### 2022-01-08 paper/paper.tex
just getting started with LIPIcs format for ITP 2022

\documentclass[a4paper,USenglish]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

%\pdfoutput=1 %uncomment to ensure pdflatex processing (mandatatory e.g. to submit to arXiv)
%\hideLIPIcs  %uncomment to remove references to LIPIcs series (logo, DOI, ...), e.g. when preparing a pre-final version to be uploaded to arXiv or another public repository

%\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Subsidiary Recursion in Coq} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Aaron Stump}{Computer Science Dept., The University of Iowa, USA \and \url{http://www.cs.uiowa.edu/~astump/}}{aaron-stump@uiowa.edu}{http://orcid.org/0000-0002-9720-0003}{ }%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\authorrunning{A. Stump} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Aaron Stump} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

\ccsdesc[100]{Software and its engineering~Recursion}
\ccsdesc[100]{Software and its engineering~Polymorphism}

\keywords{strong functional programming, recursion schemes, impredicativity} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

%\acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering

%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{June Andronick and Leonardo da Moura}
\EventNoEds{2}
\EventLongTitle{Interactive Theorem Proving 2022}
\EventShortTitle{ITP 2022}
\EventAcronym{ITP}
\EventYear{2022}
\EventDate{2022}
\EventLocation{}
\EventLogo{}
\SeriesVolume{}
\ArticleNo{}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\maketitle

%TODO mandatory: add short abstract of the document
\begin{abstract}
\end{abstract}

\section{Start}

\end{document}

### 2022-01-08 paper/paper.tex
working on paper
Outline is there now, working on describing the interface we provide

@@
 
-\author{Aaron Stump}{Computer Science Dept., The University of Iowa, USA \and \url{http://www.cs.uiowa.edu/~astump/}}{aaron-stump@uiowa.edu}{http://orcid.org/0000-0002-9720-0003}{ }%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
+\author{Aaron Stump}{Computer Science Dept., The University of Iowa, USA \and \url{http://www.cs.uiowa.edu/~astump/}}{aaron-stump@uiowa.edu}{http://orcid.org/0000-0002-9720-0003}{}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty.
 Please provide at least the name of the affiliation and the country.
@@
 
-\authorrunning{A. Stump} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+\author{Alex Hubers}{Computer Science, The University of Iowa, USA}{alexander-hubers@uiowa.edu}{}{}
 
-\Copyright{Aaron Stump} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
+\author{Christopher Jenkins}{Computer Science, The University of Iowa, USA}{alexander-hubers@uiowa.edu}{http://orcid.org/
+0000-0002-5434-5018}{}
+
+\author{Benjamin Delaware}{Computer Science, Purdue University, USA \and \url{https://www.cs.purdue.edu/homes/bendy/}}{bendy@purdue.edu}{}{}
+
+\authorrunning{A. Stump, A. Hubers, C. Jenkins, and B. Delaware} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'
+
+\Copyright{Aaron Stump, Alex Hubers, Christopher Jenkins, and Benjamin Delaware} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/
 
@@
 
-\keywords{strong functional programming, recursion schemes, impredicativity} %TODO mandatory; please add comma-separated list of keywords
+\keywords{strong functional programming, recursion schemes, positive-recursive types, impredicativity} %TODO mandatory; please add comma-separated list of keywords
 
@@
 \begin{abstract}
+  This paper describes a functor-generic derivation of a recursion
+  scheme supporting the initiation of subsidiary recursions, in Coq.
+  The approach supports making outer recursive calls on the results of
+  subsidiary recursions.  The derivation utilizes a novel (necessarily
+  weakened) form of positive-recursive types in Coq.  Subsidiary
+  induction is also supported.  Using this derivation, several
+  examples are demonstrated.
 \end{abstract}
 
-\section{Start}
+\section{Introduction: subsidiary recursion}
+
+Central to interactive theorem provers like Coq, Agda, Isabelle/HOL,
+Lean and others are terminating recursive functions over user-declared
+inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
+usually enforced by a syntactic check for structural decrease.  This
+structural termination is sufficient for many basic functions.  For
+example, the well-known \texttt{span} function from Haskell's standard
+library (\verb|Data.List|) takes a list and returns a pair of the
+maximal prefix satisfying a given predicate \verb|p|, and the
+remaining suffix:
+\begin{verbatim}
+span :: (a -> Bool) -> [a] -> ([a],[a])
+span _ []    =  ([], [])
+span p (x:xs) = if p x
+                then let (ys,zs) = span p xs in (x:ys,zs)
+                else ([],x:xs)
+\end{verbatim}
+\noindent The sole recursive call is \verb|span p xs|, and it occurs
+in a clause where the input list is of the form \verb|x:xs|.  The
+input to the recursive call is a subdatum of the input for that
+clause, and hence this definition is structurally decreasing.  In the
+appropriate syntax, it can be accepted without additional effort by
+all the mentioned provers.
+
+This paper is about a more expressive form of terminating recursion,
+called \textbf{subsidiary recursion}.  While performing an outer
+recursion on some input \verb|x|, one may initiate an inner recursion
+on \verb|x| (or possibly some of its subdata), preserving the
+possibility of further recursion.  Let us see a simple example.  The
+function \verb|wordsBy| (\verb|Data.List.Extra|) breaks a list into
+the list of its maximal sublists whose elements do not satisfy a
+predicate \verb|p|.  For example, \verb|wordsBy isSpace " good day "|
+returns \verb|["good","day"]|; so \verb|wordsBy isSpace| has the same
+behavior as \verb|words| (\verb|Data.List|).  Code is in
+Figure~\ref{fig:wordsBy}. The first recursive call,
+\verb|wordsBy p tl|, is structural.  But in the second, we invoke
+\verb|wordsBy p| on a value obtained from another recursion, namely
+\verb|span|.  This is not allowed under structural termination.
+
+\begin{figure}
+\begin{verbatim}
+wordsBy :: (a -> Bool) -> [a] -> [[a]]
+wordsBy p [] = []
+wordsBy p (hd:tl) =
+  if p hd
+  then wordsBy p tl 
+  else let (w,z) = span (not . p) tl in
+        (hd:w) : wordsBy p z
+\end{verbatim}
+\caption{Haskell code for \texttt{wordsBy}, demonstrating subsidiary recursion}
+\label{fig:wordsBy}
+\end{figure}
+
+\subsection{Summary of results}
+
+This paper presents a functor-generic derivation of terminating
+subsidiary recursion and induction in Coq.  We should emphasize that
+this is a derivation of this recursion scheme within the type theory
+of Coq.  (So no axioms or other modifications to Coq of any kind are
+required.) Based on this derivation, we present several example
+functions like \verb|wordsBy|, and prove theorems about them.  For
+example, we prove the expected property that the sublists returned by
+\verb|wordsBy| contain no elements satisfying the predicate \verb|p|.
+For another, we give a definition of run-length encoding as a
+subsidiary recursion using \verb|span|, and prove that encoding and
+then decoding returns the original list.  The approach applies to the
+standard datatypes in the Coq library, and does not require switching
+libraries or datatype definitions.
+
+An important technical novelty of our approach is a derivation of a
+weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
+restrict datatypes $D$ to be strictly positive: in the type for any
+constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
+derivation needs to use positive-recursive types, where $D$ may occur
+to the left only of an even number of arrows.  Full positive-recursive
+types are incompatible with Coq's type theory.  But we present a way
+to derive a weakened form that is sufficient for our examples
+(Section~\ref{sec:mu}).  The weakening is to require just a retraction
+between recursive type $\mu$ and its one-step unfolding $F\ \mu$,
+instead of an isomorphism.  Hence, we dub these
+\textbf{retractive-positive} recursive types.
+
+We begin by summarizing the interface our derivation provides for
+subsidiary recursion (Section~\ref{sec:interface}), and then see
+examples (Section~\ref{sec:examples}).  The interface for subsidiary
+induction is covered next (Section~\ref{sec:interfacei}), and example
+proofs using it (Section~\ref{sec:examplesi}).  We conclude with a
+walk-through of the derivation of that first interface (for subsidiary
+recursion), in Section~\ref{sec:deriv}.  Related work is discussed in
+Section~\ref{sec:related}.
+
+All presented derivations have been checked with Coq version 8.13.2.
+The code may be found as release \verb|itp-2022| (dated prior to the
+ITP 2022 deadline) at
+\url{https://github.com/astump/coq-subsidiary}.
+
+\section{Interface}
+\label{sec:interface}
+
+\begin{figure}
+  \begin{verbatim}
+Inductive ListF(X : Set) : Set :=
+| Nil : ListF X
+| Cons : A -> X -> ListF X.
+
+Definition inList : ListF List -> List := inn ListF.
+Definition outList : List -> ListF List := out ListF (fold ListF).
+Definition mkNil : List := inList Nil.
+Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
+Definition toList : list A -> List.
+Definition fromList : List -> list A.
+\end{verbatim}
+  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists}
+  \label{fig:listf}
+\end{figure}
+
+For purposes of presenting the interface for subsidiary recursion, let us consider three files in our codebase:
+
+\begin{itemize}
+\item \texttt{Subrec.v}: parametrized by a signature functor \verb|F| of type \verb|Set -> Set|,
+  this provides (among much else, to be discussed below) \verb|Subrec : Set|, as the type to use
+  for subsidiary recursion; and also a constructor \verb|inn : F Subrec -> Subrec| for that type.
+
+\item \texttt{List.v}: parametrized by the type \verb|A : Set| of list
+  elements, this file specializes the development in \verb|Subrec.v|
+  to the case of lists.  This is done using defining the signature
+  functor \verb|ListF|, shown in Figure~\ref{fig:listf}, giving us the
+  type \verb|List|.  This is not to be confused with the type
+  \verb|list| of lists in Coq's standard library.  The figure also
+  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
+  and types for conversion functions between \verb|List| and
+  \verb|list| (code omitted).
+
+\item \texttt{Span.v}: this implements the \verb|span| function
+  mentioned above, as an \emph{algebra}.
+
+\end{itemize}
+
+\subsection{Algebras for subsidiary recursion}
+
+Our approach is within a long line of work using ideas from
+universal algebra and category theory to implement inductive datatypes
+and recursion principles, in type theory.  On this approach, one describes
+transformations to be performed on data as \emph{algebras}, which can then
+be \textit{folded} over data.
+
+\begin{figure}
+\begin{verbatim}
+Definition KAlg  : Type := Set -> (Set -> Set) -> Set.
+
+Definition FoldT(alg : KAlg)(C : Set) : Set :=
+  forall (X : Set -> Set) (FunX : Functor X), alg C X -> C -> X C.
+
+Definition AlgF(Alg: KAlg)(C : Set)(X : Set -> Set) : Set :=
+  forall (R : Set)
+         (reveal : R -> C)        
+         (fold : FoldT Alg R)
+         (eval : R -> X R)      
+         (d : F R),             
+         X R.
+
+Definition Alg : KAlg := MuAlg AlgF.
+\end{verbatim}
+\caption{The interface for algebras}
+\label{fig:algf}
+\end{figure}
+
+The interface for algebras is presented in Figure~\ref{fig:algf}.
+Let us consider the definitions.  \verb|KAlg| is the kind for the type-constructor for algebras,
+as we see in the very last definition of the figure, for \verb|Alg|.  This type-constructor \verb|Alg|
+is a fixed-point of the type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|,
+to be discussed in Section~\ref{sec:mu} below.  Doing so requires that \verb|AlgF| only use its parameter
+\verb|Alg| positively.  We will confirm this shortly.
+
+The type \verb|FoldT Alg C| is the type for fold functions which apply
+algebras of type \verb|Alg| to data of type \verb|C|.  At the top
+level of code, \verb|C| would just be \verb|List| (for example).  When
+one initiates a subsidiary recursion, though, this type \verb|C| will
+instead by what we like to call the \emph{recursion universe} of that
+recursion.  Each recursion is based on an abstract type \verb|R|,
+representing the data upon which we will recurse.  Indeed, \verb|AlgF|
+says that algebras take in such an \verb|R| as their first argument.
+This means that an algebra can assume nothing about \verb|R| except
+that it has the operations given next in the listing of \verb|AlgF|.
+Thus, \verb|R| functions like a universe for recursion: one may
+perform the various given operations, without leaving the type
+\verb|R| (which is good, because we may recurse on elements of type \verb|R|).
+
+And what are the operations on \verb|R|?  First there is \verb|reveal|, which
+turns an \verb|R| into a \verb|C|.  This reveals that the data of type
+\verb|R| are really lists (if this is a top-level recursion) or else
+really belong to some outer recursion universe (if this is an inner
+recursion).  Next we have \texttt{fold}, which will allow us to fold
+another algebra over data of type \verb|R|.  We will use \verb|fold|
+to initiate subsidiary recursions.  Then there is \verb|eval|, which
+is used to make recursive calls on data of type \verb|R|.
+
+This is a good place to highlight that for subsidiary recursion, it is
+crucial that algebras have a carrier \verb|X| which depends
+(functorially) on a type.  This is so that (i) inside an inner
+recursion we may compute a result of some type that may mention
+\verb|R|, but (ii) outside that recursion, the result will mention the
+type \verb|C|.  The \verb|eval| function returns something of type
+\verb|X R|, demonstrating (i).  For (ii): if we look at the definition
+of \verb|FoldT| in the figure, we see that folding an algebra of type
+\verb|alg C X| over a value of type \verb|C| produces a result of type
+\verb|X C|.
+
+\section{Examples of subsidiary recursion}
+\label{sec:examples}
+
+\section{Interface for subsidiary induction}
+\label{sec:interfacei}
+
+\section{Examples of subsidiary induction}
+\label{sec:interfacei}
+
+\section{Derivation of subsidiary recursion}
+\label{sec:deriv}
+
+\subsection{Retractive-positive recursive types}
+\label{sec:mu}
+
+\section{Related Work}
+\label{sec:related}
+
+In some tools, like Coq, Agda, and Lean, termination is checked
+statically, based on structural decrease at recursive calls.  Others,
+like Isabelle/HOL, allow one to write recursions first, and prove
+(possibly with automated help) their termination
+afterwards~\cite{krauss}.
+
+It has not escaped the notice of designers of ITPs that structural
+recursion is not the only form of terminating recursion.  All the
+mentioned tools provide support for well-founded recursion, where for
+recursive calls, one must show that the parameter of recursion has
+decreased in some well-founded order.
+
+Subsidiary recursion can be seen as a generalization
+of nested recursion, which is a form of recursion allowing recursive
+calls of the form \verb|f\ (f\ x)|.  In subsidiary recursion, these
+are generalized to the form \verb|f\ (g\ x)|, where \verb|g| is
+another recursively defined function.
+
+\bibliography{biblio}
 

### 2022-01-08 paper/Examples.hs

@@ -1,0 +1,16 @@
+module Examples where
+
+span' :: (a -> Bool) -> [a] -> ([a],[a])
+span' _ []    =  ([], [])
+span' p (x:xs) = if p x then
+                   let (ys,zs) = span' p xs in (x:ys,zs)
+                 else
+                   ([],x:xs)
+
+wordsBy :: (a -> Bool) -> [a] -> [[a]]
+wordsBy p [] = []
+wordsBy p (hd:tl) =
+  if p hd
+  then wordsBy p tl 
+  else let (w,z) = span (not . p) tl in
+        (hd:w) : wordsBy p z

### 2022-01-08 code/List/Scratch.hs
tweaking some names as I start writing the paper

@@ -15,10 +15,11 @@ wordsBy f s = case dropWhile f s of
 
 wordsBy' :: (a -> Bool) -> [a] -> [[a]]
 wordsBy' p [] = []
-wordsBy' p (hd:tl) | p hd = wordsBy' p tl
-                   | otherwise =
-   let (w,z) = break p tl in
-     (hd:w) : wordsBy' p z
+wordsBy' p (hd:tl) =
+  if p hd
+  then wordsBy' p tl 
+  else let (w,z) = break p tl in
+        (hd:w) : wordsBy' p z
 
 rle :: Eq a => [a] -> [(Int,a)]
 rle = mapThrough compressSpan

### 2022-01-10 paper/paper.tex
further text

@@
 
+\newcommand{\all}[2]{\forall\, #1.\, #2}
+
 \begin{document}
@@
 \begin{abstract}
-  This paper describes a functor-generic derivation of a recursion
-  scheme supporting the initiation of subsidiary recursions, in Coq.
-  The approach supports making outer recursive calls on the results of
-  subsidiary recursions.  The derivation utilizes a novel (necessarily
-  weakened) form of positive-recursive types in Coq.  Subsidiary
-  induction is also supported.  Using this derivation, several
-  examples are demonstrated.
+  This paper describes a functor-generic derivation in Coq of
+  subsidiary recursion.  On this recursion scheme, inner recursions
+  may be initiated within outer ones, in such a way that outer
+  recursive calls may be made on results from inner ones.  The
+  derivation utilizes a novel (necessarily weakened) form of
+  positive-recursive types in Coq, dubbed retractive-positive
+  recursive types.  A corresponding form of induction is also
+  supported.  The method is demonstrated through several examples.
 \end{abstract}
@@
 \noindent The sole recursive call is \verb|span p xs|, and it occurs
-in a clause where the input list is of the form \verb|x:xs|.  The
-input to the recursive call is a subdatum of the input for that
-clause, and hence this definition is structurally decreasing.  In the
-appropriate syntax, it can be accepted without additional effort by
-all the mentioned provers.
+in a clause where the input list is of the form \verb|x:xs|.  So the
+input to the recursive call is a subdatum of the input, and hence this
+definition is structurally decreasing.  In the appropriate syntax, it
+can be accepted without additional effort by all the mentioned
+provers.
 
@@
 on \verb|x| (or possibly some of its subdata), preserving the
-possibility of further recursion.  Let us see a simple example.  The
-function \verb|wordsBy| (\verb|Data.List.Extra|) breaks a list into
-the list of its maximal sublists whose elements do not satisfy a
-predicate \verb|p|.  For example, \verb|wordsBy isSpace " good day "|
-returns \verb|["good","day"]|; so \verb|wordsBy isSpace| has the same
-behavior as \verb|words| (\verb|Data.List|).  Code is in
+possibility of further invocations of the outer recursive function.
+Let us see a simple example.  The function \verb|wordsBy|
+(from \verb|Data.List.Extra|) breaks a list into its maximal
+sublists whose elements do not satisfy a predicate \verb|p|.  For
+example, \verb|wordsBy isSpace " good day "| returns
+\verb|["good","day"]|; so \verb|wordsBy isSpace| has the same behavior
+as \verb|words| (from \verb|Data.List|).  Code is in
 Figure~\ref{fig:wordsBy}. The first recursive call,
@@
 \verb|wordsBy p| on a value obtained from another recursion, namely
-\verb|span|.  This is not allowed under structural termination.
+\verb|span|.  This is not allowed under structural termination,
+but will be permitted by subsidiary recursion as derived below.
 
@@
 this is a derivation of this recursion scheme within the type theory
-of Coq.  (So no axioms or other modifications to Coq of any kind are
-required.) Based on this derivation, we present several example
+of Coq.  No axioms or other modifications to Coq of any kind are
+required. Based on this derivation, we present several example
 functions like \verb|wordsBy|, and prove theorems about them.  For
 example, we prove the expected property that the sublists returned by
-\verb|wordsBy| contain no elements satisfying the predicate \verb|p|.
+\verb|wordsBy| consist of elements satisfying \verb|not . p|.
 For another, we give a definition of run-length encoding as a
 subsidiary recursion using \verb|span|, and prove that encoding and
-then decoding returns the original list.  The approach applies to the
+then decoding returns the original list.  Our approach applies to the
 standard datatypes in the Coq library, and does not require switching
@@
 derivation needs to use positive-recursive types, where $D$ may occur
-to the left only of an even number of arrows.  Full positive-recursive
-types are incompatible with Coq's type theory.  But we present a way
-to derive a weakened form that is sufficient for our examples
-(Section~\ref{sec:mu}).  The weakening is to require just a retraction
-between recursive type $\mu$ and its one-step unfolding $F\ \mu$,
-instead of an isomorphism.  Hence, we dub these
-\textbf{retractive-positive} recursive types.
+to the left only of an even number of arrows.  The restriction to
+strict positivity is enforced because full positive-recursive types
+are incompatible with Coq's type theory~\cite{coquand88}.  But we
+present a way to derive a weakened form that is sufficient for our
+examples (Section~\ref{sec:mu}).  The weakening is to require only
+that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
+and $F\ \mu$ its one-step unfolding.  Usually these types are
+isomorphic.  Hence, we dub these \textbf{retractive-positive}
+recursive types.  This weakening does have some negative consequences,
+but we will see how to handle them.  Our definition of
+retractive-positive recursive types makes essential use of impredicate
+quantification, and hence cannot be soundly recapitulated in a
+predicative theory like Agda's.
 
@@
 subsidiary recursion (Section~\ref{sec:interface}), and then see
-examples (Section~\ref{sec:examples}).  The interface for subsidiary
-induction is covered next (Section~\ref{sec:interfacei}), and example
-proofs using it (Section~\ref{sec:examplesi}).  We conclude with a
-walk-through of the derivation of that first interface (for subsidiary
-recursion), in Section~\ref{sec:deriv}.  Related work is discussed in
+examples (Section~\ref{sec:examples}).  We next explain how the
+interface is actually implemented (Section~\ref{sec:deriv}), including
+our retractive-positive recursive types (Section~\ref{sec:mu}).  The
+interface for subsidiary induction is covered next
+(Section~\ref{sec:interfacei}), and example proofs using it
+(Section~\ref{sec:examplesi}).  Related work is discussed in
 Section~\ref{sec:related}.
 
-All presented derivations have been checked with Coq version 8.13.2.
-The code may be found as release \verb|itp-2022| (dated prior to the
-ITP 2022 deadline) at
-\url{https://github.com/astump/coq-subsidiary}.
+All presented derivations have been checked with Coq version 8.13.2,
+using command-line option \verb|-impredicative-set|.  The code may be
+found as release \verb|itp-2022| (dated prior to the ITP 2022
+deadline) at \url{https://github.com/astump/coq-subsidiary}.
 
-\section{Interface}
+\section{Interface for subsidiary recursion}
 \label{sec:interface}
+
+This section presents the interface our Coq development provides
+for subsidiary recursion.
+
+\subsection{The recursion universe}
+
+Our approach is within a long line of work using ideas from universal
+algebra and category theory to describe inductive datatypes and their
+recursion principles.  On this approach, one describes transformations
+to be performed on data as \emph{algebras}, which can then be
+\textit{folded} over data.  The simplest form of algebras, namely
+$F$-algebras, are morphisms from $F\ A$ to $A$, for carrier object
+$A$.  From a programming perspective, an $F$-algebra is presented
+with $F\ A$, and must compute a value of type $A$.
+
+Algebras for our subsidiary recursion are quite a bit more complex than
+this.  First, for reasons we will explain further below, the carrier
+of the algebra will be an \verb|X : Set -> Set|.  Second, instead of
+being given just $F\ A$ as input, the algebra is presented with:
+
+\begin{itemize}
+\item a type \verb|R : Set|
+\item a function \verb|reveal : R -> C|, which reveals values of type \verb|R| as really having \emph{anchor type} \verb|C|.
+This type \verb|C| will either by the abstract type of some outer recursion, or else our development's version of the actual datatype one is recursing over.
+\item a function \verb|fold : FoldT Alg R|, which allows one to initiate subsidiary recursions in which the anchor type is \verb|R|.  We will present its type \verb|FoldT Alg R| below.
+\item a function \verb|eval : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
+\item and an \emph{subdata structure} \verb|d : F R|, where \verb|F| is the signature functor for the datatype.
+\end{itemize}
+
+\noindent The algebra is then required to produce a value of type \verb|F R|.
+
+We use Coq inductive types for the signature functors \verb|F| of
+various datatypes, thus enabling recursions to use Coq's
+pattern-matching on the subdata structure \verb|d|.  So the style of
+coding against this interface retains a similar feel to structural
+recursions.  Unlike with structural termination, though, the interface
+here is type-based and hence compositional.  As we will see, it
+supports nested and higher-order recursions.
+
+As in previous work, we dub this interface a \emph{recursion
+universe}~\cite{stump20}.  As in other domains using the term
+``universe'', we have an entity (here, \verb|R|) from which one cannot
+escape by using the available operations (for other cases:
+$\epsilon_0$ and $\omega^-$, the physical universe and traveling at
+the speed of light).  Staying in the recursion universe is good,
+because we may recurse (via \verb|eval|) on any value of type
+\verb|R|.
+
+Some points must still be explained, particularly why \verb|X| has type \verb|Set -> Set|,
+and the definition of \verb|FoldT|.  Let us see these details next.
+
+\subsection{The interface in more detail}
+
+Let us consider two files from our development.
 
@@
 Definition inList : ListF List -> List := inn ListF.
-Definition outList : List -> ListF List := out ListF (fold ListF).
 Definition mkNil : List := inList Nil.
@@
 \end{verbatim}
-  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists}
+  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
   \label{fig:listf}
@@
 
-For purposes of presenting the interface for subsidiary recursion, let us consider three files in our codebase:
+\subsubsection{Subrec.v}
 
-\begin{itemize}
-\item \texttt{Subrec.v}: parametrized by a signature functor \verb|F| of type \verb|Set -> Set|,
-  this provides (among much else, to be discussed below) \verb|Subrec : Set|, as the type to use
-  for subsidiary recursion; and also a constructor \verb|inn : F Subrec -> Subrec| for that type.
+  This file is parametrized by a signature functor \verb|F| of type
+  \verb|Set -> Set|.  It provides the implementation of subsidiary
+  recursion.  Two crucial values are \verb|Subrec : Set|, which is the
+  type to use for subsidiary recursion; and
+  \verb|inn : F Subrec -> Subrec|, which is to be used as a
+  constructor for that type.  An important point, however, is that
+  \verb|Subrec.v| does not provide an induction principle based on
+  \verb|inn|.  Induction is derived later
+  (Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use
+  of retractive-positive recursive types, to take a fixed-point based
+  on \verb|F|.  We will present the definition of those in
+  Section~\ref{sec:mu}.
 
-\item \texttt{List.v}: parametrized by the type \verb|A : Set| of list
-  elements, this file specializes the development in \verb|Subrec.v|
-  to the case of lists.  This is done using defining the signature
-  functor \verb|ListF|, shown in Figure~\ref{fig:listf}, giving us the
-  type \verb|List|.  This is not to be confused with the type
-  \verb|list| of lists in Coq's standard library.  The figure also
-  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
-  and types for conversion functions between \verb|List| and
-  \verb|list| (code omitted).
+  \subsubsection{List.v}
 
-\item \texttt{Span.v}: this implements the \verb|span| function
-  mentioned above, as an \emph{algebra}.
-
-\end{itemize}
+  This file specializes the development in \verb|Subrec.v| to the case
+  of lists (parametrized by the type \verb|A| of elements).  In
+  general, to use our development to get subsidiary recursion over
+  some datatype, one will have a similar ``shim'' file.  The file
+  defines the signature functor \verb|ListF|, shown in
+  Figure~\ref{fig:listf}.  Using \verb|Subrec|, we then get a type
+  \verb|List|.  This is not to be confused with the type \verb|list|
+  of lists in Coq's standard library.  As noted previously, our
+  development is meant to be used in extension of existing inductive
+  datatypes, not replacing them.  The figure also shows constructors
+  \verb|mkNil| and \verb|mkCons| for \verb|List|, and types for
+  conversion functions between \verb|List| and \verb|list| (see
+  Section~\ref{sec:deriv} for code).
 
@@
 
-Our approach is within a long line of work using ideas from
-universal algebra and category theory to implement inductive datatypes
-and recursion principles, in type theory.  On this approach, one describes
-transformations to be performed on data as \emph{algebras}, which can then
-be \textit{folded} over data.
+\verb|Subrec.v| also defines the notion of algebra that is used for writing
+recursions.  The central definitions are in Figure~\ref{fig:algf}.
+\verb|KAlg| is the kind for the type-constructor for algebras, as we
+see in the very last definition of the figure, for \verb|Alg|.  This
+type-constructor \verb|Alg| is a fixed-point of the type \verb|AlgF|.
+The fixed-point is taken using \verb|MuAlg| (Section~\ref{sec:mu}
+below).  Doing so requires that \verb|AlgF| only use its parameter
+\verb|Alg| positively.  We will confirm this shortly.
 
@@
 \end{verbatim}
-\caption{The interface for algebras}
+\caption{The type for algebras (\texttt{Subrec.v})}
 \label{fig:algf}
@@
 
-The interface for algebras is presented in Figure~\ref{fig:algf}.
-Let us consider the definitions.  \verb|KAlg| is the kind for the type-constructor for algebras,
-as we see in the very last definition of the figure, for \verb|Alg|.  This type-constructor \verb|Alg|
-is a fixed-point of the type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|,
-to be discussed in Section~\ref{sec:mu} below.  Doing so requires that \verb|AlgF| only use its parameter
-\verb|Alg| positively.  We will confirm this shortly.
+The type \verb|FoldT Alg C| is the type for fold functions which apply
+algebras of type \verb|Alg| to data of type \verb|C|, which we have
+already dubbed the \emph{anchor type} of the recursion.  At the top
+level of code, the anchor type would just be \verb|List| (for
+example).  When one initiates a subsidiary recursion, though, the
+anchor type will instead by the abstract type \verb|R| for the outer recursion.
 
-The type \verb|FoldT Alg C| is the type for fold functions which apply
-algebras of type \verb|Alg| to data of type \verb|C|.  At the top
-level of code, \verb|C| would just be \verb|List| (for example).  When
-one initiates a subsidiary recursion, though, this type \verb|C| will
-instead by what we like to call the \emph{recursion universe} of that
-recursion.  Each recursion is based on an abstract type \verb|R|,
-representing the data upon which we will recurse.  Indeed, \verb|AlgF|
-says that algebras take in such an \verb|R| as their first argument.
-This means that an algebra can assume nothing about \verb|R| except
-that it has the operations given next in the listing of \verb|AlgF|.
-Thus, \verb|R| functions like a universe for recursion: one may
-perform the various given operations, without leaving the type
-\verb|R| (which is good, because we may recurse on elements of type \verb|R|).
+The variable \verb|Alg| occurs only positively (but not strictly
+positively) in \verb|AlgF|, because it occurs negatively in
+\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
+we can indeed take a fixed-point of \verb|AlgF| to define the constant
+\verb|Alg|.
 
-And what are the operations on \verb|R|?  First there is \verb|reveal|, which
-turns an \verb|R| into a \verb|C|.  This reveals that the data of type
-\verb|R| are really lists (if this is a top-level recursion) or else
-really belong to some outer recursion universe (if this is an inner
-recursion).  Next we have \texttt{fold}, which will allow us to fold
+Let us look at \verb|AlgF|.  As noted already, each recursion is based
+on an abstract type \verb|R|, representing the data upon which we will
+recurse.  This is the first argument to a value of type
+\verb|AlgF Alg C X|.  An algebra can assume nothing about \verb|R|
+except that it supports the following operations.  First there is
+\verb|reveal|, which turns an \verb|R| into a \verb|C|.  This reveals
+that the data of type \verb|R| are really values of the datatype
+(\verb|List|, for example), if this is a top-level recursion; or else
+really belong to some outer recursion universe, if this is an inner
+recursion.  Next we have \texttt{fold}, which will allow us to fold
 another algebra over data of type \verb|R|.  We will use \verb|fold|
@@
 
-This is a good place to highlight that for subsidiary recursion, it is
-crucial that algebras have a carrier \verb|X| which depends
-(functorially) on a type.  This is so that (i) inside an inner
-recursion we may compute a result of some type that may mention
-\verb|R|, but (ii) outside that recursion, the result will mention the
-type \verb|C|.  The \verb|eval| function returns something of type
-\verb|X R|, demonstrating (i).  For (ii): if we look at the definition
-of \verb|FoldT| in the figure, we see that folding an algebra of type
+As noted already, for subsidiary recursion, algebras have a carrier
+\verb|X| which depends (functorially) on a type.  This is so that (i)
+inside an inner recursion we may compute a result of some type that
+may mention \verb|R|, but (ii) outside that recursion, the result will
+mention the anchor type \verb|C|.  The \verb|eval| function returns
+something of type \verb|X R|, and so does the algebra itself; this
+demonstrates (i).  For (ii): if we look at the definition of
+\verb|FoldT| in the figure, we see that folding an algebra of type
 \verb|alg C X| over a value of type \verb|C| produces a result of type
-\verb|X C|.
+\verb|X C|.  Having a functor for the carrier of the algebra gives us
+the flexibility to type results inside a recursion with the abstract
+type \verb|R|, but view those results as having the anchor type
+\verb|C| outside the recursion.
 
@@
 \label{sec:examples}
-
-\section{Interface for subsidiary induction}
-\label{sec:interfacei}
-
-\section{Examples of subsidiary induction}
-\label{sec:interfacei}
 
@@
 
+As we have seen, our definitions require a form of positive-recursive
+types, to allow algebras to accept fold functions that themselves
+require algebras, and also for the definition of \verb|Subrec|.  But
+as recalled already, full positive-recursive types are incompatible
+with Coq's type theory~\cite{coquand88}.  It is worth noting that one
+can impose some restrictions on large eliminations which then allow
+positive-recursive types~\cite{blanqui05}.  This approach would
+require changing the underlying theory.  To avoid this, we here take a
+different approach, exploiting Coq's impredicative polymorphism.
+
+This is done in a file \verb|Mu.v|, whose central definitions are in
+Figure~\ref{fig:mu}.  The development is parametrized by
+\verb|F : Set -> Set| which is assumed to have an \verb|fmap| function
+(morphism part of the functor) of type
+\begin{verbatim}
+forall A B : Set, (A -> B) -> F A -> F B
+\end{verbatim}
+\noindent which satisfies the identity-preservation law for functors:
+\begin{verbatim}
+fmapId : forall (A : Set)(d : F A), fmap (fun x => x) d = d
+\end{verbatim}
+
+\begin{figure}
+\begin{verbatim}
+  Inductive Mu : Set := 
+    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
+
+  Definition inMu(d : F Mu) : Mu :=
+    mu Mu (fun x => x) d.
+
+  Definition outMu(m : Mu) : F Mu :=
+    match m with
+    | mu A r d => fmap r d
+    end.
+
+  Lemma outIn(d : F Mu) : outMu (inMu d) = d.
+\end{verbatim}
+\caption{Derivation of retractive-positive recursive types}
+\label{fig:mu}
+\end{figure}
+
+Let us consider the code in Figure~\ref{fig:mu}.  The critical
+idea is embodied in the definition of \verb|Mu|.  We would like
+to have a definition like
+\begin{verbatim}
+  Inductive Mu' : Set := mu' : F Mu' -> Mu'.
+\end{verbatim}
+\noindent This is exactly what is used in approaches to modular
+datatypes in functional programming, like
+Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
+rejected by Coq, as it would unsoundly enable instantiations of
+\verb|F| that are not strictly positive.
+
+Instead, the definition of \verb|Mu| in Figure~\ref{fig:mu} weakens
+this ideal definition to a strictly positive approximation:
+\begin{verbatim}
+  Inductive Mu : Set := 
+    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
+\end{verbatim}
+\noindent Instead of taking in \verb|F Mu|, constructor \verb|mu|
+accepts an input of type \verb|F R|, for some type \verb|R| for which
+we have a function of type \verb|R -> Mu|.  The impredicative
+quantification of \verb|R| is essential here: we instantiate it with
+\verb|Mu| itself in the definition of \verb|inMu|
+(Figure~\ref{fig:mu}).  So this approach would not work in a
+predicative theory like Agda's.  The quantification of \verb|R| can be
+seen as applying a technique due to Mendler, of introducing
+universally quantified variables for problematic type occurrences, to
+a datatype constructor.  We will review this in
+Section~\ref{sec:related}.
+
+Returning to Figure~\ref{fig:mu}, we have functions \verb|inMu| and
+\verb|outMu|, which make \verb|F Mu| a retraction (\verb|outIn|) of
+\verb|Mu|: the composition of \verb|outMu| and \verb|inMu| is
+(extensionally) the identity on \verb|F Mu|.  But the reverse
+composition cannot be proved to be the identity, because of the basic
+problem of \textbf{noncanonicity} that arises with this definition.
+
+For a simple example of noncanonicity, suppose we instantiate \verb|F|
+with \verb|ListF| (of Figure~\ref{fig:listf}).  Please note that as
+\verb|Mu| is used in our derivation of subsidiary recursion, we will
+not instantiate this \verb|F| with the signature functor of a datatype
+directly; but this will show the issue in a simple form.  Let us
+temporarily define \verb|List A| as \verb|Mu (ListF A)| (again, for
+subsidiary recursion we use a different functor than just \verb|ListF|
+directly).  The canonical way to define the empty list would be, 
+implicitly instantiating \verb|F| to \verb|ListF A|,
+\begin{verbatim}
+Definition mkNil := mu (List A) (fun x => x) (NilF A)
+\end{verbatim}
+\noindent But given this, there are infinitely many other equivalent
+definitions.  For any \verb|Q : Set|, we could take
+\begin{verbatim}
+Definition mkNil' := mu Q (fun x => mkNil) (NilF A)
+\end{verbatim}
+\noindent Since \verb|fmap f (NilF A)| equals just \verb|NilF B| for
+\verb|f : A -> B|, if we apply \verb|outMu| (of Figure~\ref{fig:mu})
+to \verb|mkNil'| or \verb|mkNil|, we will get \verb|NilF (List A)|.
+But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
+definitionally nor provably.  One can define a function that puts
+\verb|Mu| values in normal form by folding \verb|inMu| over them.
+Then \verb|mkNil| and \verb|mkNil'| will have the same normal form,
+and be equivalent in that sense.  But the fact that they are not
+provably equal is what we term noncanonicity.
+
+Noncanonicity leads to some issues, as we turn next to the problem of
+inductive reasoning about subsidiary recursions.  With some care,
+however, we can avoid pitfalls, leaving us with a form of
+positive-recursive type that enables our definitions to go through.
+
+\section{Interface for subsidiary induction}
+\label{sec:interfacei}
+
+\section{Examples of subsidiary induction}
+\label{sec:examplesi}
+
 \section{Related Work}
@@
 
+\subsection{Termination}
 In some tools, like Coq, Agda, and Lean, termination is checked
@@
 
-Subsidiary recursion can be seen as a generalization
-of nested recursion, which is a form of recursion allowing recursive
-calls of the form \verb|f\ (f\ x)|.  In subsidiary recursion, these
-are generalized to the form \verb|f\ (g\ x)|, where \verb|g| is
-another recursively defined function.
+Subsidiary recursion can be seen as a generalization of \emph{nested
+recursion}, which allows recursive calls of the form \verb|f (f x)|.
+In subsidiary recursion, these are generalized to the form
+\verb|f (g x)|, where \verb|g| could be \verb|f| or another
+recursively defined function.
+
+\subsection{Mendler encoding}
+
+Mendler introduced the basic idea of using universal abstraction to
+support compositional termination checking; an accessible source
+is~\cite{mendler91}.  This recursor has type
+\[
+\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to X}
+\]
+\noindent We have adopted this idea to the constructor of the type
+\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
+categorical perspective on Mendler-style recursion~\cite{uustalu99}.
+Others have explored the possibility of using it with negative type
+schemes~\cite{ahn11}.
 

### 2022-01-12 paper/paper.tex
revising text, description of Mu

@@
 derivation needs to use positive-recursive types, where $D$ may occur
-to the left only of an even number of arrows.  The restriction to
-strict positivity is enforced because full positive-recursive types
-are incompatible with Coq's type theory~\cite{coquand88}.  But we
-present a way to derive a weakened form that is sufficient for our
-examples (Section~\ref{sec:mu}).  The weakening is to require only
-that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
-and $F\ \mu$ its one-step unfolding.  Usually these types are
-isomorphic.  Hence, we dub these \textbf{retractive-positive}
-recursive types.  This weakening does have some negative consequences,
-but we will see how to handle them.  Our definition of
+to the left of an even number (only) of arrows.  Coq requires strict
+positivity because in the presence of other features of Coq's theory,
+full positive-recursive types lead to a paradox~\cite{coquand88}.  We
+present a way to derive a weakened form of positive-recursive type
+that is sufficient for our examples (Section~\ref{sec:mu}).  The
+weakening is to require only that $F\ \mu$ is a retract of $\mu$,
+where $\mu$ is the recursive type and $F\ \mu$ its one-step unfolding.
+Usually these types are isomorphic.  Hence, we dub these
+\textbf{retractive-positive} recursive types.  This weakening has the
+negative consequence of leading to a form of noncanonicity, but we
+will see how to work around this.  Our definition of
 retractive-positive recursive types makes essential use of impredicate
@@
 $F$-algebras, are morphisms from $F\ A$ to $A$, for carrier object
-$A$.  From a programming perspective, an $F$-algebra is presented
-with $F\ A$, and must compute a value of type $A$.
+$A$.  From a programming perspective, an $F$-algebra is given input of
+type $F\ A$, and must compute a result of type $A$.
 
-Algebras for our subsidiary recursion are quite a bit more complex than
-this.  First, for reasons we will explain further below, the carrier
-of the algebra will be an \verb|X : Set -> Set|.  Second, instead of
-being given just $F\ A$ as input, the algebra is presented with:
+Algebras for our subsidiary recursion are more complex.  First, for
+reasons we will explain further below, the carrier of the algebra will
+be a functor \verb|X : Set -> Set|.  Second, algebras have a specified
+\emph{anchor type} \verb|C|, which we can think of as the datatype
+\emph{as viewed by a containing recursion} or else, if this is a
+top-level recursion, our development's version of the actual datatype
+(e.g., \verb|List|).  The algebra is presented with:
 
 \begin{itemize}
-\item a type \verb|R : Set|
-\item a function \verb|reveal : R -> C|, which reveals values of type \verb|R| as really having \emph{anchor type} \verb|C|.
-This type \verb|C| will either by the abstract type of some outer recursion, or else our development's version of the actual datatype one is recursing over.
-\item a function \verb|fold : FoldT Alg R|, which allows one to initiate subsidiary recursions in which the anchor type is \verb|R|.  We will present its type \verb|FoldT Alg R| below.
+\item a type \verb|R : Set|, which will be this recursion's view of the datatype.
+\item a function \verb|reveal : R -> C|, which reveals values of type \verb|R| as really having the anchor type.  
+\item a function \verb|fold : FoldT Alg R|, which allows one to initiate subsidiary recursions in which the anchor type is \verb|R|.
+Note that the algebra's anchor type is \verb|C|, but for subsidiary recursions the anchor type changes (to \verb|R|).
+We will present the type \verb|FoldT Alg R| below.
 \item a function \verb|eval : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
-\item and an \emph{subdata structure} \verb|d : F R|, where \verb|F| is the signature functor for the datatype.
+\item and a \emph{subdata structure} \verb|d : F R|, where \verb|F| is the signature functor for the datatype.
 \end{itemize}
 
-\noindent The algebra is then required to produce a value of type \verb|F R|.
+\noindent The algebra is then required to produce a value of type \verb|X R|.
 
-We use Coq inductive types for the signature functors \verb|F| of
+We will use Coq inductive types for the signature functors \verb|F| of
 various datatypes, thus enabling recursions to use Coq's
@@
 escape by using the available operations (for other cases:
-$\epsilon_0$ and $\omega^-$, the physical universe and traveling at
+the ordinal $\epsilon_0$ and $\omega^-$, the physical universe and traveling at
 the speed of light).  Staying in the recursion universe is good,
@@
 Some points must still be explained, particularly why \verb|X| has type \verb|Set -> Set|,
-and the definition of \verb|FoldT|.  Let us see these details next.
+and the definition of \verb|FoldT|.  Let us see these and other details next.
 
@@
 
-Let us consider two files from our development.
+Let us consider two central files from our development.
 
@@
   (Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use
-  of retractive-positive recursive types, to take a fixed-point based
-  on \verb|F|.  We will present the definition of those in
-  Section~\ref{sec:mu}.
+  of retractive-positive recursive types, to take a fixed-point of a
+  construction based on \verb|F|.  We present these recursive
+  types in Section~\ref{sec:mu} below.
 
@@
   conversion functions between \verb|List| and \verb|list| (see
-  Section~\ref{sec:deriv} for code).
+  Section~\ref{sec:deriv} for the code).
 
@@
 
-\verb|Subrec.v| also defines the notion of algebra that is used for writing
-recursions.  The central definitions are in Figure~\ref{fig:algf}.
-\verb|KAlg| is the kind for the type-constructor for algebras, as we
-see in the very last definition of the figure, for \verb|Alg|.  This
-type-constructor \verb|Alg| is a fixed-point of the type \verb|AlgF|.
-The fixed-point is taken using \verb|MuAlg| (Section~\ref{sec:mu}
-below).  Doing so requires that \verb|AlgF| only use its parameter
-\verb|Alg| positively.  We will confirm this shortly.
+\verb|Subrec.v| also defines the notion of algebra that is used for
+writing recursions.  The central definitions are in
+Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
+type-constructor for algebras, as we see in the definition of
+\verb|Alg|.  This type-constructor \verb|Alg| is a fixed-point of the
+type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|
+(Section~\ref{sec:mu}), which implements our retractive-positive
+recursive types at kind \verb|KAlg|.  Using \verb|Alg| will require
+that \verb|AlgF| only uses its parameter \verb|Alg| positively.  We
+will confirm this shortly.
 
@@
 Definition Alg : KAlg := MuAlg AlgF.
+
+Definition fold : FoldT Alg Subrec.
 \end{verbatim}
@@
 \verb|reveal|, which turns an \verb|R| into a \verb|C|.  This reveals
-that the data of type \verb|R| are really values of the datatype
-(\verb|List|, for example), if this is a top-level recursion; or else
-really belong to some outer recursion universe, if this is an inner
-recursion.  Next we have \texttt{fold}, which will allow us to fold
+that the data of type \verb|R| are really values of the anchor type
+of this recursion.  Next we have \texttt{fold}, which will allow us to fold
 another algebra over data of type \verb|R|.  We will use \verb|fold|
-to initiate subsidiary recursions.  Then there is \verb|eval|, which
-is used to make recursive calls on data of type \verb|R|.
+to initiate subsidiary recursions.  Then there is \verb|eval|, for
+recursive calls on data of type \verb|R|.
 
@@
 
+The final definition in the figure is for \verb|fold|, which allows
+us to fold an \verb|Alg| over a value of type \verb|Subrec|.  We will
+return to the code for this, and definitions of \verb|Subrec| and \verb|inn|,
+in Section~\ref{sec:deriv}.
+
 \section{Examples of subsidiary recursion}
 \label{sec:examples}
+
+Having seen now the interface for subsidiary recursion in Coq, let us
+consider now some examples, listed by their filename in the development.
+
+\subsection{\texttt{Span.v}}
 
@@
 Let us consider the code in Figure~\ref{fig:mu}.  The critical
-idea is embodied in the definition of \verb|Mu|.  We would like
+idea is embodied in the definition of \verb|Mu|.  Ideally, we would like
 to have a definition like
@@
 Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
-rejected by Coq, as it would unsoundly enable instantiations of
-\verb|F| that are not strictly positive.
+rejected by Coq, as instantiations of
+\verb|F| that are not strictly positive would be unsound.
 

### 2022-01-12 paper/paper.tex
writing up SpanAlg

@@
 Definition fold : FoldT Alg Subrec.
+Definition rollAlg : forall {C : Set} {X : Set -> Set}, AlgF Alg C X -> Alg C X.
+Definition unrollAlg : forall {C : Set} {X : Set -> Set}, Alg C X -> AlgF Alg C X.
 \end{verbatim}
@@
 
-The final definition in the figure is for \verb|fold|, which allows
-us to fold an \verb|Alg| over a value of type \verb|Subrec|.  We will
-return to the code for this, and definitions of \verb|Subrec| and \verb|inn|,
-in Section~\ref{sec:deriv}.
+The final definitions in the figure are for \verb|fold|, which allows
+us to fold an \verb|Alg| over a value of type \verb|Subrec|; and for mapping
+between \verb|Alg| and its unfolding in terms of \verb|AlgF|.  We will
+return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.
 
@@
 
-Having seen now the interface for subsidiary recursion in Coq, let us
-consider now some examples, listed by their filename in the development.
+Having seen the interface for subsidiary recursion in Coq, let us
+consider now some examples.
 
-\subsection{\texttt{Span.v}}
+\subsection{The \texttt{span} function (\texttt{Span.v})}
+
+Given a predicate \verb|p : A -> bool|, and a value of type
+\verb|List A|, we would like to compute a pair of type
+\verb|list A * List A|, where the first component is the maximal
+prefix whose elements satisfy \verb|p|, and the second is the
+remaining suffix.  This is the typing for a top-level recursion.  More
+generally, though, given an anchor type \verb|R : Set| along with a
+fold function for that anchor type (i.e., of type \verb|FoldT (Alg (ListF A)) R|),
+we would like to map an input list of type \verb|R| to a pair of
+type \verb|list A * R|.  The first component of this pair is
+going to be built up from scratch, and so cannot have type \verb|R|.
+But the second component will be a subdatum of the input list,
+and so can still have type \verb|R|.  This will enable outer recursions
+to continue on that component.  So we want: 
+\begin{verbatim}
+Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
+                (p : A -> bool)(xs : R) : list A * R.
+\end{verbatim}
+\noindent From this we can also define the top-level recursion, by
+supplying \verb|fold (ListF A)|, which is the function for folding an algebra
+over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
+\begin{verbatim}
+Definition span(p : A -> bool)(xs : List A) : list A * List A
+  := spanr (fold (ListF A)) p xs.
+\end{verbatim}
+
+Before we define \verb|spanr|, we must resolve a small problem.
+If the first element of the input list
+\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
+should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
+will see it only in the form of a subdata structure of type
+\verb|F R|.  We will not be able to return it from our recursion at
+type \verb|R|, and hence we would not be able to return \verb|([],xs)|
+as desired.  To work around this, we will have our recursion return a value
+of type \verb|SpanF R|:
+\begin{verbatim}
+Inductive SpanF(X : Set) : Set :=
+  SpanNoMatch : SpanF X
+| SpanSomeMatch : list A -> X -> SpanF X.
+\end{verbatim}
+\noindent The idea is that the recursion will signal if it is in the
+one tricky case where \verb|p| does not match the first element, by
+returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
+via \verb|SpanSomeMatch|, a prefix and the suffix at type
+\verb|R|. The prefix will be nonempty, and hence the suffix will be at
+most the tail of \verb|xs|.  This tail is available to the algebra
+in the subdata structure of type \verb|F R|.
+
+Figure~\ref{fig:span} gives the algebra \verb|SpanAlg| for computing
+\verb|span|, and the code for \verb|spanr|.  We elide the proof
+\verb|SpanFunctor| that \verb|SpanF| is indeed a \verb|Functor|,
+and make \verb|X| implicit in the constructors of \verb|SpanF|.
+The type of \verb|SpanAlg| is
+\begin{verbatim}
+Alg (ListF A) C SpanF
+\end{verbatim}
+This states that we are defining an algebra (\verb|Alg|)
+for the \verb|ListF A| functor, with anchor type \verb|C| and carrier
+\verb|SpanF|.  The algebra does not depend on what the anchor type is,
+which is good, as it means we can use \verb|SpanAlg| for top-level or
+subsidiary recursions.
+
+Let us continue through the code for \verb|SpanAlg|
+(Figure~\ref{fig:span}).  We use \verb|rollAlg| to create an algebra
+from something whose type is an application of \verb|AlgF|.  This
+takes in all the components of the recursion universe: the abstract
+type \verb|R|, the \verb|reveal| function (not needed in this case),
+the \verb|fold| function for any subsidiary recursions (also not
+needed here), a function we choose to name \verb|span| for making
+recursive calls, and finally \verb|xs : ListF A R|.  The algebra
+pattern-matches on this \verb|xs|.  In the cases where it is empty or
+where its head (\verb|hd|) does not satisfy \verb|p|, we return
+\verb|SpanNoMatch|.  This signals to the caller that we really wished
+to return \verb|([],xs)|, but could not because we do not have
+\verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
+we recurse on the tail (\verb|tl|).  This is allowed because
+\verb|tl : R| and \verb|span : R -> SpanF R|.  If \verb|span tl|
+returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
+the suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily,
+we have \verb|tl : R| here, so we can do this.  In either case (for
+return value of \verb|span tl|), we add the head to the front of the
+prefix.  We define \verb|spanhr| to invoke the fold function it is
+given, on the algebra (\verb|SpanAlg|).
+
+The final twist is now in the definition of \verb|spanr|.  We call
+\verb|spanhr| on the input \verb|xs : R|.  If \verb|spanhr| returns
+\verb|SpanNoMatch|, then we are supposed to return \verb|([],xs)|,
+which we can do here, because we have \verb|xs : R|.  It was only
+inside the algebra that we lost the information that the subdata
+structure of type \verb|F R| is derived from a value of type \verb|R|.
+If \verb|spanhr| returns \verb|SpanSomeMatch|, then the return value
+gives us the nonempty prefix (\verb|l|) and the suffix (\verb|r|), which we then return.
+
+\begin{figure}
+\begin{verbatim}
+Definition SpanAlg(p : A -> bool)(C : Set)
+  : Alg (ListF A) C SpanF :=
+  rollAlg (fun R reveal fo span xs => 
+     match xs with
+         Nil => SpanNoMatch 
+       | Cons hd tl =>
+          if p hd then
+            match (span tl) with
+              SpanNoMatch => SpanSomeMatch [hd] tl
+            | SpanSomeMatch l r => SpanSomeMatch (hd::l) r
+            end
+          else
+            SpanNoMatch 
+       end).
+
+Definition spanhr{R : Set}(fo:FoldT (Alg (ListF A)) R)
+                 (p : A -> bool)(xs : R) : SpanF R :=
+  fo SpanF SpanFunctor (SpanAlg p R) xs.
+
+Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
+                (p : A -> bool)(xs : R) : list A * R
+  := match spanhr fo p xs with
+       SpanNoMatch => ([],xs)
+     | SpanSomeMatch l r => (l,r)
+     end.
+\end{verbatim}
+\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function, and some functions based on it}
+\label{fig:span}
+\end{figure}
 

### 2022-01-13 paper/paper.tex
added rle example
I think that is all the example recursions I will add.  We have then
just the derivation, interface for subsidiary induction, and examples
of subsidiary induction left to write, and filling in the related
work.  Space is getting tight, so we might need to shrink some earlier
parts a bit.

@@
 \section{Introduction: subsidiary recursion}
+\label{sec:intro}
 
@@
 possibility of further invocations of the outer recursive function.
-Let us see a simple example.  The function \verb|wordsBy|
-(from \verb|Data.List.Extra|) breaks a list into its maximal
-sublists whose elements do not satisfy a predicate \verb|p|.  For
-example, \verb|wordsBy isSpace " good day "| returns
-\verb|["good","day"]|; so \verb|wordsBy isSpace| has the same behavior
-as \verb|words| (from \verb|Data.List|).  Code is in
-Figure~\ref{fig:wordsBy}. The first recursive call,
+Let us see a simple example.  The function \verb|wordsBy| (from
+\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
+elements do not satisfy a predicate \verb|p|.  For example,
+\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|; so
+\verb|wordsBy isSpace| has the same behavior as \verb|words| (from
+\verb|Data.List|).  Code is in Figure~\ref{fig:wordsBy}, using
+also \verb|break| (\verb|Data.List|). Recall that
+\verb|break p| is equivalent to \verb|span (not . p)|: it splits its
+input into the maximal prefix whose elements do not satisfy \verb|p|,
+and the following suffix.  The first recursive call,
 \verb|wordsBy p tl|, is structural.  But in the second, we invoke
 \verb|wordsBy p| on a value obtained from another recursion, namely
-\verb|span|.  This is not allowed under structural termination,
-but will be permitted by subsidiary recursion as derived below.
+\verb|span|.  This is not allowed under structural termination, but
+will be permitted by subsidiary recursion as derived below.
 
@@
   then wordsBy p tl 
-  else let (w,z) = span (not . p) tl in
+  else let (w,z) = break p tl in
         (hd:w) : wordsBy p z
@@
 found as release \verb|itp-2022| (dated prior to the ITP 2022
-deadline) at \url{https://github.com/astump/coq-subsidiary}.
+deadline) at \url{https://github.com/astump/coq-subsidiary}.  The
+paper references files in this codebase, as an aid to the reader
+wishing to peruse the code.
 
@@
 Definition fold : FoldT Alg Subrec.
-Definition rollAlg : forall {C : Set} {X : Set -> Set}, AlgF Alg C X -> Alg C X.
-Definition unrollAlg : forall {C : Set} {X : Set -> Set}, Alg C X -> AlgF Alg C X.
+Definition rollAlg :
+  forall {C : Set} {X : Set -> Set}, AlgF Alg C X -> Alg C X.
+Definition unrollAlg : 
+  forall {C : Set} {X : Set -> Set}, Alg C X -> AlgF Alg C X.
 \end{verbatim}
@@
 and make \verb|X| implicit in the constructors of \verb|SpanF|.
-The type of \verb|SpanAlg| is
+The type of \verb|SpanAlg p C| is
 \begin{verbatim}
@@
 \end{verbatim}
-This states that we are defining an algebra (\verb|Alg|)
-for the \verb|ListF A| functor, with anchor type \verb|C| and carrier
-\verb|SpanF|.  The algebra does not depend on what the anchor type is,
-which is good, as it means we can use \verb|SpanAlg| for top-level or
-subsidiary recursions.
+This states that we are defining an algebra (\verb|Alg|) for the
+\verb|ListF A| functor, with anchor type \verb|C| and carrier
+\verb|SpanF|.  \verb|SpanF| has type \verb|Set -> Set|, as required
+for the carriers of our algebras. The definition of \verb|SpanAlg| is
+actually parametrized by \verb|C|, which is good, as it means we can
+use \verb|SpanAlg| for top-level or subsidiary recursions.
 
@@
 type \verb|R|, the \verb|reveal| function (not needed in this case),
-the \verb|fold| function for any subsidiary recursions (also not
+the fold function (\verb|fo|) for any subsidiary recursions (also not
 needed here), a function we choose to name \verb|span| for making
@@
 \verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
-we recurse on the tail (\verb|tl|).  This is allowed because
-\verb|tl : R| and \verb|span : R -> SpanF R|.  If \verb|span tl|
+we recurse on the tail (\verb|tl : R|) by calling the provided \verb|span : R -> SpanF R|.
+If \verb|span tl|
 returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
@@
 \label{fig:span}
+\end{figure}
+
+We can easily define \verb|break|, in Figure~\ref{fig:break}.  The
+function \verb|breakr| is a version of \verb|break| that can be used
+for subsidiary recursion, similarly to \verb|spanr| for \verb|span|.
+Such a function always takes in a fold function (\verb|fo|) with
+anchor type \verb|R|, which then is used to fold the algebra in
+question.
+
+\begin{figure}
+\begin{verbatim}
+Definition breakr{R : Set}(fo:FoldT (Alg (ListF A)) R)
+                 (p : A -> bool)(xs : R) : list A * R :=
+  spanr fo (fun x => negb (p x)) xs.
+
+Definition break(p : A -> bool)(xs : List A) : list A * List A :=
+  breakr (fold (ListF A)) p xs.
+\end{verbatim}
+\caption{The \texttt{break} function and its more flexible version, \texttt{breakr}, defined in terms of \texttt{spanr} (Figure~\ref{fig:span})}
+\label{fig:break}
+\end{figure}
+
+\subsection{The \texttt{wordsBy} function (\texttt{WordsBy.v})}
+
+Let us now see how to write \verb|wordsBy|, our example function from
+Section~\ref{sec:intro}, using \verb|spanr| as a subsidiary recursion.
+The code is in Figure~\ref{fig:wordsby}, assuming a type
+\verb|A : Set|.  The setup is similar to that for \verb|span|.  We first define
+an algebra \verb|WordsBy|, parametrized by anchor type \verb|C| (and also
+the predicate \verb|p|), of type
+\begin{verbatim}
+Alg (ListF A) C (Const (list (list A)))
+\end{verbatim}
+\noindent This says that \verb|WordsBy p C| is an algebra (\verb|Alg|)
+for the \verb|ListF A| functor, with anchor type \verb|C|, and carrier
+\verb|Const (list (list A))|.  \verb|Const| is the combinator for
+creating the object part of constant functors; \verb|FunConst| creates
+the morphism part (i.e., the \verb|fmap| function).  We use it
+\verb|Const| here and in other examples where the return type of the
+algebra will not depend on its abstract type \verb|R|.  Here, we are
+constructing from scratch a list of lists, so it will not be legal to
+recurse on the list itself, or its (list) elements.  Intead, we just
+use the \verb|list| type of Coq's standard library.
+
+The code for \verb|WordsBy| is, except for the noise of \verb|rollAlg|
+and accepting the components of the recursion universe, essentially
+the same as what we saw in Section~\ref{sec:intro}.  We pattern match
+on \verb|xs : ListF A R|.  Recall that for this function, we are
+trying to drop elements which satisfy \verb|p|, and return a list of
+the sublists between maximal sequences of such elements.  In the
+\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we are
+supposed to drop it and recurse.  This is legal, because \verb|tl : R|
+and \verb|wordsBy : R -> list (list A)|.  In the \verb|else| case,
+we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
+does not satisfy \verb|p|, and the remaining suffix \verb|z|.
+
+Here we see the benefit of our approach.  From Figure~\ref{fig:break},
+the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
+is the anchor type of the provided fold function \verb|fo|.  And
+\verb|fo| has type \verb|FoldT (ListF A) Alg R|, from the definition
+of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
+with \verb|ListF A|).  This means that from the invocation of
+\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
+can indeed apply \verb|wordsBy : R -> list (list A)| to
+\verb|z| to recurse.
+
+\begin{figure}
+\begin{verbatim}
+  Definition WordsBy(p : A -> bool)(C : Set)
+    : Alg (ListF A) C (Const (list (list A))) :=
+    rollAlg (fun R reveal fo wordsBy xs => 
+         match xs with
+           Nil => [] 
+         | Cons hd tl =>
+           if p hd then
+             wordsBy tl
+           else
+             let (w,z) := breakr fo p tl in
+             (hd :: w) :: wordsBy z
+         end).
+
+Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
+                   (p : A -> bool)(xs : R) : list (list A) :=
+  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p R) xs.
+
+Definition wordsBy(p : A -> bool)(xs : List A) : list (list A) :=
+  wordsByr (fold (ListF A)) p xs.
+\end{verbatim}
+\caption{The \texttt{wordsBy} and \texttt{wordsByr} function, defined using an algebra}
+\label{fig:wordsby}
+\end{figure}
+
+\subsection{The \texttt{mapThrough} function (\texttt{MapThrough.v})}
+
+The Haskell library \verb|Data.List.Extra| has a function \verb|repeatedly|, which
+is defined essentially as follows; I have attempted a more informative name:
+\begin{verbatim}
+mapThrough :: (a -> [a] -> (b, [a])) -> [a] -> [b]
+mapThrough f [] = []
+mapThrough f (a:as) = b : mapThrough f as'
+    where (b, as') = f a as
+\end{verbatim}
+\noindent The idea is that the function is like the standard \verb|map| function
+on lists, except that here, the function \verb|f| that we are mapping (or ``mapping
+through'') takes in not just the current element \verb|a|, but also the tail \verb|as|.
+It then returns the value \verb|b| to include in the output list, and whatever other
+list it wishes, upon which \verb|mapThrough| will recurse.
+
+We can write this combinatory using our infrastructure for subsidiary recursion.
+For this to work, we need to supply the mapped function with the fold function
+for \verb|mapThrough|'s recursion.  This is so that the mapped function can
+initiate a subsidiary recursion, returning a value in the abstract type \verb|R|
+of \verb|mapThrough|'s recursion.  So the type we will use for mapped functions
+is:
+\begin{verbatim}
+Definition mappedT(A B : Set) : Set :=
+  forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
+\end{verbatim}
+\noindent This type is more informative than the Haskell type,
+since it shows that the second component of the returned value
+must have type \verb|R|, and hence must be (hereditarily) a tail
+of the input of type \verb|R|.
+
+Given this definition, the code for \verb|mapThrough| and
+\verb|mapThroughr| is in Figure~\ref{fig:mapthrough}.  The code for
+\verb|MapThroughAlg| is very similar (discounting syntax) to the
+Haskell code above.  Here, though, when we call \verb|f|, we must
+supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
+from the definition of \verb|mappedT|, we have that \verb|b : B| and
+\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> List B|
+on \verb|c|.  Note that as we are building up a new list from scratch
+(rather than just extracting some tail of the input list), we just
+return \verb|list B|; we cannot perform further subsidiary recursion
+on the output.
+
+\begin{figure}
+\begin{verbatim}
+Definition MapThroughAlg{B : Set}(f:mappedT A B)
+           (C : Set) : Alg (ListF A) C (Const (list B)) :=
+  rollAlg (fun R reveal fo mapThrough xs => 
+    match xs with
+      Nil => []
+    | Cons hd tl =>
+      let (b,c) := f R fo hd tl in
+        b :: mapThrough c
+    end).
+
+Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
+                      {B : Set}(f:mappedT A B) : R -> list B :=
+  fo (Const (list B)) (FunConst (list B)) (MapThroughAlg f R).
+
+Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B :=
+  mapThroughr (fold (ListF A)) f.
+\end{verbatim}
+\caption{The \texttt{mapThrough} and \texttt{mapThroughr} functions,
+  with their defining algebra}
+\label{fig:mapthrough}
+\end{figure}
+
+\subsection{Run-length encoding}
+
+Using \verb|mapThrough|, we can write a quite concise function for
+\emph{run-length encoding}, a basic data-compression algorithm where
+maximal sequences of $n$ occurrences of element $e$ are summarized by
+the pair $(n,e)$~\cite{datacomp}.  In Haskell, invoking \verb|span| and
+\verb|mapThrough| (defined above), the code is simply
+\begin{verbatim}
+rle :: Eq a => [a] -> [(Int,a)]
+rle = mapThrough compressSpan
+  where compressSpan a as =
+          let (p,s) = span (== a) as in
+            ((1 + length p, a),s)
+\end{verbatim} 
+\noindent (Recall that \verb|(== a)| is a Haskell \emph{section}
+testing its input for equality with \verb|a|.)  The
+\verb|compressSpan| helper function gathers up all elements at the
+start of the tail \verb|as| that are equal to the head \verb|a|.  This
+prefix is returned as \verb|p|, with the remaining suffix as \verb|s|.
+The pair \verb|(1 + length p, a)| is returned to summarize
+\verb|a :: p|.  We then use \verb|mapThrough| to iterate
+\verb|compressSpan| through the suffix \verb|s|.
+
+Assuming \verb|A : Set| and an equality test \verb|eqb : A -> A -> bool| on it,
+code for run-length encoding using our infrastructure is listed in Figure~\ref{fig:rle}.
+The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
+will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
+\verb|compressSpan| has type:
+\begin{verbatim}
+forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> (nat * A) * R.
+\end{verbatim}
+\noindent It will be invoked by the code for \verb|mapThrough| with a
+fold function \verb|fo| with anchor type \verb|R|, and then has the
+responsibility of mapping the tail at type \verb|R| (second input) to
+a result upon which \verb|mapThrough| should recurse (second component
+of the output pair).  Then we define an algebra \verb|RleAlg| by
+supplying \verb|compressSpan| as the function to map through, to
+\verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).  Following the
+pattern we have seen in all the examples above, we may then define
+function \verb|mapThroughr| for subsidiary recursions, and
+\verb|mapThrough| for top-level recursions.
+
+\begin{figure}
+\begin{verbatim}
+Definition compressSpan : mappedT A (nat * A) :=
+  fun R fo hd tl => 
+    let (p,s) := spanr fo (eqb hd) tl in
+       ((succ (length p),hd), s).
+
+Definition RleCarr := Const (list (nat * A)).
+Definition RleAlg(C : Set) : Alg (ListF A) C RleCarr :=
+  MapThroughAlg compressSpan C.
+
+Definition rle(xs : List A) : list (nat * A)
+  := @fold (ListF A) RleCarr (FunConst (list (nat * A))) (RleAlg (List A)) xs.
+\end{verbatim}
+\caption{The function \texttt{rle} for run-length encoding, and the algebra \texttt{RleAlg} defining it
+in terms of \texttt{MapThroughAlg} (Figure~\ref{fig:mapthrough})}
+\label{fig:rle}
 \end{figure}

### 2022-01-13 paper/paper.tex
minor

@@
 \verb|wordsBy isSpace| has the same behavior as \verb|words| (from
-\verb|Data.List|).  Code is in Figure~\ref{fig:wordsBy}, using
-also \verb|break| (\verb|Data.List|). Recall that
-\verb|break p| is equivalent to \verb|span (not . p)|: it splits its
-input into the maximal prefix whose elements do not satisfy \verb|p|,
-and the following suffix.  The first recursive call,
-\verb|wordsBy p tl|, is structural.  But in the second, we invoke
-\verb|wordsBy p| on a value obtained from another recursion, namely
-\verb|span|.  This is not allowed under structural termination, but
-will be permitted by subsidiary recursion as derived below.
+\verb|Data.List|).  Code is in Figure~\ref{fig:wordsBy}.  The first
+recursive call, \verb|wordsBy p tl|, is structural.  But in the
+second, we invoke \verb|wordsBy p| on a value obtained from another
+recursion, namely \verb|span|.  This is not allowed under structural
+termination, but will be permitted by subsidiary recursion as derived
+below.
 
@@
   then wordsBy p tl 
-  else let (w,z) = break p tl in
+  else let (w,z) = span (not . p) tl in
         (hd:w) : wordsBy p z
@@
 
-\subsection{Run-length encoding}
+\subsection{Run-length encoding (\texttt{Rle.v})}
 

### 2022-01-14 paper/paper.tex
working on Section 4 (implementation of Subrec)

@@
 
+Definition List := Subrec ListF .
 Definition inList : ListF List -> List := inn ListF.
@@
   defines the signature functor \verb|ListF|, shown in
-  Figure~\ref{fig:listf}.  Using \verb|Subrec|, we then get a type
-  \verb|List|.  This is not to be confused with the type \verb|list|
-  of lists in Coq's standard library.  As noted previously, our
-  development is meant to be used in extension of existing inductive
-  datatypes, not replacing them.  The figure also shows constructors
-  \verb|mkNil| and \verb|mkCons| for \verb|List|, and types for
-  conversion functions between \verb|List| and \verb|list| (see
-  Section~\ref{sec:deriv} for the code).
+  Figure~\ref{fig:listf}.  We then define \verb|List| to be
+  \verb|Subrec|, with the instantiation of \verb|F| to \verb|ListF A|.
+  This type \verb|List| is not to be confused with the type
+  \verb|list| of lists in Coq's standard library.  As noted
+  previously, our development is meant to be used in extension of
+  existing inductive datatypes, not replacing them.  The figure also
+  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
+  and types for conversion functions between \verb|List| and
+  \verb|list| (see Section~\ref{sec:deriv} for the code).
 
@@
 
-We can write this combinatory using our infrastructure for subsidiary recursion.
+We can write this combinator using our infrastructure for subsidiary recursion.
 For this to work, we need to supply the mapped function with the fold function
@@
 
-Using \verb|mapThrough|, we can write a quite concise function for
+Using \verb|mapThrough|, we can quite concisely implement
 \emph{run-length encoding}, a basic data-compression algorithm where
@@
 
+Let us now consider the implementation of the
+interface we have used for the preceding examples.
+The first step is our weakened form of positive-recursive
+types.
+
 \subsection{Retractive-positive recursive types}
@@
 types, to allow algebras to accept fold functions that themselves
-require algebras, and also for the definition of \verb|Subrec|.  But
-as recalled already, full positive-recursive types are incompatible
-with Coq's type theory~\cite{coquand88}.  It is worth noting that one
-can impose some restrictions on large eliminations which then allow
-positive-recursive types~\cite{blanqui05}.  This approach would
-require changing the underlying theory.  To avoid this, we here take a
-different approach, exploiting Coq's impredicative polymorphism.
+require algebras, and also for the definition of \verb|Subrec| (which
+we will see in more detail in the next section).  As already noted,
+full positive-recursive types are incompatible with Coq's type
+theory~\cite{coquand88}.  One can impose some restrictions on large
+eliminations which then enable positive-recursive
+types~\cite{blanqui05}.  This approach would require changing the
+underlying theory.  To avoid this, we here take a different approach,
+exploiting Coq's impredicative polymorphism.
 
@@
 
-Noncanonicity leads to some issues, as we turn next to the problem of
-inductive reasoning about subsidiary recursions.  With some care,
-however, we can avoid pitfalls, leaving us with a form of
-positive-recursive type that enables our definitions to go through.
+Noncanonicity must be handled carefully when reasoning about functions
+defined with our interface.  We will see examples in
+Section~\ref{sec:examplesi}.  First, though, let us complete the
+exposition of our implementation of subsidiary recursion.
+
+\subsection{The implementation of \texttt{Subrec} (\texttt{Subrec.v})}
+
+The type \verb|Subrec| is defined in Figure~\ref{fig:subrec}, as a
+fixed-point of \verb|SubrecF : Set -> Set|.  We take this fixed-point
+with \verb|Mu|, discussed in the previous section, and obtain \verb|roll|
+and \verb|unroll| functions between \verb|SubrecF Subrec| and \verb|Subrec|.
+Unrolling \verb|Subrec| gives us the type
+\begin{verbatim}
+forall (X : Set -> Set) (FunX : Functor X), Alg Subrec X -> X Subrec
+\end{verbatim}
+\noindent So we see that \verb|Subrec| is the type of functions which,
+for all algebras with anchor type \verb|Subrec| and functorial carrier
+\verb|X|, compute a value of type \verb|X Subrec|.  This is a generalization
+of the functor-generic type for the Church encoding:
+\[
+\forall\ X.\ \textit{Alg}\ X \to X
+\]
+\noindent where $\textit{Alg}\ X$ is $F\ X \to X$.  
+We elide the implementation of the \verb|roll| and \verb|unroll| functions, but
+note that \verb|unroll| makes use of functoriality of carriers \verb|X|.
+
+\begin{figure}
+\begin{verbatim}
+Definition SubrecF(C : Set) := 
+  forall (X : Set -> Set) (FunX : Functor X), Alg C X -> X C.
+Definition Subrec := Mu SubrecF.
+Definition roll: SubrecF Subrec -> Subrec.
+Definition unroll: Subrec -> SubrecF Subrec.
+\end{verbatim}
+\caption{Definition of \texttt{Subrec} as a fixed-point of \texttt{SubrecF}}
+\label{fig:subrec}
+\end{figure}
+
+The rest of the interface for \verb|Subrec| is shown in
+Figure~\ref{fig:subrecb}.  We have \verb|fold|, which is a fold
+function with anchor type \verb|Subrec|.  To fold an algebra
+\verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
+\verb|FunX|) over \verb|d : Subrec|, \verb|fold| \verb|unroll|s the
+definition of \verb|Subrec| and applies that to the algebra (with its
+carrier).
+
+More interesting is the definition of \verb|inn|.
+
+\begin{figure}
+\begin{verbatim}
+Definition fold : FoldT Alg Subrec := 
+  fun X FunX alg d => unroll d X FunX alg.
+
+Definition inn : F Subrec -> Subrec :=
+  fun d => roll (fun X xmap alg =>
+                 unrollAlg alg Subrec (fun x => x) fold (fold X xmap alg) d).
+
+Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
+  fo F FunF (rollAlg (fun _ _ _ _ d => d)).
+\end{verbatim}
+\caption{The rest of the interface for \texttt{Subrec}}
+\label{fig:subrecb}
+\end{figure}
 
@@
 support compositional termination checking; an accessible source
-is~\cite{mendler91}.  This recursor has type
+is~\cite{mendler91}.  He introduces a functor-generic recursor of type
 \[

### 2022-01-15 paper/paper.tex
almost done, but paper is too long
I skimped on the example proofs and still I am a page over without
writing a conclusion or finishing the related work section.  So I will
need to do some trimming once those last things are written.

@@
 required. Based on this derivation, we present several example
-functions like \verb|wordsBy|, and prove theorems about them.  For
-example, we prove the expected property that the sublists returned by
-\verb|wordsBy| consist of elements satisfying \verb|not . p|.
-For another, we give a definition of run-length encoding as a
-subsidiary recursion using \verb|span|, and prove that encoding and
-then decoding returns the original list.  Our approach applies to the
-standard datatypes in the Coq library, and does not require switching
-libraries or datatype definitions.
+functions like \verb|wordsBy|, and prove theorems about them.  A nice
+example is a definition of run-length encoding using \verb|span| as a
+subsidiary recursion, where we prove that encoding and then decoding
+returns the original list.  Our approach applies to the standard
+datatypes in the Coq library, and does not require switching libraries
+or datatype definitions.
 
@@
 \subsection{The recursion universe}
+\label{sec:recu}
 
@@
   general, to use our development to get subsidiary recursion over
-  some datatype, one will have a similar ``shim'' file.  The file
+  some datatype, one will have a similar ``shim'' file.  For space
+  reasons, we just give the example of lists.  The file
   defines the signature functor \verb|ListF|, shown in
@@
 \subsection{The implementation of \texttt{Subrec} (\texttt{Subrec.v})}
+\label{sec:subrecimpl}
 
@@
 
-More interesting is the definition of \verb|inn|.
+More interesting is the definition of \verb|inn|, which is the
+critical point where the recursion universe is implemented.  To create
+a value of type \verb|Subrec| from data of type \verb|F Subrec|, the
+definition of \verb|inn| \verb|roll|s a value of type
+\verb|SubrecF Subrec| (we saw this type unfolded at the start of
+Section~\ref{sec:subrecimpl}).  This value takes in a carrier
+\verb|X|, its fmap function \verb|xmap|, and an algebra \verb|alg|
+with that carrier.  Note that the anchor type of this algebra is
+\verb|Subrec|.  It will then call \verb|alg| (after \verb|unroll|ing
+it) with implementations for the components of the recursion universe
+(cf. Section~\ref{sec:recu}, also Figure~\ref{fig:algf}):
+\begin{itemize}
+\item \verb|Subrec| is passed as the value for the abstract type \verb|R|.
+\item the identity function is passed as the value for \verb|reveal : R -> Subrec| (since \verb|R| has
+  been instantiated to \verb|Subrec|).
+\item The function \verb|fold|, which expects an algebra with anchor type \verb|Subrec|, is passed as
+  the fold function of type \verb|FoldT Alg R|.
+\item For the \verb|eval : R -> X R| function, we pass \verb|(fold X xmap alg) : Subrec -> X Subrec|.
+\item For the subdata structure of type \verb|F R|, we pass \verb|d : F Subrec| (again, since we
+  are instantiating \verb|R| with \verb|Subrec|).
+\end{itemize}
+
+Finally, Figure~\ref{fig:subrecb} defines \verb|out| as a subsidiary
+recursion, given any fold function with its anchor type \verb|R|.  The
+code for \verb|out| just folds an algebra over the input of type
+\verb|R|, where that algebra simply returns the subdata structure it
+is given.  Outside the recusion, this has type \verb|F R|; inside the
+recursion it has type \verb|F R'| where \verb|R'| is the abstract type
+of the subsidiary recursion (named in the figure just for discussion
+here).  So \verb|out| implements the idea that unfolding an abstract
+type one step is just a trivial case of subsidiary recursion.
 
@@
 Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
-  fo F FunF (rollAlg (fun _ _ _ _ d => d)).
+  fo F FunF (rollAlg (fun R' _ _ _ d => d)).
 \end{verbatim}
@@
 
-\section{Interface for subsidiary induction}
+\subsection{Converting between \texttt{list} and \texttt{List} (\texttt{List.v})}
+
+Figure~\ref{fig:listb} gives the definitions (specified in
+Figure~\ref{fig:listf} above) for conversion functions between
+\texttt{list} (the list type of Coq's standard library) and our
+\texttt{List} type for subsidiary recursion on lists.  To convert a
+\verb|list| to a \verb|List|, we use \verb|listFold| to recursively
+apply the function \verb|inn| (Figure~\ref{fig:subrecb}) throughout a
+\verb|list|.  This \verb|listFold| provides a functorial view of the
+usual fold-right function on lists.  To convert a \verb|List| to a
+\verb|list|, we use our \verb|fold| function
+(Figure~\ref{fig:subrecb}) to fold an algebra over the input
+\verb|List|.  This algebra does an \verb|fmap| (for \verb|ListF|) with
+\verb|eval : R -> list A|.  The resulting value has type
+\verb|ListF (list A)|.  We may then apply \verb|listIn| to produce a
+\verb|list A|.  This \verb|listIn| function converts from
+\verb|ListF (list A)| to \verb|list A| in the obvious way, converting
+\verb|Nil| to the empty list, and \verb|Cons| to an application of
+\verb|::|.  These functions were not necessary in implementing
+examples of subsidiary recursion above, but they will be needed
+for reasoning about them.  We turn to this topic next.
+
+\begin{figure}
+\begin{verbatim}
+Fixpoint listFold(l : list A){X : Set}(alg : ListF X -> X) : X :=
+  match l with
+    nil => alg Nil
+  | cons hd tl => alg (Cons hd (listFold tl alg))
+  end.
+
+Definition toList (xs : list A) : List := listFold xs (inn ListF).
+
+Definition listIn(d : ListF (list A)) : list A :=
+  match d with
+    Nil => []
+  | Cons hd tl => hd :: tl
+  end.
+
+Definition fromList : List -> list A :=
+  fold ListF (Const (list A)) (FunConst (list A))
+       (rollAlg (fun R reveal fo eval fr => listIn (fmap eval fr))) .
+\end{verbatim}
+\caption{Additional functions for the \texttt{List} type (\texttt{Subrec ListF})}
+\label{fig:listb}
+\end{figure}
+
+\section{Interface for subsidiary induction (\texttt{Subreci.v})}
 \label{sec:interfacei}
+
+We have seen how to write subsidiary recursions in Coq.  But can one
+reason about these?  To wrap up this paper, we will see an interface
+for subsidiary induction in Coq, and example proofs written using this
+interface.  Subsidiary induction is written just as the natural
+extension of subsidiary recursion, which worked over \verb|Set|s, to
+\verb|Subrec|-predicates.  The development is parametrized by a
+functor \verb|F| and a functor
+\verb|Fi : (Subrec -> Prop) -> (Subrec -> Prop)| over
+\verb|Subrec|-indexed propositions (i.e., predicates).  Just as
+functors need an \verb|fmap| function, we here need an indexed
+version, of type \verb|fmapiT Subrec Fi| (definition elided.)
+
+The central definitions for the type \verb|Subreci : Subrec -> Prop|
+are given in Figure~\ref{fig:subreci}.  Where having a value \verb|x|
+of \verb|Subrec| entitles us to define subsidiary recursions to
+inhabit types \verb|X Subrec|, a value of type \verb|Subreci x| lets
+us prove properties of \verb|x| by subsidiary induction.  Briefly:
+\verb|kMo| is the kind for \emph{motives}, namely predicates on
+\verb|Subrec|~\cite{mcbride00}.  \verb|KAlgi| is the kind for indexed
+algebras.  \verb|FoldTi| is the indexed version of \verb|FoldT|: it is
+expresses provability of \verb|X C| for \verb|d|, based on an indexed
+algebra and a value of type \verb|C d|, where \verb|C| is the anchor
+type, now indexed.  \verb|AlgFi| and \verb|Algi| are indexed versions
+of the algebras we saw for recursion.  The \verb|eval| function
+(Figure~\ref{fig:algf}) has now become an induction hypothesis: given
+any \verb|d| where \verb|R d| holds, \verb|ih| proves \verb|X R d|.  A
+value of type \verb|R d| is thus a license to induct on \verb|d|.
+Finally, the algebra is given a subdata structure indexed by
+\verb|d : Subrec|, and must produce a proof of \verb|X R d|.  \verb|Subreci|
+is defined as the suitably indexed fixed-point of \verb|SubrecFi|, which
+is the natural indexed version of \verb|SubrecF|.
+
+\begin{figure}
+\begin{verbatim}
+Definition kMo := Subrec -> Prop.
+Definition KAlgi := kMo -> (kMo -> kMo) -> Set.
+Definition FoldTi(alg : KAlgi)(C : kMo) : kMo :=
+  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), alg C X -> C d -> X C d.
+
+Definition AlgFi(A: KAlgi)(C : kMo)(X : kMo -> kMo) : Set :=
+  forall (R : kMo)
+    (reveal : (forall (d : Subrec), R d -> C d))        
+    (fo : (forall (d : Subrec), FoldTi A R d))
+    (ih : (forall (d : Subrec), R d -> X R d))
+    (d : Subrec),
+    Fi R d -> X R d.
+
+Definition Algi := MuAlgi Subrec AlgFi.
+
+Definition SubrecFi(C : kMo) : kMo := 
+  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), Algi C X -> X C d.
+Definition Subreci := Mui Subrec SubrecFi.
+
+Definition foldi(i : Subrec) : FoldTi Algi Subreci i.
+Definition inni(i : Subrec)(fd : Fi Subreci i) : Subreci i.
+\end{verbatim}
+\caption{Interface for subsidiary induction}
+\label{fig:subreci}
+\end{figure}
+
+For lists, we instantiate \verb|Fi| with \verb|ListFi|, shown in
+Figure~\ref{fig:listfi}.  This is just the indexed version of
+\verb|ListF|.  Given a \verb|list A|, \verb|toListi| returns
+a value of type \verb|Listi (toList xs)|.  This can be understood
+as saying that for any list (from Coq's standard library),
+we can reason by subsidiary induction to prove properties of \verb|toList xs|.
+We also introduce an abbreviation \verb|ListFoldTi| for the type
+of indexed fold functions over lists.
+
+\begin{figure}
+\begin{verbatim}
+Definition lkMo := List -> Prop.
+
+Inductive ListFi(R : lkMo) : lkMo :=
+  nilFi : ListFi R mkNil
+| consFi : forall (h : A)(t : List), R t -> ListFi R (mkCons h t).
+
+Definition Listi := Subreci ListF ListFi.
+Definition toListi(xs : list A) : Listi (toList xs) := listFoldi xs Listi inni.
+Definition ListFoldTi(R : List -> Prop)(d : List) : Prop :=
+  FoldTi ListF (Algi ListF ListFi) R d.
+\end{verbatim}
+\caption{The indexed version \texttt{ListFi} of \texttt{ListF}}
+\label{fig:listfi}
+\end{figure}
 
@@
 \label{sec:examplesi}
+
+For proving the main theorem about run-length encoding, we need
+several lemmas about \verb|span|, shown in Figure~\ref{fig:spanlem}.
+For lack of space, we just state the properties.  The first says that
+appending the results of a call to span returns the original list
+(module some conversions to \verb|list| from \verb|List|).  The second
+uses the inductive type \verb|Forall| from Coq's standard library to
+state that all the elements of the prefix returned by \verb|span|
+satisfy \verb|p|.  These lemmas are proved using an indexed algebra
+where the indexed anchor type is not used (so the carriers are
+constant indexed-functor returning the types shown).  But for the
+third, \verb|GuardPres|, we are showing that whenever \verb|spanh|
+(see Figure~\ref{fig:span}) returns a suffix \verb|r|, that suffix
+satisfies the indexed anchor type.  This enables us to invoke an
+outer induction hypothesis on this suffix, when using \verb|span|
+subsidiarily.
+
+\begin{figure}
+\begin{verbatim}
+Definition SpanAppendF(p : A -> bool)(xs : List A) : Prop :=
+  forall (l : list A)(r : List A) ,
+    span p xs = (l,r) ->
+    fromList xs = l ++ (fromList r).
+
+Definition spanForallF(p : A -> bool)(xs : List A) : Prop :=
+  forall (l : list A)(r : List A),
+    span p xs = (l,r) ->
+    Forall (fun a => p a = true) l.
+
+Definition GuardPresF(p : A -> bool)(S : List A -> Prop)(xs : List A) : Prop :=
+  forall (l : list A)(r : List A),
+    spanh p xs = SpanSomeMatch l r ->
+    S r.
+\end{verbatim}
+\caption{Statements of three lemmas about \texttt{span}}
+\label{fig:spanlem}
+\end{figure}
+
+Using these lemmas, we can write a short proof by subsidiary induction of:
+
+\begin{verbatim}
+Theorem RldRle (xs : list A): rld (rle (toList xs)) = xs.  
+\end{verbatim}
+
+Finally, as promised, a note on noncanonicity. When proving properties about subsidiary
+recursions on \verb|xs : List A|, one should be aware that nothing prevents
+the property from being applied to noncanonical \verb|List|s.  For example,
+suppose we wish to prove that if all elements of a  list satisfy \verb|p|, then
+the suffix returned by \verb|span| is empty.  It is dangerous to phrase
+this as ``the suffix equals \verb|mkNil|'', because for a noncanonical input \verb|xs|,
+\verb|span| will return that same  noncanonical \verb|xs| as the suffix (and so
+it may be a noncanonical empty list, not equal to \verb|mkNil|).  The solution
+in this case is to use a function \verb|getNil| (\verb|List.v|) that computes
+an empty list from \verb|xs|.  So the statement that one can prove is:
+
+\begin{verbatim}
+Definition spanForall2F(p : A -> bool)(xs : List A) : Prop :=
+  Forall (fun a => p a = true) (fromList xs) ->
+  span p xs = (fromList xs, getNil xs).
+\end{verbatim}
 
@@
 
+\section{Conclusion}
+
 \bibliography{biblio}

### 2022-01-15 paper/paper.tex
polishing, shrinking
Now just around 16 pages, still need conclusion and revised related work.

@@
 inductive datatypes~\cite{agda,coq,isabelle-hol,lean}.  Termination is
-usually enforced by a syntactic check for structural decrease.  This
-structural termination is sufficient for many basic functions.  For
-example, the well-known \texttt{span} function from Haskell's standard
-library (\verb|Data.List|) takes a list and returns a pair of the
-maximal prefix satisfying a given predicate \verb|p|, and the
-remaining suffix:
+usually enforced by a syntactic check for structural decrease, which
+is sufficient for many basic functions.  For example, the
+\texttt{span} function from Haskell's prelude (\verb|Data.List|) takes
+a list and returns a pair of the maximal prefix whose elements satisfy
+a given predicate \verb|p|, and the remaining suffix:
 \begin{verbatim}
@@
 \noindent The sole recursive call is \verb|span p xs|, and it occurs
-in a clause where the input list is of the form \verb|x:xs|.  So the
-input to the recursive call is a subdatum of the input, and hence this
-definition is structurally decreasing.  In the appropriate syntax, it
-can be accepted without additional effort by all the mentioned
-provers.
+in a clause where the input list is of the form \verb|x:xs|.  Hence it
+is structurally decreasing.  In the appropriate syntax, this
+definition can be accepted without additional effort by all the
+mentioned provers.
 
@@
 possibility of further invocations of the outer recursive function.
-Let us see a simple example.  The function \verb|wordsBy| (from
-\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
+Let us see a simple example.  The function \verb|wordsBy|
+(\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
 elements do not satisfy a predicate \verb|p|.  For example,
-\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|; so
-\verb|wordsBy isSpace| has the same behavior as \verb|words| (from
-\verb|Data.List|).  Code is in Figure~\ref{fig:wordsBy}.  The first
-recursive call, \verb|wordsBy p tl|, is structural.  But in the
-second, we invoke \verb|wordsBy p| on a value obtained from another
-recursion, namely \verb|span|.  This is not allowed under structural
-termination, but will be permitted by subsidiary recursion as derived
-below.
+\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|.
+Code is in Figure~\ref{fig:wordsBy}.  Recall that
+\verb|break p| is equivalent to \verb|span (not . p)|. The first recursive call,
+\verb|wordsBy p tl|, is structural.  But in the second, we invoke
+\verb|wordsBy p| on a value obtained from another recursion, namely
+\verb|span|.  This is not allowed under structural termination, but
+will be permitted by subsidiary recursion.
 
@@
 subsidiary recursion and induction in Coq.  We should emphasize that
-this is a derivation of this recursion scheme within the type theory
-of Coq.  No axioms or other modifications to Coq of any kind are
-required. Based on this derivation, we present several example
-functions like \verb|wordsBy|, and prove theorems about them.  A nice
-example is a definition of run-length encoding using \verb|span| as a
-subsidiary recursion, where we prove that encoding and then decoding
-returns the original list.  Our approach applies to the standard
-datatypes in the Coq library, and does not require switching libraries
-or datatype definitions.
-
-An important technical novelty of our approach is a derivation of a
-weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
-restrict datatypes $D$ to be strictly positive: in the type for any
-constructor of $D$, $D$ cannot occur to the left of any arrows.  Our
-derivation needs to use positive-recursive types, where $D$ may occur
-to the left of an even number (only) of arrows.  Coq requires strict
-positivity because in the presence of other features of Coq's theory,
-full positive-recursive types lead to a paradox~\cite{coquand88}.  We
-present a way to derive a weakened form of positive-recursive type
-that is sufficient for our examples (Section~\ref{sec:mu}).  The
-weakening is to require only that $F\ \mu$ is a retract of $\mu$,
-where $\mu$ is the recursive type and $F\ \mu$ its one-step unfolding.
-Usually these types are isomorphic.  Hence, we dub these
-\textbf{retractive-positive} recursive types.  This weakening has the
-negative consequence of leading to a form of noncanonicity, but we
-will see how to work around this.  Our definition of
+this is a derivation within the type theory of Coq, and requires no
+axioms or other modifications to Coq. Using this derivation, we
+present several example functions like \verb|wordsBy|, and prove
+theorems about them.  A nice example is a definition of run-length
+encoding using \verb|span| as a subsidiary recursion, where we prove
+that encoding and then decoding returns the original list.  Our
+approach applies to the standard datatypes in the Coq library, and
+does not require switching libraries or datatype definitions.
+
+An important technical novelty is a derivation of a weakened form of
+positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
+datatypes $D$ to be strictly positive: in the type for any constructor
+of $D$, $D$ cannot occur to the left of any arrows.  Our derivation
+needs to use positive-recursive types, where $D$ may occur to the left
+of an even number (only) of arrows.  We present a way to derive a
+weakened form of positive-recursive type that is sufficient for our
+examples (Section~\ref{sec:mu}).  The weakening is to require only
+that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
+and $F\ \mu$ its one-step unfolding.  Usually these types are
+isomorphic.  Hence, we dub these \textbf{retractive-positive}
+recursive types.  This weakening leads to noncanonical elements of
+$\mu$, but we will see how to work around this.  Our definition of
 retractive-positive recursive types makes essential use of impredicate
-quantification, and hence cannot be soundly recapitulated in a
-predicative theory like Agda's.
+quantification, and hence is not available in predicative theories
+like Agda's.
 
@@
 coding against this interface retains a similar feel to structural
-recursions.  Unlike with structural termination, though, the interface
-here is type-based and hence compositional.  As we will see, it
-supports nested and higher-order recursions.
+recursion.  Unlike with structural termination, though, the interface
+here is type-based and hence compositional.
 
@@
 because we may recurse (via \verb|eval|) on any value of type
-\verb|R|.
-
-Some points must still be explained, particularly why \verb|X| has type \verb|Set -> Set|,
-and the definition of \verb|FoldT|.  Let us see these and other details next.
+\verb|R|.  Some points must still be explained: why \verb|X| has type \verb|Set -> Set|,
+and the definition of \verb|FoldT|.  Let us see these details next.
 
@@
   some datatype, one will have a similar ``shim'' file.  For space
-  reasons, we just give the example of lists.  The file
-  defines the signature functor \verb|ListF|, shown in
-  Figure~\ref{fig:listf}.  We then define \verb|List| to be
-  \verb|Subrec|, with the instantiation of \verb|F| to \verb|ListF A|.
-  This type \verb|List| is not to be confused with the type
-  \verb|list| of lists in Coq's standard library.  As noted
-  previously, our development is meant to be used in extension of
-  existing inductive datatypes, not replacing them.  The figure also
-  shows constructors \verb|mkNil| and \verb|mkCons| for \verb|List|,
-  and types for conversion functions between \verb|List| and
-  \verb|list| (see Section~\ref{sec:deriv} for the code).
+  reasons, we just give the example of lists.  The file defines the
+  signature functor \verb|ListF|, shown in Figure~\ref{fig:listf}.  We
+  then define \verb|List| to be \verb|Subrec|, with the instantiation
+  of \verb|F| to \verb|ListF A|.  This type \verb|List| is not to be
+  confused with the type \verb|list| of lists in Coq's standard
+  library.  As noted previously, our development is meant to be used
+  in extension of existing inductive datatypes, not replacing them.
+  The figure also shows constructors \verb|mkNil| and \verb|mkCons|
+  for \verb|List|, and types for conversion functions between
+  \verb|List| and \verb|list|; one direction uses Coq's structural
+  recursion, the other uses our subsidiary recursion (code elided).
 
@@
 \verb|Alg|.  This type-constructor \verb|Alg| is a fixed-point of the
-type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|
-(Section~\ref{sec:mu}), which implements our retractive-positive
-recursive types at kind \verb|KAlg|.  Using \verb|Alg| will require
-that \verb|AlgF| only uses its parameter \verb|Alg| positively.  We
-will confirm this shortly.
+type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|, which
+implements our retractive-positive recursive types
+(Section~\ref{sec:mu}) at kind \verb|KAlg|.  Using \verb|Alg| will
+require that \verb|AlgF| only uses its parameter \verb|Alg|
+positively.  We will confirm this shortly.
 
@@
 
-The type \verb|FoldT Alg C| is the type for fold functions which apply
+\verb|FoldT Alg C| is the type for fold functions which apply
 algebras of type \verb|Alg| to data of type \verb|C|, which we have
@@
 anchor type will instead by the abstract type \verb|R| for the outer recursion.
-
 The variable \verb|Alg| occurs only positively (but not strictly
@@
 type \verb|list A * R|.  The first component of this pair is
-going to be built up from scratch, and so cannot have type \verb|R|.
+going to be built up from scratch, and so cannot have type \verb|R|;
+we cannot statically ensure that outer recursions on it are legal.
 But the second component will be a subdatum of the input list,
-and so can still have type \verb|R|.  This will enable outer recursions
-to continue on that component.  So we want: 
+and so can still have type \verb|R|, enabling outer recursive calls.  So we want: 
 \begin{verbatim}
@@
 as desired.  To work around this, we will have our recursion return a value
-of type \verb|SpanF R|:
+of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
 \begin{verbatim}
@@
 
-Figure~\ref{fig:span} gives the algebra \verb|SpanAlg| for computing
-\verb|span|, and the code for \verb|spanr|.  We elide the proof
-\verb|SpanFunctor| that \verb|SpanF| is indeed a \verb|Functor|,
-and make \verb|X| implicit in the constructors of \verb|SpanF|.
-The type of \verb|SpanAlg p C| is
+\begin{figure}
+\begin{verbatim}
+Definition SpanAlg(p : A -> bool)(C : Set)
+  : Alg (ListF A) C SpanF :=
+  rollAlg (fun R reveal fo span xs => 
+     match xs with
+         Nil => SpanNoMatch 
+       | Cons hd tl =>
+          if p hd then
+            match (span tl) with
+              SpanNoMatch => SpanSomeMatch [hd] tl
+            | SpanSomeMatch l r => SpanSomeMatch (hd::l) r
+            end
+          else
+            SpanNoMatch 
+       end).
+\end{verbatim}
+\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function}
+\label{fig:spanalg}
+\end{figure}
+
+Figure~\ref{fig:spanalg} gives the algebra \verb|SpanAlg| for computing
+\verb|span|.  The type of \verb|SpanAlg p C| is
 \begin{verbatim}
@@
 return value of \verb|span tl|), we add the head to the front of the
-prefix.  We define \verb|spanhr| to invoke the fold function it is
-given, on the algebra (\verb|SpanAlg|).
-
-The final twist is now in the definition of \verb|spanr|.  We call
-\verb|spanhr| on the input \verb|xs : R|.  If \verb|spanhr| returns
-\verb|SpanNoMatch|, then we are supposed to return \verb|([],xs)|,
-which we can do here, because we have \verb|xs : R|.  It was only
-inside the algebra that we lost the information that the subdata
-structure of type \verb|F R| is derived from a value of type \verb|R|.
-If \verb|spanhr| returns \verb|SpanSomeMatch|, then the return value
-gives us the nonempty prefix (\verb|l|) and the suffix (\verb|r|), which we then return.
-
-\begin{figure}
-\begin{verbatim}
-Definition SpanAlg(p : A -> bool)(C : Set)
-  : Alg (ListF A) C SpanF :=
-  rollAlg (fun R reveal fo span xs => 
-     match xs with
-         Nil => SpanNoMatch 
-       | Cons hd tl =>
-          if p hd then
-            match (span tl) with
-              SpanNoMatch => SpanSomeMatch [hd] tl
-            | SpanSomeMatch l r => SpanSomeMatch (hd::l) r
-            end
-          else
-            SpanNoMatch 
-       end).
-
+prefix.
+
+\verb|SpanAlg| is used in the definition of \verb|spanhr|, in
+Figure~\ref{fig:span}.  This function invokes the fold function it is
+given, on \verb|SpanAlg|.  The final twist is now in the definition of
+\verb|spanr|.  We call \verb|spanhr| on the input \verb|xs : R|.  If
+\verb|spanhr| returns \verb|SpanNoMatch|, then we are supposed to
+return \verb|([],xs)|, which we can do here, because we have
+\verb|xs : R|.  It was only inside the algebra that we lost the
+information that the subdata structure of type \verb|F R| is derived
+from a value of type \verb|R|.  If \verb|spanhr| returns
+\verb|SpanSomeMatch|, then the return value gives us the nonempty
+prefix (\verb|l|) and the suffix (\verb|r|), which we then return.  We
+also define a version of \verb|break| for subsidiary recursion.
+
+\begin{figure}
+\begin{verbatim}
 Definition spanhr{R : Set}(fo:FoldT (Alg (ListF A)) R)
@@
      end.
-\end{verbatim}
-\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function, and some functions based on it}
-\label{fig:span}
-\end{figure}
-
-We can easily define \verb|break|, in Figure~\ref{fig:break}.  The
-function \verb|breakr| is a version of \verb|break| that can be used
-for subsidiary recursion, similarly to \verb|spanr| for \verb|span|.
-Such a function always takes in a fold function (\verb|fo|) with
-anchor type \verb|R|, which then is used to fold the algebra in
-question.
-
-\begin{figure}
-\begin{verbatim}
+
 Definition breakr{R : Set}(fo:FoldT (Alg (ListF A)) R)
@@
   spanr fo (fun x => negb (p x)) xs.
-
-Definition break(p : A -> bool)(xs : List A) : list A * List A :=
-  breakr (fold (ListF A)) p xs.
-\end{verbatim}
-\caption{The \texttt{break} function and its more flexible version, \texttt{breakr}, defined in terms of \texttt{spanr} (Figure~\ref{fig:span})}
-\label{fig:break}
+\end{verbatim}
+\caption{Functions derived from \texttt{SpanAlg}}
+\label{fig:span}
 \end{figure}
@@
 Let us now see how to write \verb|wordsBy|, our example function from
-Section~\ref{sec:intro}, using \verb|spanr| as a subsidiary recursion.
+Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.
 The code is in Figure~\ref{fig:wordsby}, assuming a type
@@
 creating the object part of constant functors; \verb|FunConst| creates
-the morphism part (i.e., the \verb|fmap| function).  We use it
-\verb|Const| here and in other examples where the return type of the
+the morphism part (i.e., the \verb|fmap| function).  We use 
+\verb|Const| where the return type of the
 algebra will not depend on its abstract type \verb|R|.  Here, we are
 constructing from scratch a list of lists, so it will not be legal to
-recurse on the list itself, or its (list) elements.  Intead, we just
+recurse on the list itself, or its (list) elements.  So we just
 use the \verb|list| type of Coq's standard library.
@@
 
-Here we see the benefit of our approach.  From Figure~\ref{fig:break},
+Here we see the benefit of our approach.  From Figure~\ref{fig:span},
 the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
@@
 
-The Haskell library \verb|Data.List.Extra| has a function \verb|repeatedly|, which
-is defined essentially as follows; I have attempted a more informative name:
+\begin{figure}
 \begin{verbatim}
@@
 \end{verbatim}
-\noindent The idea is that the function is like the standard \verb|map| function
-on lists, except that here, the function \verb|f| that we are mapping (or ``mapping
-through'') takes in not just the current element \verb|a|, but also the tail \verb|as|.
-It then returns the value \verb|b| to include in the output list, and whatever other
-list it wishes, upon which \verb|mapThrough| will recurse.
-
-We can write this combinator using our infrastructure for subsidiary recursion.
-For this to work, we need to supply the mapped function with the fold function
-for \verb|mapThrough|'s recursion.  This is so that the mapped function can
-initiate a subsidiary recursion, returning a value in the abstract type \verb|R|
-of \verb|mapThrough|'s recursion.  So the type we will use for mapped functions
-is:
+\caption{The \texttt{mapThrough} function in Haskell}
+\label{fig:mapthroughhs}
+\end{figure}
+
+The Haskell library \verb|Data.List.Extra| has a function
+\verb|repeatedly|, defined essentially as in
+Figure~\ref{fig:mapthroughhs}. We attempt a more informative
+name.  This is like the standard \verb|map| function on lists, except
+that the function \verb|f| that we are mapping (or ``mapping
+through'') takes in not just the current element \verb|a|, but also
+the tail \verb|as|.  It then returns the value \verb|b| to include in
+the output list, and whatever other list it wishes, upon which
+\verb|mapThrough| will recurse.
+
+To write this combinator using our infrastructure for subsidiary
+recursion, we need to supply the mapped function with the fold
+function for \verb|mapThrough|'s recursion.  This is so that the
+mapped function can initiate a subsidiary recursion, returning a value
+in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  So
+the type we will use for mapped functions is:
 \begin{verbatim}
@@
 must have type \verb|R|, and hence must be (hereditarily) a tail
-of the input of type \verb|R|.
-
-Given this definition, the code for \verb|mapThrough| and
-\verb|mapThroughr| is in Figure~\ref{fig:mapthrough}.  The code for
+of the input.
+
+Given this definition, the code is in Figure~\ref{fig:mapthrough}.  
 \verb|MapThroughAlg| is very similar (discounting syntax) to the
@@
 from the definition of \verb|mappedT|, we have that \verb|b : B| and
-\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> List B|
+\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> list B|
 on \verb|c|.  Note that as we are building up a new list from scratch
@@
 Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
-                      {B : Set}(f:mappedT A B) : R -> list B :=
-  fo (Const (list B)) (FunConst (list B)) (MapThroughAlg f R).
-
-Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B :=
-  mapThroughr (fold (ListF A)) f.
-\end{verbatim}
-\caption{The \texttt{mapThrough} and \texttt{mapThroughr} functions,
-  with their defining algebra}
+                      {B : Set}(f:mappedT A B) : R -> list B.
+Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B.
+\end{verbatim}
+\caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
+  code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
+  we omit it}
 \label{fig:mapthrough}
@@
 
-Using \verb|mapThrough|, we can quite concisely implement
-\emph{run-length encoding}, a basic data-compression algorithm where
-maximal sequences of $n$ occurrences of element $e$ are summarized by
-the pair $(n,e)$~\cite{datacomp}.  In Haskell, invoking \verb|span| and
-\verb|mapThrough| (defined above), the code is simply
+\begin{figure}
 \begin{verbatim}
@@
 \end{verbatim} 
-\noindent (Recall that \verb|(== a)| is a Haskell \emph{section}
-testing its input for equality with \verb|a|.)  The
+\caption{Run-length encoding in Haskell, using \texttt{mapThrough} and \texttt{span}}
+\label{fig:rlehs}
+\end{figure}
+
+Using \verb|mapThrough|, we can quite concisely implement
+\emph{run-length encoding}, a basic data-compression algorithm where
+maximal sequences of $n$ occurrences of element $e$ are summarized by
+the pair $(n,e)$~\cite{datacomp}.  Haskell code is in Figure~\ref{fig:rlehs}.
+Recall that \verb|(== a)| 
+tests its input for equality with \verb|a|.  The
 \verb|compressSpan| helper function gathers up all elements at the
@@
 The pair \verb|(1 + length p, a)| is returned to summarize
-\verb|a :: p|.  We then use \verb|mapThrough| to iterate
+\verb|a :: p|.  The \verb|mapThrough| combinator then iterates
 \verb|compressSpan| through the suffix \verb|s|.
@@
 Assuming \verb|A : Set| and an equality test \verb|eqb : A -> A -> bool| on it,
-code for run-length encoding using our infrastructure is listed in Figure~\ref{fig:rle}.
+we port this code to our Coq infrastructure in Figure~\ref{fig:rle}.
 The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
@@
 \verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).  Following the
-pattern we have seen in all the examples above, we may then define
-function \verb|mapThroughr| for subsidiary recursions, and
-\verb|mapThrough| for top-level recursions.
+pattern seen above, we define function \verb|rle| for top-level
+recursions using \verb|fold| (we could also define a subsidiary version
+\verb|rler|).
 
@@
   MapThroughAlg compressSpan C.
-
 Definition rle(xs : List A) : list (nat * A)
-  := @fold (ListF A) RleCarr (FunConst (list (nat * A))) (RleAlg (List A)) xs.
+  := fold (ListF A) RleCarr (FunConst (list (nat * A))) (RleAlg (List A)) xs.
 \end{verbatim}
@@
 require algebras, and also for the definition of \verb|Subrec| (which
-we will see in more detail in the next section).  As already noted,
-full positive-recursive types are incompatible with Coq's type
+we will see in more detail in the next section).  Full
+positive-recursive types are incompatible with Coq's type
 theory~\cite{coquand88}.  One can impose some restrictions on large
 eliminations which then enable positive-recursive
-types~\cite{blanqui05}.  This approach would require changing the
-underlying theory.  To avoid this, we here take a different approach,
-exploiting Coq's impredicative polymorphism.
+types~\cite{blanqui05}, but this requires changing the underlying
+theory.  Here we take a different approach, exploiting Coq's
+impredicative polymorphism.
 
@@
 
-For a simple example of noncanonicity, suppose we instantiate \verb|F|
-with \verb|ListF| (of Figure~\ref{fig:listf}).  Please note that as
-\verb|Mu| is used in our derivation of subsidiary recursion, we will
-not instantiate this \verb|F| with the signature functor of a datatype
-directly; but this will show the issue in a simple form.  Let us
-temporarily define \verb|List A| as \verb|Mu (ListF A)| (again, for
-subsidiary recursion we use a different functor than just \verb|ListF|
-directly).  The canonical way to define the empty list would be, 
-implicitly instantiating \verb|F| to \verb|ListF A|,
+For a simple example: suppose we instantiate \verb|F| with
+\verb|ListF| (of Figure~\ref{fig:listf}).  Our derivation uses a
+different type that wraps \verb|F|, but this will show the issue in a
+simple form.  Let us temporarily define \verb|List A| as
+\verb|Mu (ListF A)| (again, for subsidiary recursion do not use just
+\verb|ListF| directly).  The canonical way to define the empty list
+would be, implicitly instantiating \verb|F| to \verb|ListF A|,
 \begin{verbatim}
@@
 Noncanonicity must be handled carefully when reasoning about functions
-defined with our interface.  We will see examples in
+defined with our interface.  We will see an example in
 Section~\ref{sec:examplesi}.  First, though, let us complete the
@@
 for all algebras with anchor type \verb|Subrec| and functorial carrier
-\verb|X|, compute a value of type \verb|X Subrec|.  This is a generalization
-of the functor-generic type for the Church encoding:
-\[
-\forall\ X.\ \textit{Alg}\ X \to X
-\]
-\noindent where $\textit{Alg}\ X$ is $F\ X \to X$.  
-We elide the implementation of the \verb|roll| and \verb|unroll| functions, but
-note that \verb|unroll| makes use of functoriality of carriers \verb|X|.
+\verb|X|, compute a value of type \verb|X Subrec|.  This is a
+generalization of the functor-generic type
+$\forall\ X.\ \textit{Alg}\ X \to X$ for the Church encoding, where
+$\textit{Alg}\ X$ is $F\ X \to X$.  We elide the implementation of the
+\verb|roll| and \verb|unroll| functions, but note that \verb|unroll|
+makes use of functoriality of carriers \verb|X|.
 
@@
 \verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
-\verb|FunX|) over \verb|d : Subrec|, \verb|fold| \verb|unroll|s the
-definition of \verb|Subrec| and applies that to the algebra (with its
+\verb|FunX|) over \verb|d : Subrec|, we \verb|unroll| the
+definition of \verb|Subrec| and apply that to the algebra (with its
 carrier).
@@
 \verb|SubrecF Subrec| (we saw this type unfolded at the start of
-Section~\ref{sec:subrecimpl}).  This value takes in a carrier
+this section).  This value takes in a carrier
 \verb|X|, its fmap function \verb|xmap|, and an algebra \verb|alg|
@@
 \begin{itemize}
-\item \verb|Subrec| is passed as the value for the abstract type \verb|R|.
-\item the identity function is passed as the value for \verb|reveal : R -> Subrec| (since \verb|R| has
-  been instantiated to \verb|Subrec|).
+\item \verb|Subrec| is passed as the value for the abstract type \verb|R|; this is what enables all
+  the rest of the components to have the desired types, since we will pass values that have \verb|Subrec|
+  where the interface mentions \verb|R|.
+\item the identity function is passed as the value for \verb|reveal : R -> Subrec|.
 \item The function \verb|fold|, which expects an algebra with anchor type \verb|Subrec|, is passed as
@@
 \item For the \verb|eval : R -> X R| function, we pass \verb|(fold X xmap alg) : Subrec -> X Subrec|.
-\item For the subdata structure of type \verb|F R|, we pass \verb|d : F Subrec| (again, since we
-  are instantiating \verb|R| with \verb|Subrec|).
+\item For the subdata structure of type \verb|F R|, we pass \verb|d : F Subrec|. 
 \end{itemize}
@@
 \label{fig:subrecb}
-\end{figure}
-
-\subsection{Converting between \texttt{list} and \texttt{List} (\texttt{List.v})}
-
-Figure~\ref{fig:listb} gives the definitions (specified in
-Figure~\ref{fig:listf} above) for conversion functions between
-\texttt{list} (the list type of Coq's standard library) and our
-\texttt{List} type for subsidiary recursion on lists.  To convert a
-\verb|list| to a \verb|List|, we use \verb|listFold| to recursively
-apply the function \verb|inn| (Figure~\ref{fig:subrecb}) throughout a
-\verb|list|.  This \verb|listFold| provides a functorial view of the
-usual fold-right function on lists.  To convert a \verb|List| to a
-\verb|list|, we use our \verb|fold| function
-(Figure~\ref{fig:subrecb}) to fold an algebra over the input
-\verb|List|.  This algebra does an \verb|fmap| (for \verb|ListF|) with
-\verb|eval : R -> list A|.  The resulting value has type
-\verb|ListF (list A)|.  We may then apply \verb|listIn| to produce a
-\verb|list A|.  This \verb|listIn| function converts from
-\verb|ListF (list A)| to \verb|list A| in the obvious way, converting
-\verb|Nil| to the empty list, and \verb|Cons| to an application of
-\verb|::|.  These functions were not necessary in implementing
-examples of subsidiary recursion above, but they will be needed
-for reasoning about them.  We turn to this topic next.
-
-\begin{figure}
-\begin{verbatim}
-Fixpoint listFold(l : list A){X : Set}(alg : ListF X -> X) : X :=
-  match l with
-    nil => alg Nil
-  | cons hd tl => alg (Cons hd (listFold tl alg))
-  end.
-
-Definition toList (xs : list A) : List := listFold xs (inn ListF).
-
-Definition listIn(d : ListF (list A)) : list A :=
-  match d with
-    Nil => []
-  | Cons hd tl => hd :: tl
-  end.
-
-Definition fromList : List -> list A :=
-  fold ListF (Const (list A)) (FunConst (list A))
-       (rollAlg (fun R reveal fo eval fr => listIn (fmap eval fr))) .
-\end{verbatim}
-\caption{Additional functions for the \texttt{List} type (\texttt{Subrec ListF})}
-\label{fig:listb}
 \end{figure}
@@
 Definition FoldTi(alg : KAlgi)(C : kMo) : kMo :=
-  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), alg C X -> C d -> X C d.
+  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), 
+           alg C X -> C d -> X C d.
 
@@
 where the indexed anchor type is not used (so the carriers are
-constant indexed-functor returning the types shown).  But for the
-third, \verb|GuardPres|, we are showing that whenever \verb|spanh|
-(see Figure~\ref{fig:span}) returns a suffix \verb|r|, that suffix
-satisfies the indexed anchor type.  This enables us to invoke an
-outer induction hypothesis on this suffix, when using \verb|span|
-subsidiarily.
+constant indexed-functor returning the types shown).  But
+\verb|GuardPresF| uses the indexed anchor type (its argument \verb|S|),
+to express that whenever \verb|spanh| returns a suffix \verb|r|, that
+suffix satisfies the indexed anchor type.  This enables us to invoke
+an outer induction hypothesis on this suffix, when using \verb|span|
+subsidiarily.  Using these lemmas, we can write a short proof by
+subsidiary induction of the following, where
+\verb|rld : list (nat * A) -> list A| is the obvious decoding
+function:
+
+\begin{verbatim}
+Theorem RldRle (xs : list A): rld (rle (toList xs)) = xs.  
+\end{verbatim}
+
+\noindent We invoke the lemmas about \verb|span| subsidiarily,
+so that we may apply our induction hypothesis to the suffix
+that \verb|span| returns (on which \verb|mapThrough| then recurses).
+For example, the lemma for \verb|GuardPresF| takes in the
+indexed fold function \verb|foi| from the outer induction (for \verb|RldRle|),
+to show that the abstract predicate \verb|R| applies to the suffix
+\verb|r| returned by \verb|span|.  This enables the outer induction
+hypothesis (for \verb|RldRle|) to be applied.
+
+\begin{verbatim}
+Lemma guardPres{R : List A -> Prop}(foi:forall d : List A, ListFoldTi R d)
+      (p : A -> bool)(xs : List A)(rxs : R xs)
+      (l:list A)(r : List A)(e: span p xs = (l,r)) : R r.
+\end{verbatim}
 
@@
 \end{figure}
-
-Using these lemmas, we can write a short proof by subsidiary induction of:
-
-\begin{verbatim}
-Theorem RldRle (xs : list A): rld (rle (toList xs)) = xs.  
-\end{verbatim}
 
@@
 
-\subsection{Termination}
+\textbf{Termination.}
 In some tools, like Coq, Agda, and Lean, termination is checked
@@
 
-\subsection{Mendler encoding}
-
+\textbf{Mendler encoding.}
 Mendler introduced the basic idea of using universal abstraction to

### 2022-01-16 paper/paper.tex
complete draft
I will read over again, polish, may add more related work, then
circulate to the co-authors.

@@
 
-\subsection{Retractive-positive recursive types}
+\subsection{Retractive-positive recursive types (\texttt{Mu.v})}
 \label{sec:mu}
@@
 
-\textbf{Termination.}
-In some tools, like Coq, Agda, and Lean, termination is checked
-statically, based on structural decrease at recursive calls.  Others,
-like Isabelle/HOL, allow one to write recursions first, and prove
-(possibly with automated help) their termination
-afterwards~\cite{krauss}.
-
-It has not escaped the notice of designers of ITPs that structural
-recursion is not the only form of terminating recursion.  All the
-mentioned tools provide support for well-founded recursion, where for
-recursive calls, one must show that the parameter of recursion has
-decreased in some well-founded order.
+\textbf{Termination.}  In some tools, like Coq, Agda, and Lean,
+termination is checked statically, based on structural decrease.
+Others, like Isabelle/HOL, allow one to write recursions first, and
+prove (possibly with automated help) their termination
+afterwards~\cite{krauss}.  Well-founded recursion replaces structural
+decrease with decrease in a well-founded ordering. At least in
+constructive type theory, evidence of well-foundedness then propagates
+through code.  In contrast, our approach here, while less general,
+does not clutter code with proofs.  For more on partiality and recursion
+in theorem provers, see~\cite{bove+16}.
 
 Subsidiary recursion can be seen as a generalization of \emph{nested
-recursion}, which allows recursive calls of the form \verb|f (f x)|.
+recursion}, which allows recursive calls of the form \verb|f (f x)|~\cite{krauss10}.
 In subsidiary recursion, these are generalized to the form
@@
 
-\textbf{Mendler encoding.}
-Mendler introduced the basic idea of using universal abstraction to
-support compositional termination checking; an accessible source
-is~\cite{mendler91}.  He introduces a functor-generic recursor of type
-\[
-\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to X}
-\]
-\noindent We have adopted this idea to the constructor of the type
-\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
-categorical perspective on Mendler-style recursion~\cite{uustalu99}.
-Others have explored the possibility of using it with negative type
-schemes~\cite{ahn11}.
+\textbf{Mendler encoding.}  Mendler introduced the basic idea of using
+universal abstraction to support compositional termination checking;
+an accessible source is~\cite{mendler91}.  He introduces a
+functor-generic recursor of type $\all{X}{(\all{R}{(R \to X) \to F\ R
+    \to X}) \to \mu\ F\ \to X}$. We have adopted this idea to the
+constructor of the type \verb|Mu| (Section~\ref{sec:mu}).  Previous
+works explored the categorical perspective on Mendler-style
+recursion~\cite{uustalu99}, and its use with negative type
+schemes~\cite{ahn11}.  Previous work from our group showed
+how to derive inductive datatypes in Cedille using encodings
+extending the Mendler encoding~\cite{firsov+18b,firsov+18a,stump18}.
 
@@
 
+We have seen a derivation in Coq of a scheme for terminating
+subsidiary recursion, where recursions may be nested and outer
+recursive calls may be made on results of inner recursions.  We saw
+examples invoking the \verb|span| function as a subsidiary recursion,
+for functions \verb|wordsBy| and run-length encoding.  We also looked
+briefly at the extension of this interface to support subsidiary
+induction, with example lemmas about \verb|span|, and the decoding
+correctness theorem for run-length encoding.  There are many other
+interesting examples we can develop in Coq with this interface,
+including natural-number division, which may invoke subtraction as a
+subsidiary recursion.  Another example is Harper's regular-expression
+matcher, which previous work showed can be implemented in Cedille
+using a form of nested recursion that is subsumed by subsidiary
+recursion~\cite{stump20}.  We may also attempt to extend the recursion
+universe further, to allow other forms of recursion like
+divide-and-conquer, where some (necessarily limited) ability to
+recurse on values built using constructors is required.
+
+\pagebreak
 \bibliography{biblio}

### 2022-01-16 paper/paper.tex
revised first draft

@@
 $\mu$, but we will see how to work around this.  Our definition of
-retractive-positive recursive types makes essential use of impredicate
-quantification, and hence is not available in predicative theories
+retractive-positive recursive types makes essential use of impredicative
+quantification, and hence is not legal in predicative theories
 like Agda's.
@@
 algebra and category theory to describe inductive datatypes and their
-recursion principles.  On this approach, one describes transformations
+recursion principles (cf.~\cite{traytel12,cockett92,hagino87}).  On this approach, one describes transformations
 to be performed on data as \emph{algebras}, which can then be
 \textit{folded} over data.  The simplest form of algebras, namely
-$F$-algebras, are morphisms from $F\ A$ to $A$, for carrier object
+$F$-algebras for functor $F$, are morphisms from $F\ A$ to $A$, for carrier object
 $A$.  From a programming perspective, an $F$-algebra is given input of
-type $F\ A$, and must compute a result of type $A$.
-
-Algebras for our subsidiary recursion are more complex.  First, for
+type $F\ A$, and must compute a result of type $A$.  An example of $F$
+is the signature functor for lists, which we will use below:
+
+\begin{verbatim}
+Inductive ListF(X : Set) : Set :=
+| Nil : ListF X
+| Cons : A -> X -> ListF X.
+\end{verbatim}
+
+Algebras for our subsidiary recursion are more complex than $F$-algebras.  First, for
 reasons we will explain further below, the carrier of the algebra will
@@
   \begin{verbatim}
-Inductive ListF(X : Set) : Set :=
-| Nil : ListF X
-| Cons : A -> X -> ListF X.
-
 Definition List := Subrec ListF .
@@
   This file specializes the development in \verb|Subrec.v| to the case
-  of lists (parametrized by the type \verb|A| of elements).  In
+  of lists, parametrized by the type \verb|A| of elements.  In
   general, to use our development to get subsidiary recursion over
@@
   reasons, we just give the example of lists.  The file defines the
-  signature functor \verb|ListF|, shown in Figure~\ref{fig:listf}.  We
-  then define \verb|List| to be \verb|Subrec|, with the instantiation
-  of \verb|F| to \verb|ListF A|.  This type \verb|List| is not to be
+  signature functor \verb|ListF|, already shown above.  We have
+  also the definitions of Figure~\ref{fig:listf}.  \verb|List| is defined to be \verb|Subrec|, with
+   \verb|F| instantiated to \verb|ListF A|.  This type \verb|List| is not to be
   confused with the type \verb|list| of lists in Coq's standard
@@
   for \verb|List|, and types for conversion functions between
-  \verb|List| and \verb|list|; one direction uses Coq's structural
-  recursion, the other uses our subsidiary recursion (code elided).
+  \verb|List| and \verb|list| (code elided). One direction uses Coq's structural
+  recursion, the other uses subsidiary recursion, which we will
+  see next.
 
@@
 
-\verb|Subrec.v| also defines the notion of algebra that is used for
-writing recursions.  The central definitions are in
+\verb|Subrec.v| also implements the notion of algebra we introduced informally above.  The central definitions are in
 Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
@@
 in the subdata structure of type \verb|F R|.
+
+\subsubsection{The algebra for \texttt{span}}
+
+Figure~\ref{fig:spanalg} gives the algebra \verb|SpanAlg| for computing
+\verb|span|.  The type of \verb|SpanAlg p C| is
+\begin{verbatim}
+Alg (ListF A) C SpanF
+\end{verbatim}
+This states that we are defining an algebra (\verb|Alg|) for the
+\verb|ListF A| functor, with anchor type \verb|C| and carrier
+\verb|SpanF|.  \verb|SpanF| has type \verb|Set -> Set|, as required
+for the carriers of our algebras. The definition of \verb|SpanAlg| is
+actually parametrized by \verb|C|, which is good, as it means we can
+use \verb|SpanAlg| for top-level or subsidiary recursions.
 
@@
 \end{verbatim}
-\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function}
+\caption{The algebra \texttt{SpanAlg} for the \texttt{span} function (\texttt{Span.v})}
 \label{fig:spanalg}
@@
 
-Figure~\ref{fig:spanalg} gives the algebra \verb|SpanAlg| for computing
-\verb|span|.  The type of \verb|SpanAlg p C| is
-\begin{verbatim}
-Alg (ListF A) C SpanF
-\end{verbatim}
-This states that we are defining an algebra (\verb|Alg|) for the
-\verb|ListF A| functor, with anchor type \verb|C| and carrier
-\verb|SpanF|.  \verb|SpanF| has type \verb|Set -> Set|, as required
-for the carriers of our algebras. The definition of \verb|SpanAlg| is
-actually parametrized by \verb|C|, which is good, as it means we can
-use \verb|SpanAlg| for top-level or subsidiary recursions.
-
 Let us continue through the code for \verb|SpanAlg|
-(Figure~\ref{fig:span}).  We use \verb|rollAlg| to create an algebra
+(Figure~\ref{fig:spanalg}).  We use \verb|rollAlg| to create an algebra
 from something whose type is an application of \verb|AlgF|.  This
@@
 
+\subsubsection{Defining \texttt{span} from \texttt{SpanAlg}}
+
 \verb|SpanAlg| is used in the definition of \verb|spanhr|, in
@@
 from a value of type \verb|R|.  If \verb|spanhr| returns
-\verb|SpanSomeMatch|, then the return value gives us the nonempty
-prefix (\verb|l|) and the suffix (\verb|r|), which we then return.  We
-also define a version of \verb|break| for subsidiary recursion.
+\verb|SpanSomeMatch l r|, then we return the nonempty
+prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| for subsidiary recursion.
 
@@
 \end{verbatim}
-\caption{Functions derived from \texttt{SpanAlg}}
+\caption{Functions derived from \texttt{SpanAlg} (\texttt{Span.v})}
 \label{fig:span}
@@
 an algebra \verb|WordsBy|, parametrized by anchor type \verb|C| (and also
-the predicate \verb|p|), of type
-\begin{verbatim}
-Alg (ListF A) C (Const (list (list A)))
-\end{verbatim}
-\noindent This says that \verb|WordsBy p C| is an algebra (\verb|Alg|)
+the predicate \verb|p|), of type \verb|Alg (ListF A) C (Const (list (list A)))|.
+ This says that \verb|WordsBy p C| is an algebra (\verb|Alg|)
 for the \verb|ListF A| functor, with anchor type \verb|C|, and carrier
@@
 
-The code for \verb|WordsBy| is, except for the noise of \verb|rollAlg|
-and accepting the components of the recursion universe, essentially
+The code for \verb|WordsBy| is essentially
 the same as what we saw in Section~\ref{sec:intro}.  We pattern match
@@
 can indeed apply \verb|wordsBy : R -> list (list A)| to
-\verb|z| to recurse.
+\verb|z| to recurse.  The figure also shows the code for the subsidiary
+recursion \verb|wordsByr|.
 
@@
   fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p R) xs.
-
-Definition wordsBy(p : A -> bool)(xs : List A) : list (list A) :=
-  wordsByr (fold (ListF A)) p xs.
-\end{verbatim}
-\caption{The \texttt{wordsBy} and \texttt{wordsByr} function, defined using an algebra}
+\end{verbatim}
+\caption{Functions functions \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
+  fold (\texttt{WordsBy.v})}
 \label{fig:wordsby}
@@
 \verb|repeatedly|, defined essentially as in
-Figure~\ref{fig:mapthroughhs}. We attempt a more informative
+Figure~\ref{fig:mapthroughhs}, though we attempt a more informative
 name.  This is like the standard \verb|map| function on lists, except
@@
 Given this definition, the code is in Figure~\ref{fig:mapthrough}.  
-\verb|MapThroughAlg| is very similar (discounting syntax) to the
-Haskell code above.  Here, though, when we call \verb|f|, we must
+\verb|MapThroughAlg| is similar to the
+Haskell code above, though when we call \verb|f|, we must
 supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
@@
   code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
-  we omit it}
+  we omit it (\texttt{MapThrough.v})}
 \label{fig:mapthrough}
@@
 fold function \verb|fo| with anchor type \verb|R|, and then has the
-responsibility of mapping the tail at type \verb|R| (second input) to
+responsibility of extracting from the tail at type \verb|R| (second input) 
 a result upon which \verb|mapThrough| should recurse (second component
@@
 \caption{The function \texttt{rle} for run-length encoding, and the algebra \texttt{RleAlg} defining it
-in terms of \texttt{MapThroughAlg} (Figure~\ref{fig:mapthrough})}
+in terms of \texttt{MapThroughAlg} of Figure~\ref{fig:mapthrough} (\texttt{Rle.v})}
 \label{fig:rle}
@@
 types~\cite{blanqui05}, but this requires changing the underlying
-theory.  Here we take a different approach, exploiting Coq's
-impredicative polymorphism.
-
-This is done in a file \verb|Mu.v|, whose central definitions are in
-Figure~\ref{fig:mu}.  The development is parametrized by
-\verb|F : Set -> Set| which is assumed to have an \verb|fmap| function
+theory.  Here we exploit Coq's
+impredicative polymorphism for a different solution.
+
+Assume \verb|F : Set -> Set|, with an \verb|fmap| function
 (morphism part of the functor) of type
@@
 \end{verbatim}
-
-\begin{figure}
-\begin{verbatim}
-  Inductive Mu : Set := 
-    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
-
-  Definition inMu(d : F Mu) : Mu :=
-    mu Mu (fun x => x) d.
-
-  Definition outMu(m : Mu) : F Mu :=
-    match m with
-    | mu A r d => fmap r d
-    end.
-
-  Lemma outIn(d : F Mu) : outMu (inMu d) = d.
-\end{verbatim}
-\caption{Derivation of retractive-positive recursive types}
-\label{fig:mu}
-\end{figure}
-
-Let us consider the code in Figure~\ref{fig:mu}.  The critical
+\noindent Then we make the definitions of Figure~\ref{fig:mu}.  The critical
 idea is embodied in the definition of \verb|Mu|.  Ideally, we would like
@@
 
-Instead, the definition of \verb|Mu| in Figure~\ref{fig:mu} weakens
-this ideal definition to a strictly positive approximation:
+\begin{figure}
+\begin{verbatim}
+  Inductive Mu : Set := 
+    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
+
+  Definition inMu(d : F Mu) : Mu :=
+    mu Mu (fun x => x) d.
+
+  Definition outMu(m : Mu) : F Mu :=
+    match m with
+    | mu A r d => fmap r d
+    end.
+
+  Lemma outIn(d : F Mu) : outMu (inMu d) = d.
+\end{verbatim}
+\caption{Derivation of retractive-positive recursive types (\texttt{Mu.v})}
+\label{fig:mu}
+\end{figure}
+
+The definition of \verb|Mu| weakens
+this to a strictly positive approximation:
 \begin{verbatim}
@@
 \end{verbatim}
-\caption{Definition of \texttt{Subrec} as a fixed-point of \texttt{SubrecF}}
+\caption{Definition of \texttt{Subrec} as a fixed-point of \texttt{SubrecF} (\texttt{Subrec.v})}
 \label{fig:subrec}
@@
 \end{verbatim}
-\caption{The rest of the interface for \texttt{Subrec}}
+\caption{The rest of the interface for \texttt{Subrec} (\texttt{Subrec.v})}
 \label{fig:subrecb}
@@
 \verb|Subrec|~\cite{mcbride00}.  \verb|KAlgi| is the kind for indexed
-algebras.  \verb|FoldTi| is the indexed version of \verb|FoldT|: it is
+algebras.  \verb|FoldTi| is the indexed version of \verb|FoldT|: it 
 expresses provability of \verb|X C| for \verb|d|, based on an indexed
-algebra and a value of type \verb|C d|, where \verb|C| is the anchor
-type, now indexed.  \verb|AlgFi| and \verb|Algi| are indexed versions
+algebra and a value of type \verb|C d|, where \verb|C| is the (indexed) anchor
+type.  \verb|AlgFi| and \verb|Algi| are indexed versions
 of the algebras we saw for recursion.  The \verb|eval| function
@@
 \end{verbatim}
-\caption{Interface for subsidiary induction}
+\caption{Interface for subsidiary induction (\texttt{Subreci.v})}
 \label{fig:subreci}
@@
 \end{verbatim}
-\caption{The indexed version \texttt{ListFi} of \texttt{ListF}}
+\caption{The indexed version \texttt{ListFi} of \texttt{ListF} (\texttt{List.v})}
 \label{fig:listfi}
@@
 where the indexed anchor type is not used (so the carriers are
-constant indexed-functor returning the types shown).  But
+constant indexed functors returning the types shown).  But
 \verb|GuardPresF| uses the indexed anchor type (its argument \verb|S|),
@@
 suffix satisfies the indexed anchor type.  This enables us to invoke
-an outer induction hypothesis on this suffix, when using \verb|span|
-subsidiarily.  Using these lemmas, we can write a short proof by
+an outer induction hypothesis on this suffix, when reasoning
+subsidiarily about \verb|span|.  Using these lemmas, we can write a short proof by
 subsidiary induction of the following, where
@@
 \end{verbatim}
-\caption{Statements of three lemmas about \texttt{span}}
+\caption{Statements of three lemmas about \texttt{span} (directory \texttt{SpanPfs})}
 \label{fig:spanlem}
@@
 in this case is to use a function \verb|getNil| (\verb|List.v|) that computes
-an empty list from \verb|xs|.  So the statement that one can prove is:
-
+an empty list from \verb|xs|.  The statement that one can prove is shown in Figure~\ref{fig:spanforallt}.
+
+\begin{figure}
 \begin{verbatim}
@@
 \end{verbatim}
+\caption{A statement of the property that \texttt{span} returns the empty suffix, computed using \texttt{getNil} to avoid noncanonicity problems, if all elements satisfy \texttt{p}}
+\label{fig:spanforallt}
+\end{figure}
 
@@
 prove (possibly with automated help) their termination
-afterwards~\cite{krauss}.  Well-founded recursion replaces structural
-decrease with decrease in a well-founded ordering. At least in
-constructive type theory, evidence of well-foundedness then propagates
-through code.  In contrast, our approach here, while less general,
-does not clutter code with proofs.  For more on partiality and recursion
-in theorem provers, see~\cite{bove+16}.
-
+afterwards~\cite{krauss}.  These tools all support well-founded
+recursion, but in constructive type theory, evidence of
+well-foundedness then propagates through code.  In contrast, our
+approach here, while less general, does not clutter code with proofs.
 Subsidiary recursion can be seen as a generalization of \emph{nested
-recursion}, which allows recursive calls of the form \verb|f (f x)|~\cite{krauss10}.
-In subsidiary recursion, these are generalized to the form
-\verb|f (g x)|, where \verb|g| could be \verb|f| or another
-recursively defined function.
-
-\textbf{Mendler encoding.}  Mendler introduced the basic idea of using
-universal abstraction to support compositional termination checking;
-an accessible source is~\cite{mendler91}.  He introduces a
+recursion}, which allows recursive calls of the form
+\verb|f (f x)|~\cite{krauss10}.  In subsidiary recursion, these are
+generalized to the form \verb|f (g x)|, where \verb|g| could be
+\verb|f| or another recursively defined function.  For more on
+partiality and recursion in theorem provers, see~\cite{bove+16}.
+
+Our work contributes to the program proposed by Owens and Slind, of
+broadening the scope of functional programs that can be accommodated
+in ITPs~\cite{owens+08}.  The goal of terminating recursion has been
+advocated in the literature on programming languages under the name
+\emph{strong functional programming}~\cite{turner95}.  Uustalu and
+Vene developed a categorical view of a recursion scheme allowing one
+level of subsidiary recursion, and illustrated it with an artificial
+example~\cite{uustalu11}.  In contrast, our scheme allows arbitrary
+finite nestings of recursion, and we illustrate it with realistic
+examples drawn from Haskell's \verb|Data.List.Extra|.
+
+\textbf{Mendler-style recursion.}  Mendler introduced the idea of using
+universal abstraction to support compositional termination checking~\cite{mendler91}.  He proposed a
 functor-generic recursor of type $\all{X}{(\all{R}{(R \to X) \to F\ R
@@
 how to derive inductive datatypes in Cedille using encodings
-extending the Mendler encoding~\cite{firsov+18b,firsov+18a,stump18}.
+extending the Mendler encoding~\cite{firsov+18b,firsov+18a}.
+Here, we do not derive inductive types using these methods,
+but rather apply them to justify a terminating recursion scheme.
 

### 2022-01-22 paper/paper.tex
Tiny, Tiny code fix in paper.
revising paper based on Chris's feedback (more to come)

@@
 This paper presents a functor-generic derivation of terminating
-subsidiary recursion and induction in Coq.  We should emphasize that
-this is a derivation within the type theory of Coq, and requires no
-axioms or other modifications to Coq. Using this derivation, we
-present several example functions like \verb|wordsBy|, and prove
-theorems about them.  A nice example is a definition of run-length
-encoding using \verb|span| as a subsidiary recursion, where we prove
-that encoding and then decoding returns the original list.  Our
-approach applies to the standard datatypes in the Coq library, and
-does not require switching libraries or datatype definitions.
+subsidiary recursion and induction in Coq.  We emphasize that this is
+a derivation within the type theory of Coq, and requires no axioms or
+other modifications to Coq, except the \verb|-impredicative-set|
+flag. Using this derivation, we present several example functions like
+\verb|wordsBy|, and prove theorems about them.  A nice example is a
+definition of run-length encoding using \verb|span| as a subsidiary
+recursion, where we prove that encoding and then decoding returns the
+original list.  Our approach applies to the standard datatypes in the
+Coq library, and does not require switching libraries or datatype
+definitions.
 
@@
 positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
-datatypes $D$ to be strictly positive: in the type for any constructor
-of $D$, $D$ cannot occur to the left of any arrows.  Our derivation
-needs to use positive-recursive types, where $D$ may occur to the left
-of an even number (only) of arrows.  We present a way to derive a
-weakened form of positive-recursive type that is sufficient for our
-examples (Section~\ref{sec:mu}).  The weakening is to require only
-that $F\ \mu$ is a retract of $\mu$, where $\mu$ is the recursive type
-and $F\ \mu$ its one-step unfolding.  Usually these types are
+datatypes $D$ to be strictly positive: in the input types of
+constructors of $D$, $D$ cannot occur to the left of any arrows.  Our
+derivation needs to use positive-recursive types, where $D$ may occur
+to the left of an even number (only) of arrows.  We present a way to
+derive a weakened form of positive-recursive type that is sufficient
+for our examples (Section~\ref{sec:mu}).  The weakening is to require
+only that $F\ (\mu F)$ is a retract of $\mu F$.  Usually these types are
 isomorphic.  Hence, we dub these \textbf{retractive-positive}
@@
 $\mu$, but we will see how to work around this.  Our definition of
-retractive-positive recursive types makes essential use of impredicative
-quantification, and hence is not legal in predicative theories
-like Agda's.
+retractive-positive recursive types makes essential use of
+impredicative quantification, and hence is not legal in predicative
+theories like Agda's.
 
@@
 
-All presented derivations have been checked with Coq version 8.13.2,
-using command-line option \verb|-impredicative-set|.  The code may be
+All presented derivations have been checked with Coq version 8.13.2.  The code may be
 found as release \verb|itp-2022| (dated prior to the ITP 2022
@@
 top-level recursion, our development's version of the actual datatype
-(e.g., \verb|List|).  The algebra is presented with:
+(e.g., \verb|List A|, for some \verb|A|).  The algebra is presented with:
 
@@
 universe}~\cite{stump20}.  As in other domains using the term
-``universe'', we have an entity (here, \verb|R|) from which one cannot
-escape by using the available operations (for other cases:
-the ordinal $\epsilon_0$ and $\omega^-$, the physical universe and traveling at
-the speed of light).  Staying in the recursion universe is good,
-because we may recurse (via \verb|eval|) on any value of type
-\verb|R|.  Some points must still be explained: why \verb|X| has type \verb|Set -> Set|,
-and the definition of \verb|FoldT|.  Let us see these details next.
+``universe'', we have a kind of space (here, \verb|R|) with operations
+that keep one in that space (for other cases: the ordinal $\epsilon_0$
+and $\omega^-$, the physical universe and traveling at the speed of
+light).  Staying in the recursion universe is good, because we may
+recurse (via \verb|eval|) on any value of type \verb|R|.  One can
+use \verb|reveal| to leave, but then \verb|eval| can no longer be used. Some points
+must still be explained: why \verb|X| has type \verb|Set -> Set|, and
+the definition of \verb|FoldT|.  Let us see these details next.
 
@@
 \verb|FoldT Alg C| is the type for fold functions which apply
-algebras of type \verb|Alg| to data of type \verb|C|, which we have
-already dubbed the \emph{anchor type} of the recursion.  At the top
-level of code, the anchor type would just be \verb|List| (for
-example).  When one initiates a subsidiary recursion, though, the
-anchor type will instead by the abstract type \verb|R| for the outer recursion.
-The variable \verb|Alg| occurs only positively (but not strictly
+algebras (\verb|Alg|) to data of type \verb|C|, which we have already
+dubbed the \emph{anchor type} of the recursion.  At the top level of
+code, the anchor type would just be \verb|List| (for example).  When
+one initiates a subsidiary recursion, though, the anchor type will
+instead by the abstract type \verb|R| for the outer recursion.  The
+variable \verb|Alg| occurs only positively (but not strictly
 positively) in \verb|AlgF|, because it occurs negatively in
@@
 recurse.  This is the first argument to a value of type
-\verb|AlgF Alg C X|.  An algebra can assume nothing about \verb|R|
-except that it supports the following operations.  First there is
-\verb|reveal|, which turns an \verb|R| into a \verb|C|.  This reveals
-that the data of type \verb|R| are really values of the anchor type
-of this recursion.  Next we have \texttt{fold}, which will allow us to fold
-another algebra over data of type \verb|R|.  We will use \verb|fold|
-to initiate subsidiary recursions.  Then there is \verb|eval|, for
-recursive calls on data of type \verb|R|.
+\verb|AlgF Alg C X|.  Reasoning parametrically, an algebra can assume
+nothing about \verb|R| except that it supports the following
+operations.  First there is \verb|reveal|, which turns an \verb|R|
+into a \verb|C|.  This reveals that the data of type \verb|R| are
+really values of the anchor type of this recursion.  Next we have
+\texttt{fold}, which will allow us to fold another algebra over data
+of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
+recursions.  Then there is \verb|eval|, for recursive calls on data of
+type \verb|R|.
 
@@
 Finally, Figure~\ref{fig:subrecb} defines \verb|out| as a subsidiary
-recursion, given any fold function with its anchor type \verb|R|.  The
-code for \verb|out| just folds an algebra over the input of type
-\verb|R|, where that algebra simply returns the subdata structure it
-is given.  Outside the recusion, this has type \verb|F R|; inside the
+recursion, given any fold function with its anchor type \verb|R|.
+Outside the recursion, \verb|d| has type \verb|F R|; inside the
 recursion it has type \verb|F R'| where \verb|R'| is the abstract type
-of the subsidiary recursion (named in the figure just for discussion
-here).  So \verb|out| implements the idea that unfolding an abstract
-type one step is just a trivial case of subsidiary recursion.
+of the subsidiary recursion.  So \verb|out| implements the idea that
+unfolding an abstract type one step is just a trivial case of
+subsidiary recursion.
 
@@
 Vene developed a categorical view of a recursion scheme allowing one
-level of subsidiary recursion, and illustrated it with an artificial
-example~\cite{uustalu11}.  In contrast, our scheme allows arbitrary
-finite nestings of recursion, and we illustrate it with realistic
-examples drawn from Haskell's \verb|Data.List.Extra|.
+level of subsidiary recursion, and illustrated it in Haskell with an
+artificial example~\cite{uustalu11}.  In contrast, our scheme allows
+arbitrary finite nestings of recursion, and we illustrate it in Coq
+with realistic examples.  Like them, we find that subsidiary recursion
+subsumes course-of-value recursion.
 
@@
 Here, we do not derive inductive types using these methods,
-but rather apply them to justify a terminating recursion scheme.
+but rather apply them to justify a terminating recursion scheme
+for existing datatypes.
 

### 2022-01-22 paper/paper.tex
completed revisions based on Chris's comments
Paper is too long now, but I will incorporate Ben's comments and then
shrink.

@@
   \begin{verbatim}
-Definition List := Subrec ListF .
+Definition List := Subrec ListF.
 Definition inList : ListF List -> List := inn ListF.
@@
 
-The final definitions in the figure are for \verb|fold|, which allows
+The last definitions in the figure are for \verb|fold|, which allows
 us to fold an \verb|Alg| over a value of type \verb|Subrec|; and for mapping
@@
 return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.
+
+Finally, for a recursion scheme, one would like to see not just the
+typed interface, but also the computation law.  This is stated as a
+theorem in Figure~\ref{fig:comp}.  Intuitively, it states that
+\verb|fold|ing an algebra over constructed data \verb|inn d| is equal
+to invoking the algebra on the identity function for \verb|reveal|
+from the interface for algebras (Figure~\ref{fig:algf}); \verb|fold|
+for the fold function; an invocation of \verb|fold| with the algebra
+for the \verb|eval| function; and \verb|d| for the subdata structure.
+
+\begin{figure}
+\begin{verbatim}
+Theorem FoldChar :
+ forall (X : Set -> Set) (FunX : Functor X) (IdF : FmapId X FunX)
+        (algf : AlgF Alg Subrec X) (d : F Subrec),
+ fold X FunX (rollAlg algf) (inn d) =
+   algf _ (fun x => x) fold (fold X FunX (rollAlg algf)) d .
+\end{verbatim}
+\end{figure}
 
@@
 will see it only in the form of a subdata structure of type
-\verb|F R|.  We will not be able to return it from our recursion at
+\verb|ListF A R|.  We will not be able to return it from our recursion at
 type \verb|R|, and hence we would not be able to return \verb|([],xs)|
@@
 \verb|R|. The prefix will be nonempty, and hence the suffix will be at
-most the tail of \verb|xs|.  This tail is available to the algebra
-in the subdata structure of type \verb|F R|.
+most the tail of \verb|xs|.  This suffix is available to the algebra
+in the subdata structure of type \verb|ListF A R|.
 
@@
 \verb|SpanSomeMatch l r|, then we return the nonempty
-prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| for subsidiary recursion.
+prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| for subsidiary recursion (e.g., in \verb|wordsBy|, below).
 
@@
 with realistic examples.  Like them, we find that subsidiary recursion
-subsumes course-of-value recursion.
-
-\textbf{Mendler-style recursion.}  Mendler introduced the idea of using
-universal abstraction to support compositional termination checking~\cite{mendler91}.  He proposed a
-functor-generic recursor of type $\all{X}{(\all{R}{(R \to X) \to F\ R
-    \to X}) \to \mu\ F\ \to X}$. We have adopted this idea to the
-constructor of the type \verb|Mu| (Section~\ref{sec:mu}).  Previous
-works explored the categorical perspective on Mendler-style
-recursion~\cite{uustalu99}, and its use with negative type
-schemes~\cite{ahn11}.  Previous work from our group showed
-how to derive inductive datatypes in Cedille using encodings
-extending the Mendler encoding~\cite{firsov+18b,firsov+18a}.
-Here, we do not derive inductive types using these methods,
-but rather apply them to justify a terminating recursion scheme
-for existing datatypes.
+subsumes course-of-value recursion.  The technique of sized types is
+similar to our method in providing a type-based method for
+termination~\cite{barthe04}.  This technique enriches datatypes with
+abstract sizes, which must then be propagated through code, using
+dependent types.  Our \verb|Subrec| does not require dependent types,
+resulting in just polymorphically typed code for the examples we saw
+(\verb|Subreci|, for proofs, of course does use dependent types).
+
+\textbf{Mendler-style recursion.}  Mendler introduced the idea of
+using universal abstraction to support compositional termination
+checking~\cite{mendler91}.  He proposed a functor-generic recursor of
+type $\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to
+  X}$. We have adopted this idea to the constructor of the type
+\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
+categorical perspective on Mendler-style recursion, and showed how to
+reduce it to basic catamorphisms (i.e., structural
+recursion)~\cite{uustalu99}.  Another considered its use with negative
+type schemes~\cite{ahn11}.  Previous work from our group showed how to
+derive inductive datatypes in Cedille using extensions of the Mendler
+encoding~\cite{firsov+18b,firsov+18a}.  Here, we do not derive
+inductive types using these methods, but rather apply them to justify
+a terminating recursion scheme for existing datatypes.
 

### 2022-01-24 paper/paper.tex

@@
 %\graphicspath{{./graphics/}}%helpful if your graphic files are in another directory
+\usepackage{minted}
 
@@
 a given predicate \verb|p|, and the remaining suffix:
-\begin{verbatim}
+\begin{minted}{haskell}
 span :: (a -> Bool) -> [a] -> ([a],[a])
@@
                 else ([],x:xs)
-\end{verbatim}
+\end{minted}
 \noindent The sole recursive call is \verb|span p xs|, and it occurs
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{haskell}
 wordsBy :: (a -> Bool) -> [a] -> [[a]]
@@
         (hd:w) : wordsBy p z
-\end{verbatim}
+\end{minted}
 \caption{Haskell code for \texttt{wordsBy}, demonstrating subsidiary recursion}
@@
 
-\begin{verbatim}
+\begin{minted}{coq}
 Inductive ListF(X : Set) : Set :=
@@
 | Cons : A -> X -> ListF X.
-\end{verbatim}
+\end{minted}
 
@@
 We will present the type \verb|FoldT Alg R| below.
-\item a function \verb|eval : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
+\item a function \verb|rec : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
 \item and a \emph{subdata structure} \verb|d : F R|, where \verb|F| is the signature functor for the datatype.
@@
 light).  Staying in the recursion universe is good, because we may
-recurse (via \verb|eval|) on any value of type \verb|R|.  One can
-use \verb|reveal| to leave, but then \verb|eval| can no longer be used. Some points
+recurse (via \verb|rec|) on any value of type \verb|R|.  One can
+use \verb|reveal| to leave, but then \verb|rec| can no longer be used. Some points
 must still be explained: why \verb|X| has type \verb|Set -> Set|, and
@@
 
-\subsection{The interface in more detail}
-
-Let us consider two central files from our development.
-
-\begin{figure}
-  \begin{verbatim}
+\subsection{Types for subsidiary recursion (\texttt{Subrec.v}, \texttt{List.v})}
+
+\begin{figure}
+\begin{minted}{coq}
 Definition List := Subrec ListF.
@@
 Definition fromList : List -> list A.
-\end{verbatim}
+\end{minted}
   \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
@@
 
-\subsubsection{Subrec.v}
-
-  This file is parametrized by a signature functor \verb|F| of type
-  \verb|Set -> Set|.  It provides the implementation of subsidiary
-  recursion.  Two crucial values are \verb|Subrec : Set|, which is the
-  type to use for subsidiary recursion; and
-  \verb|inn : F Subrec -> Subrec|, which is to be used as a
-  constructor for that type.  An important point, however, is that
-  \verb|Subrec.v| does not provide an induction principle based on
-  \verb|inn|.  Induction is derived later
-  (Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use
-  of retractive-positive recursive types, to take a fixed-point of a
-  construction based on \verb|F|.  We present these recursive
-  types in Section~\ref{sec:mu} below.
-
-  \subsubsection{List.v}
-
-  This file specializes the development in \verb|Subrec.v| to the case
-  of lists, parametrized by the type \verb|A| of elements.  In
-  general, to use our development to get subsidiary recursion over
-  some datatype, one will have a similar ``shim'' file.  For space
-  reasons, we just give the example of lists.  The file defines the
-  signature functor \verb|ListF|, already shown above.  We have
-  also the definitions of Figure~\ref{fig:listf}.  \verb|List| is defined to be \verb|Subrec|, with
-   \verb|F| instantiated to \verb|ListF A|.  This type \verb|List| is not to be
-  confused with the type \verb|list| of lists in Coq's standard
-  library.  As noted previously, our development is meant to be used
-  in extension of existing inductive datatypes, not replacing them.
-  The figure also shows constructors \verb|mkNil| and \verb|mkCons|
-  for \verb|List|, and types for conversion functions between
-  \verb|List| and \verb|list| (code elided). One direction uses Coq's structural
-  recursion, the other uses subsidiary recursion, which we will
-  see next.
+Our development is parametrized by a signature functor \verb|F| of
+type \verb|Set -> Set|.  It implements \verb|Subrec : Set|, which is
+the type to use for subsidiary recursion.  This type comes with
+\verb|inn : F Subrec -> Subrec|, which behaves computationally like a
+constructor.  \verb|Subrec.v| does not itself provide an induction
+principle based on \verb|inn|, however.  Induction is derived later
+(Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use of
+retractive-positive recursive types, to take a fixed-point of a
+construction based on \verb|F|.  We present these recursive types in
+Section~\ref{sec:mu} below.
+
+For our examples, we will consider the specialization to the case of
+lists, parametrized by the type \verb|A| of elements.  In general, to
+use our development to get subsidiary recursion over some datatype,
+one must define a signature functor for the datatype.  For lists, this
+is \verb|ListF| of Figure~\ref{fig:listf}.  \verb|List| is defined to
+be \verb|Subrec|, with \verb|F| instantiated to \verb|ListF A|.  This
+type \verb|List| is not to be confused with the type \verb|list| of
+lists in Coq's standard library.  As noted previously, our development
+is meant to be used in extension of existing inductive datatypes, not
+replacing them.  The figure also shows constructors \verb|mkNil| and
+\verb|mkCons| for \verb|List|, and types for conversion functions
+between \verb|List| and \verb|list| (code elided). One direction uses
+Coq's structural recursion, the other uses subsidiary recursion, which
+we will see next.
 
@@
 
-\verb|Subrec.v| also implements the notion of algebra we introduced informally above.  The central definitions are in
+\verb|Subrec.v| also implements the notion of algebra we introduced
+informally above.  The central definitions are in
 Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
@@
 implements our retractive-positive recursive types
-(Section~\ref{sec:mu}) at kind \verb|KAlg|.  Using \verb|Alg| will
-require that \verb|AlgF| only uses its parameter \verb|Alg|
-positively.  We will confirm this shortly.
-
-\begin{figure}
-\begin{verbatim}
+(Section~\ref{sec:mu}) at kind \verb|KAlg|.
+
+We need a fixed-point here due to the input \verb|fold| (\verb|Algf|)
+of type \verb|FoldT Alg C|.  This is the type for fold functions which
+apply algebras (\verb|Alg|) to data of type \verb|C|, which we have
+already dubbed the \emph{anchor type} of the recursion.  At the top
+level of code, the anchor type would just be \verb|List A| (for
+example).  When one initiates a subsidiary recursion, though, the
+anchor type will instead by the abstract type \verb|R| for the outer
+recursion.  The variable \verb|Alg| occurs only positively (but not
+strictly positively) in \verb|AlgF|, because it occurs negatively in
+\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
+we can indeed take a fixed-point of \verb|AlgF| to define the constant
+\verb|Alg|.
+
+\begin{figure}
+\begin{minted}{coq}
 Definition KAlg  : Type := Set -> (Set -> Set) -> Set.
@@
          (fold : FoldT Alg R)
-         (eval : R -> X R)      
+         (rec : R -> X R)      
          (d : F R),             
@@
   forall {C : Set} {X : Set -> Set}, Alg C X -> AlgF Alg C X.
-\end{verbatim}
-\caption{The type for algebras (\texttt{Subrec.v})}
+\end{minted}
+\caption{The type for algebras, parametrized over \texttt{F : Set -> Set} (\texttt{Subrec.v})}
 \label{fig:algf}
 \end{figure}
-
-\verb|FoldT Alg C| is the type for fold functions which apply
-algebras (\verb|Alg|) to data of type \verb|C|, which we have already
-dubbed the \emph{anchor type} of the recursion.  At the top level of
-code, the anchor type would just be \verb|List| (for example).  When
-one initiates a subsidiary recursion, though, the anchor type will
-instead by the abstract type \verb|R| for the outer recursion.  The
-variable \verb|Alg| occurs only positively (but not strictly
-positively) in \verb|AlgF|, because it occurs negatively in
-\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
-we can indeed take a fixed-point of \verb|AlgF| to define the constant
-\verb|Alg|.
 
@@
 of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
-recursions.  Then there is \verb|eval|, for recursive calls on data of
+recursions.  Then there is \verb|rec|, for recursive calls on data of
 type \verb|R|.
@@
 may mention \verb|R|, but (ii) outside that recursion, the result will
-mention the anchor type \verb|C|.  The \verb|eval| function returns
+mention the anchor type \verb|C|.  The \verb|rec| function returns
 something of type \verb|X R|, and so does the algebra itself; this
@@
 for the fold function; an invocation of \verb|fold| with the algebra
-for the \verb|eval| function; and \verb|d| for the subdata structure.
-
-\begin{figure}
-\begin{verbatim}
+for the \verb|rec| function; and \verb|d| for the subdata structure.
+
+\begin{figure}
+\begin{minted}{coq}
 Theorem FoldChar :
@@
    algf _ (fun x => x) fold (fold X FunX (rollAlg algf)) d .
-\end{verbatim}
+\end{minted}
 \end{figure}
@@
 
+This first example does not invoke subsidiary recursions, but will
+itself be used as a subsidiary recursion in other examples to follow.
 Given a predicate \verb|p : A -> bool|, and a value of type
@@
 generally, though, given an anchor type \verb|R : Set| along with a
-fold function for that anchor type (i.e., of type \verb|FoldT (Alg (ListF A)) R|),
-we would like to map an input list of type \verb|R| to a pair of
-type \verb|list A * R|.  The first component of this pair is
-going to be built up from scratch, and so cannot have type \verb|R|;
-we cannot statically ensure that outer recursions on it are legal.
-But the second component will be a subdatum of the input list,
-and so can still have type \verb|R|, enabling outer recursive calls.  So we want: 
-\begin{verbatim}
+fold function for that anchor type (i.e., of type
+\verb|FoldT (Alg (ListF A)) R|), we would like to map an input list of
+type \verb|R| to a pair of type \verb|list A * R|.  The first
+component of this pair is going to be built up from scratch, and so
+cannot have type \verb|R|; we cannot statically ensure that outer
+recursions on it are legal.  But the second component will be a
+subdatum of the input list, and so can still have type \verb|R|,
+enabling outer recursive calls.  So we want:
+\begin{minted}{coq}
 Definition spanr{R : Set}(fo:FoldT (Alg (ListF A)) R)
                 (p : A -> bool)(xs : R) : list A * R.
-\end{verbatim}
+\end{minted}
 \noindent From this we can also define the top-level recursion, by
@@
 over a list (Figure~\ref{fig:algf}), for the argument \verb|fo| of \verb|spanr|:
-\begin{verbatim}
+\begin{minted}{coq}
 Definition span(p : A -> bool)(xs : List A) : list A * List A
   := spanr (fold (ListF A)) p xs.
-\end{verbatim}
+\end{minted}
 
@@
 of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
-\begin{verbatim}
+\begin{minted}{coq}
 Inductive SpanF(X : Set) : Set :=
@@
 | SpanSomeMatch : list A -> X -> SpanF X.
-\end{verbatim}
+\end{minted}
 \noindent The idea is that the recursion will signal if it is in the
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition SpanAlg(p : A -> bool)(C : Set)
@@
        end).
-\end{verbatim}
+\end{minted}
 \caption{The algebra \texttt{SpanAlg} for the \texttt{span} function (\texttt{Span.v})}
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition spanhr{R : Set}(fo:FoldT (Alg (ListF A)) R)
@@
   spanr fo (fun x => negb (p x)) xs.
-\end{verbatim}
+\end{minted}
 \caption{Functions derived from \texttt{SpanAlg} (\texttt{Span.v})}
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
   Definition WordsBy(p : A -> bool)(C : Set)
@@
   fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p R) xs.
-\end{verbatim}
-\caption{Functions functions \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
+\end{minted}
+\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
   fold (\texttt{WordsBy.v})}
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{haskell}
 mapThrough :: (a -> [a] -> (b, [a])) -> [a] -> [b]
@@
     where (b, as') = f a as
-\end{verbatim}
+\end{minted}
 \caption{The \texttt{mapThrough} function in Haskell}
@@
 
-The Haskell library \verb|Data.List.Extra| has a function
-\verb|repeatedly|, defined essentially as in
+In this example, we see how to write a combinator that factors out a
+subsidiary recursion.  The Haskell library \verb|Data.List.Extra| has
+a function \verb|repeatedly|, defined essentially as in
 Figure~\ref{fig:mapthroughhs}, though we attempt a more informative
@@
 the type we will use for mapped functions is:
-\begin{verbatim}
+\begin{minted}{coq}
 Definition mappedT(A B : Set) : Set :=
   forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> B * R.
-\end{verbatim}
+\end{minted}
 \noindent This type is more informative than the Haskell type,
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition MapThroughAlg{B : Set}(f:mappedT A B)
@@
 Definition mapThrough{B : Set}(f:mappedT A B) : List A -> list B.
-\end{verbatim}
+\end{minted}
 \caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{haskell}
 rle :: Eq a => [a] -> [(Int,a)]
@@
             ((1 + length p, a),s)
-\end{verbatim} 
+\end{minted} 
 \caption{Run-length encoding in Haskell, using \texttt{mapThrough} and \texttt{span}}
@@
 
-Using \verb|mapThrough|, we can quite concisely implement
-\emph{run-length encoding}, a basic data-compression algorithm where
-maximal sequences of $n$ occurrences of element $e$ are summarized by
-the pair $(n,e)$~\cite{datacomp}.  Haskell code is in Figure~\ref{fig:rlehs}.
-Recall that \verb|(== a)| 
-tests its input for equality with \verb|a|.  The
-\verb|compressSpan| helper function gathers up all elements at the
+Finally, we have an example using our \verb|mapThrough| combinator
+together with a subsidiary recursion, to implement \emph{run-length
+encoding}.  This is a basic data-compression algorithm where maximal
+sequences of $n$ occurrences of element $e$ are summarized by the pair
+$(n,e)$~\cite{datacomp}.  Haskell code is in Figure~\ref{fig:rlehs}.
+Recall that \verb|(== a)| tests its input for equality with \verb|a|.
+The \verb|compressSpan| helper function gathers up all elements at the
 start of the tail \verb|as| that are equal to the head \verb|a|.  This
@@
 \verb|compressSpan| has type:
-\begin{verbatim}
+\begin{minted}{coq}
 forall(R : Set)(fo:FoldT (Alg (ListF A)) R), A -> R -> (nat * A) * R.
-\end{verbatim}
+\end{minted}
 \noindent It will be invoked by the code for \verb|mapThrough| with a
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition compressSpan : mappedT A (nat * A) :=
@@
   := fold (ListF A) RleCarr (FunConst (list (nat * A))) (RleAlg (List A)) xs.
-\end{verbatim}
+\end{minted}
 \caption{The function \texttt{rle} for run-length encoding, and the algebra \texttt{RleAlg} defining it
@@
 (morphism part of the functor) of type
-\begin{verbatim}
+\begin{minted}{coq}
 forall A B : Set, (A -> B) -> F A -> F B
-\end{verbatim}
+\end{minted}
 \noindent which satisfies the identity-preservation law for functors:
-\begin{verbatim}
+\begin{minted}{coq}
 fmapId : forall (A : Set)(d : F A), fmap (fun x => x) d = d
-\end{verbatim}
+\end{minted}
 \noindent Then we make the definitions of Figure~\ref{fig:mu}.  The critical
@@
 to have a definition like
-\begin{verbatim}
+\begin{minted}{coq}
   Inductive Mu' : Set := mu' : F Mu' -> Mu'.
-\end{verbatim}
+\end{minted}
 \noindent This is exactly what is used in approaches to modular
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
   Inductive Mu : Set := 
@@
   Lemma outIn(d : F Mu) : outMu (inMu d) = d.
-\end{verbatim}
+\end{minted}
 \caption{Derivation of retractive-positive recursive types (\texttt{Mu.v})}
@@
 this to a strictly positive approximation:
-\begin{verbatim}
+\begin{minted}{coq}
   Inductive Mu : Set := 
     mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
-\end{verbatim}
+\end{minted}
 \noindent Instead of taking in \verb|F Mu|, constructor \verb|mu|
@@
 would be, implicitly instantiating \verb|F| to \verb|ListF A|,
-\begin{verbatim}
+\begin{minted}{coq}
 Definition mkNil := mu (List A) (fun x => x) (NilF A)
-\end{verbatim}
+\end{minted}
 \noindent But given this, there are infinitely many other equivalent
 definitions.  For any \verb|Q : Set|, we could take
-\begin{verbatim}
+\begin{minted}{coq}
 Definition mkNil' := mu Q (fun x => mkNil) (NilF A)
-\end{verbatim}
+\end{minted}
 \noindent Since \verb|fmap f (NilF A)| equals just \verb|NilF B| for
@@
 Unrolling \verb|Subrec| gives us the type
-\begin{verbatim}
+\begin{minted}{coq}
 forall (X : Set -> Set) (FunX : Functor X), Alg Subrec X -> X Subrec
-\end{verbatim}
+\end{minted}
 \noindent So we see that \verb|Subrec| is the type of functions which,
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition SubrecF(C : Set) := 
@@
 Definition unroll: Subrec -> SubrecF Subrec.
-\end{verbatim}
+\end{minted}
 \caption{Definition of \texttt{Subrec} as a fixed-point of \texttt{SubrecF} (\texttt{Subrec.v})}
@@
   the fold function of type \verb|FoldT Alg R|. 
-\item For the \verb|eval : R -> X R| function, we pass \verb|(fold X xmap alg) : Subrec -> X Subrec|.
+\item For the \verb|rec : R -> X R| function, we pass \verb|(fold X xmap alg) : Subrec -> X Subrec|.
 \item For the subdata structure of type \verb|F R|, we pass \verb|d : F Subrec|. 
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition fold : FoldT Alg Subrec := 
@@
   fo F FunF (rollAlg (fun R' _ _ _ d => d)).
-\end{verbatim}
+\end{minted}
 \caption{The rest of the interface for \texttt{Subrec} (\texttt{Subrec.v})}
@@
 type.  \verb|AlgFi| and \verb|Algi| are indexed versions
-of the algebras we saw for recursion.  The \verb|eval| function
+of the algebras we saw for recursion.  The \verb|rec| function
 (Figure~\ref{fig:algf}) has now become an induction hypothesis: given
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition kMo := Subrec -> Prop.
@@
 Definition inni(i : Subrec)(fd : Fi Subreci i) : Subreci i.
-\end{verbatim}
+\end{minted}
 \caption{Interface for subsidiary induction (\texttt{Subreci.v})}
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition lkMo := List -> Prop.
@@
   FoldTi ListF (Algi ListF ListFi) R d.
-\end{verbatim}
+\end{minted}
 \caption{The indexed version \texttt{ListFi} of \texttt{ListF} (\texttt{List.v})}
@@
 
-\begin{verbatim}
+\begin{minted}{coq}
 Theorem RldRle (xs : list A): rld (rle (toList xs)) = xs.  
-\end{verbatim}
+\end{minted}
 
@@
 
-\begin{verbatim}
+\begin{minted}{coq}
 Lemma guardPres{R : List A -> Prop}(foi:forall d : List A, ListFoldTi R d)
@@
       (l:list A)(r : List A)(e: span p xs = (l,r)) : R r.
-\end{verbatim}
-
-\begin{figure}
-\begin{verbatim}
+\end{minted}
+
+\begin{figure}
+\begin{minted}{coq}
 Definition SpanAppendF(p : A -> bool)(xs : List A) : Prop :=
@@
     S r.
-\end{verbatim}
+\end{minted}
 \caption{Statements of three lemmas about \texttt{span} (directory \texttt{SpanPfs})}
@@
 \begin{figure}
-\begin{verbatim}
+\begin{minted}{coq}
 Definition spanForall2F(p : A -> bool)(xs : List A) : Prop :=
@@
   span p xs = (fromList xs, getNil xs).
-\end{verbatim}
+\end{minted}
 \caption{A statement of the property that \texttt{span} returns the empty suffix, computed using \texttt{getNil} to avoid noncanonicity problems, if all elements satisfy \texttt{p}}
@@
 arbitrary finite nestings of recursion, and we illustrate it in Coq
-with realistic examples.  Like them, we find that subsidiary recursion
-subsumes course-of-value recursion.  The technique of sized types is
-similar to our method in providing a type-based method for
-termination~\cite{barthe04}.  This technique enriches datatypes with
-abstract sizes, which must then be propagated through code, using
+with realistic examples.  It seems that generalizing the carriers of
+algebras to functors is the critical step enabling such examples.
+Like Uustalu and Vene, we find that subsidiary recursion subsumes
+course-of-value recursion.  Our method is also similar to the
+technique of sized types, in providing a type-based method for
+termination~\cite{barthe04}.  With sized types, datatypes are indexed
+with abstract sizes, which must then be propagated through code, using
 dependent types.  Our \verb|Subrec| does not require dependent types,
-resulting in just polymorphically typed code for the examples we saw
+resulting in just polymorphically typed code for our examples
 (\verb|Subreci|, for proofs, of course does use dependent types).

### 2022-01-26 paper/paper.tex
patching up code and paper to remove reveal functions
Also, the notion of 'anchor type' is gone now.  This is thanks to
Ben's observation that we seem not to be using 'reveal' anywhere.  It
simplifies things to remove it.

@@
 
-Algebras for our subsidiary recursion are more complex than $F$-algebras.  First, for
-reasons we will explain further below, the carrier of the algebra will
-be a functor \verb|X : Set -> Set|.  Second, algebras have a specified
-\emph{anchor type} \verb|C|, which we can think of as the datatype
-\emph{as viewed by a containing recursion} or else, if this is a
-top-level recursion, our development's version of the actual datatype
-(e.g., \verb|List A|, for some \verb|A|).  The algebra is presented with:
+Algebras for our subsidiary recursion are more complex than
+$F$-algebras.  For reasons we will explain further below, the carrier
+of the algebra will be a functor \verb|X : Set -> Set|.  The algebra
+is presented with:
 
@@
 \item a type \verb|R : Set|, which will be this recursion's view of the datatype.
-\item a function \verb|reveal : R -> C|, which reveals values of type \verb|R| as really having the anchor type.  
-\item a function \verb|fold : FoldT Alg R|, which allows one to initiate subsidiary recursions in which the anchor type is \verb|R|.
-Note that the algebra's anchor type is \verb|C|, but for subsidiary recursions the anchor type changes (to \verb|R|).
-We will present the type \verb|FoldT Alg R| below.
+\item a function \verb|fold : FoldT Alg R|, which allows one to
+  initiate subsidiary recursions over data of type \verb|R|.  We will
+  present the type \verb|FoldT Alg R| below.
 \item a function \verb|rec : R -> X R|, to use for making recursive calls, on any value of type \verb|R|.
@@
 
-As in previous work, we dub this interface a \emph{recursion
+We have previously dubbed this interface a \emph{recursion
 universe}~\cite{stump20}.  As in other domains using the term
-``universe'', we have a kind of space (here, \verb|R|) with operations
-that keep one in that space (for other cases: the ordinal $\epsilon_0$
-and $\omega^-$, the physical universe and traveling at the speed of
-light).  Staying in the recursion universe is good, because we may
-recurse (via \verb|rec|) on any value of type \verb|R|.  One can
-use \verb|reveal| to leave, but then \verb|rec| can no longer be used. Some points
-must still be explained: why \verb|X| has type \verb|Set -> Set|, and
-the definition of \verb|FoldT|.  Let us see these details next.
+``universe'', we have a kind of space (here, \verb|R|), which one
+cannot escape using certain operations.  Other examples are the
+ordinal $\epsilon_0$ and $\omega^-$, and the physical universe and
+traveling at the speed of light.  Staying in the recursion universe is
+good, because we may recurse (via \verb|rec|) on any value of type
+\verb|R|.  Some points must still be explained: why \verb|X| has type
+\verb|Set -> Set|, and the definition of \verb|FoldT|.  Let us see
+these details next.
 
@@
 We need a fixed-point here due to the input \verb|fold| (\verb|Algf|)
-of type \verb|FoldT Alg C|.  This is the type for fold functions which
-apply algebras (\verb|Alg|) to data of type \verb|C|, which we have
-already dubbed the \emph{anchor type} of the recursion.  At the top
-level of code, the anchor type would just be \verb|List A| (for
-example).  When one initiates a subsidiary recursion, though, the
-anchor type will instead by the abstract type \verb|R| for the outer
-recursion.  The variable \verb|Alg| occurs only positively (but not
+of type \verb|FoldT Alg R|.  This is the type for fold functions which
+apply algebras (\verb|Alg|) to data of type \verb|R|.  The variable \verb|Alg| occurs only positively (but not
 strictly positively) in \verb|AlgF|, because it occurs negatively in
-\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg C X|.  So
+\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg X|.  So
 we can indeed take a fixed-point of \verb|AlgF| to define the constant
@@
 \begin{minted}{coq}
-Definition KAlg  : Type := Set -> (Set -> Set) -> Set.
+Definition KAlg  : Type := (Set -> Set) -> Set.
 
@@
 
-Definition AlgF(Alg: KAlg)(C : Set)(X : Set -> Set) : Set :=
+Definition AlgF(Alg: KAlg)(X : Set -> Set) : Set :=
   forall (R : Set)
-         (reveal : R -> C)        
          (fold : FoldT Alg R)
@@
 Definition fold : FoldT Alg Subrec.
-Definition rollAlg :
-  forall {C : Set} {X : Set -> Set}, AlgF Alg C X -> Alg C X.
-Definition unrollAlg : 
-  forall {C : Set} {X : Set -> Set}, Alg C X -> AlgF Alg C X.
+Definition rollAlg : forall {X : Set -> Set}, AlgF Alg X -> Alg X.
+Definition unrollAlg : forall {X : Set -> Set}, Alg X -> AlgF Alg X.
 \end{minted}
@@
 recurse.  This is the first argument to a value of type
-\verb|AlgF Alg C X|.  Reasoning parametrically, an algebra can assume
+\verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
 nothing about \verb|R| except that it supports the following
-operations.  First there is \verb|reveal|, which turns an \verb|R|
-into a \verb|C|.  This reveals that the data of type \verb|R| are
-really values of the anchor type of this recursion.  Next we have
-\texttt{fold}, which will allow us to fold another algebra over data
+operations.  Next we have a local
+\texttt{fold} function, which will allow us to fold another algebra over data
 of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
@@
 As noted already, for subsidiary recursion, algebras have a carrier
-\verb|X| which depends (functorially) on a type.  This is so that (i)
-inside an inner recursion we may compute a result of some type that
-may mention \verb|R|, but (ii) outside that recursion, the result will
-mention the anchor type \verb|C|.  The \verb|rec| function returns
-something of type \verb|X R|, and so does the algebra itself; this
-demonstrates (i).  For (ii): if we look at the definition of
-\verb|FoldT| in the figure, we see that folding an algebra of type
-\verb|alg C X| over a value of type \verb|C| produces a result of type
-\verb|X C|.  Having a functor for the carrier of the algebra gives us
-the flexibility to type results inside a recursion with the abstract
-type \verb|R|, but view those results as having the anchor type
-\verb|C| outside the recursion.
-
-The last definitions in the figure are for \verb|fold|, which allows
-us to fold an \verb|Alg| over a value of type \verb|Subrec|; and for mapping
-between \verb|Alg| and its unfolding in terms of \verb|AlgF|.  We will
+\verb|X| which depends (functorially) on a type.  When we fold an
+algebra using a fold function (either global or local) of type
+\verb|FoldT Alg C|, (i) recursive calls may compute a result of type
+\verb|X R|, mentioning the abstract type \verb|R| for that recursion;
+and (ii) outside that recursion, the result will have type \verb|X C|.
+Having a functor for the carrier of the algebra gives us the
+flexibility to type results inside a recursion with the abstract type
+\verb|R|, but view those results as having the type \verb|C| outside
+the recursion.  The function \verb|fold| in the figure initiates
+top-level folds.  We also can have functions 
+between \verb|Alg| and its \verb|Algf|-unfolding.  We will
 return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.
@@
 \verb|fold|ing an algebra over constructed data \verb|inn d| is equal
-to invoking the algebra on the identity function for \verb|reveal|
-from the interface for algebras (Figure~\ref{fig:algf}); \verb|fold|
+to invoking the algebra on \verb|fold|
 for the fold function; an invocation of \verb|fold| with the algebra
@@
  forall (X : Set -> Set) (FunX : Functor X) (IdF : FmapId X FunX)
-        (algf : AlgF Alg Subrec X) (d : F Subrec),
+        (algf : AlgF Alg X) (d : F Subrec),
  fold X FunX (rollAlg algf) (inn d) =
-   algf _ (fun x => x) fold (fold X FunX (rollAlg algf)) d .
+   algf _ fold (fold X FunX (rollAlg algf)) d .
 \end{minted}
@@
 remaining suffix.  This is the typing for a top-level recursion.  More
-generally, though, given an anchor type \verb|R : Set| along with a
-fold function for that anchor type (i.e., of type
+generally, though, given a type \verb|R : Set| along with a
+fold function for that type (i.e., of type
 \verb|FoldT (Alg (ListF A)) R|), we would like to map an input list of
@@
 
-Figure~\ref{fig:spanalg} gives the algebra \verb|SpanAlg| for computing
-\verb|span|.  The type of \verb|SpanAlg p C| is
-\begin{verbatim}
-Alg (ListF A) C SpanF
-\end{verbatim}
-This states that we are defining an algebra (\verb|Alg|) for the
-\verb|ListF A| functor, with anchor type \verb|C| and carrier
-\verb|SpanF|.  \verb|SpanF| has type \verb|Set -> Set|, as required
-for the carriers of our algebras. The definition of \verb|SpanAlg| is
-actually parametrized by \verb|C|, which is good, as it means we can
-use \verb|SpanAlg| for top-level or subsidiary recursions.
-
-\begin{figure}
-\begin{minted}{coq}
-Definition SpanAlg(p : A -> bool)(C : Set)
-  : Alg (ListF A) C SpanF :=
-  rollAlg (fun R reveal fo span xs => 
+\begin{figure}
+\begin{minted}{coq}
+Definition SpanAlg(p : A -> bool)
+  : Alg (ListF A) SpanF :=
+  rollAlg (fun R fo span xs => 
      match xs with
@@
 
-Let us continue through the code for \verb|SpanAlg|
-(Figure~\ref{fig:spanalg}).  We use \verb|rollAlg| to create an algebra
-from something whose type is an application of \verb|AlgF|.  This
-takes in all the components of the recursion universe: the abstract
-type \verb|R|, the \verb|reveal| function (not needed in this case),
-the fold function (\verb|fo|) for any subsidiary recursions (also not
-needed here), a function we choose to name \verb|span| for making
-recursive calls, and finally \verb|xs : ListF A R|.  The algebra
-pattern-matches on this \verb|xs|.  In the cases where it is empty or
-where its head (\verb|hd|) does not satisfy \verb|p|, we return
-\verb|SpanNoMatch|.  This signals to the caller that we really wished
-to return \verb|([],xs)|, but could not because we do not have
-\verb|xs| at type \verb|R|.  If the head does satisfy \verb|p|, then
-we recurse on the tail (\verb|tl : R|) by calling the provided \verb|span : R -> SpanF R|.
-If \verb|span tl|
-returns \verb|SpanNoMatch|, that means that we should make \verb|tl|
-the suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily,
-we have \verb|tl : R| here, so we can do this.  In either case (for
+Figure~\ref{fig:spanalg} show the algebra
+\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
+an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
+\verb|SpanF| of the required type \verb|Set -> Set|.  We use
+\verb|rollAlg| to create an algebra from something whose type is an
+application of \verb|AlgF|.  This takes in all the components of the
+recursion universe: the abstract type \verb|R|, the fold function
+(\verb|fo|) for any subsidiary recursions (not needed here), a
+function we choose to name \verb|span| for making recursive calls, and
+finally \verb|xs : ListF A R|.  The algebra pattern-matches on this
+\verb|xs|.  In the cases where it is empty or where its head
+(\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
+This signals to the caller that we really wished to return
+\verb|([],xs)|, but could not because we do not have \verb|xs| at type
+\verb|R|.  If the head does satisfy \verb|p|, then we recurse on the
+tail (\verb|tl : R|) by calling the provided
+\verb|span : R -> SpanF R|.  If \verb|span tl| returns
+\verb|SpanNoMatch|, that means that we should make \verb|tl| the
+suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
+have \verb|tl : R| here, so we can do this.  In either case (for
 return value of \verb|span tl|), we add the head to the front of the
@@
 \verb|SpanSomeMatch l r|, then we return the nonempty
-prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| for subsidiary recursion (e.g., in \verb|wordsBy|, below).
+prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| 
+for subsidiary recursion (e.g., in \verb|wordsBy|, below).
 
@@
                  (p : A -> bool)(xs : R) : SpanF R :=
-  fo SpanF SpanFunctor (SpanAlg p R) xs.
+  fo SpanF SpanFunctor (SpanAlg p) xs.
 
@@
 Let us now see how to write \verb|wordsBy|, our example function from
-Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.
-The code is in Figure~\ref{fig:wordsby}, assuming a type
-\verb|A : Set|.  The setup is similar to that for \verb|span|.  We first define
-an algebra \verb|WordsBy|, parametrized by anchor type \verb|C| (and also
-the predicate \verb|p|), of type \verb|Alg (ListF A) C (Const (list (list A)))|.
- This says that \verb|WordsBy p C| is an algebra (\verb|Alg|)
-for the \verb|ListF A| functor, with anchor type \verb|C|, and carrier
-\verb|Const (list (list A))|.  \verb|Const| is the combinator for
-creating the object part of constant functors; \verb|FunConst| creates
-the morphism part (i.e., the \verb|fmap| function).  We use 
-\verb|Const| where the return type of the
-algebra will not depend on its abstract type \verb|R|.  Here, we are
+Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.  The code
+is in Figure~\ref{fig:wordsby}, assuming a type \verb|A : Set|.  The
+setup is similar to that for \verb|span|.  We first define an algebra
+\verb|WordsBy|, parametrized by the predicate \verb|p|, of type
+\verb|Alg (ListF A) (Const (list (list A)))|.  This says that
+\verb|WordsBy p| is an algebra (\verb|Alg|) for the \verb|ListF A|
+functor, with carrier \verb|Const (list (list A))|.  \verb|Const| is
+a combinator for creating the object part of constant functors;
+\verb|FunConst| creates the morphism part (i.e., the \verb|fmap|
+function).  We use \verb|Const| where the return type of the algebra
+will not depend on its abstract type \verb|R|.  Here, we are
 constructing from scratch a list of lists, so it will not be legal to
-recurse on the list itself, or its (list) elements.  So we just
-use the \verb|list| type of Coq's standard library.
+recurse on the list itself, or its (list) elements.  So we just use
+the \verb|list| type of Coq's standard library.
 
@@
 the return type of \verb|breakr| is \verb|list A * R|, where \verb|R|
-is the anchor type of the provided fold function \verb|fo|.  And
-\verb|fo| has type \verb|FoldT (ListF A) Alg R|, from the definition
+comes from the type \verb|FoldT (ListF A) Alg R| of
+\verb|fo|, from the definition
 of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
@@
 \begin{minted}{coq}
-  Definition WordsBy(p : A -> bool)(C : Set)
-    : Alg (ListF A) C (Const (list (list A))) :=
-    rollAlg (fun R reveal fo wordsBy xs => 
+  Definition WordsBy(p : A -> bool)
+    : Alg (ListF A) (Const (list (list A))) :=
+    rollAlg (fun R fo wordsBy xs => 
       match xs with
@@
                    (p : A -> bool)(xs : R) : list (list A) :=
-  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p R) xs.
+  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p) xs.
 \end{minted}
@@
 \begin{minted}{coq}
-Definition MapThroughAlg{B : Set}(f:mappedT A B)
-           (C : Set) : Alg (ListF A) C (Const (list B)) :=
-  rollAlg (fun R reveal fo mapThrough xs => 
+Definition MapThroughAlg{B : Set}(f:mappedT A B) : Alg (ListF A) (Const (list B)) :=
+  rollAlg (fun R fo mapThrough xs => 
     match xs with
@@
 \noindent It will be invoked by the code for \verb|mapThrough| with a
-fold function \verb|fo| with anchor type \verb|R|, and then has the
+fold function \verb|fo| of type \verb|FoldT (Alg (ListF A)) R|.  It then has the
 responsibility of extracting from the tail at type \verb|R| (second input) 
@@
 Definition RleCarr := Const (list (nat * A)).
-Definition RleAlg(C : Set) : Alg (ListF A) C RleCarr :=
-  MapThroughAlg compressSpan C.
+Definition RleAlg : Alg (ListF A) RleCarr :=
+  MapThroughAlg compressSpan.
 Definition rle(xs : List A) : list (nat * A)
-  := fold (ListF A) RleCarr (FunConst (list (nat * A))) (RleAlg (List A)) xs.
+  := fold (ListF A) RleCarr (FunConst (list (nat * A))) RleAlg xs.
 \end{minted}
@@
 \begin{minted}{coq}
-forall (X : Set -> Set) (FunX : Functor X), Alg Subrec X -> X Subrec
+forall (X : Set -> Set) (FunX : Functor X), Alg X -> X Subrec
 \end{minted}
 \noindent So we see that \verb|Subrec| is the type of functions which,
-for all algebras with anchor type \verb|Subrec| and functorial carrier
+for all algebras with functorial carrier
 \verb|X|, compute a value of type \verb|X Subrec|.  This is a
@@
 Definition SubrecF(C : Set) := 
-  forall (X : Set -> Set) (FunX : Functor X), Alg C X -> X C.
+  forall (X : Set -> Set) (FunX : Functor X), Alg X -> X C.
 Definition Subrec := Mu SubrecF.
@@
 The rest of the interface for \verb|Subrec| is shown in
-Figure~\ref{fig:subrecb}.  We have \verb|fold|, which is a fold
-function with anchor type \verb|Subrec|.  To fold an algebra
+Figure~\ref{fig:subrecb}.  To \verb|fold| an algebra
 \verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
@@
 \verb|X|, its fmap function \verb|xmap|, and an algebra \verb|alg|
-with that carrier.  Note that the anchor type of this algebra is
-\verb|Subrec|.  It will then call \verb|alg| (after \verb|unroll|ing
+with that carrier.  It will then call \verb|alg| (after \verb|unroll|ing
 it) with implementations for the components of the recursion universe
@@
   where the interface mentions \verb|R|.
-\item the identity function is passed as the value for \verb|reveal : R -> Subrec|.
-\item The function \verb|fold|, which expects an algebra with anchor type \verb|Subrec|, is passed as
+\item The function \verb|fold : FoldT Alg Subrec| is passed as
   the fold function of type \verb|FoldT Alg R|. 
@@
 Finally, Figure~\ref{fig:subrecb} defines \verb|out| as a subsidiary
-recursion, given any fold function with its anchor type \verb|R|.
+recursion, given a fold function.
 Outside the recursion, \verb|d| has type \verb|F R|; inside the
@@
   fun d => roll (fun X xmap alg =>
-                 unrollAlg alg Subrec (fun x => x) fold (fold X xmap alg) d).
+                 unrollAlg alg Subrec fold (fold X xmap alg) d).
 
 Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
-  fo F FunF (rollAlg (fun R' _ _ _ d => d)).
+  fo F FunF (rollAlg (fun R' _ _ d => d)).
 \end{minted}
@@
 We have seen how to write subsidiary recursions in Coq.  But can one
-reason about these?  To wrap up this paper, we will see an interface
-for subsidiary induction in Coq, and example proofs written using this
-interface.  Subsidiary induction is written just as the natural
-extension of subsidiary recursion, which worked over \verb|Set|s, to
-\verb|Subrec|-predicates.  The development is parametrized by a
-functor \verb|F| and a functor
+reason about these?  To wrap up this paper, we will briefly see the
+interface to our development of subsidiary induction in Coq, and
+example proofs written using this interface.  Subsidiary induction is
+the natural extension of subsidiary recursion, which worked over
+\verb|Set|s, to \verb|Subrec|-predicates.  The development is
+parametrized by a functor \verb|F| and a functor
 \verb|Fi : (Subrec -> Prop) -> (Subrec -> Prop)| over
@@
 Definition kMo := Subrec -> Prop.
-Definition KAlgi := kMo -> (kMo -> kMo) -> Set.
+Definition KAlgi := (kMo -> kMo) -> Set.
 Definition FoldTi(alg : KAlgi)(C : kMo) : kMo :=
   fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), 
-           alg C X -> C d -> X C d.
-
-Definition AlgFi(A: KAlgi)(C : kMo)(X : kMo -> kMo) : Set :=
+           alg X -> C d -> X C d.
+
+Definition AlgFi(A: KAlgi)(X : kMo -> kMo) : Set :=
   forall (R : kMo)
-    (reveal : (forall (d : Subrec), R d -> C d))        
     (fo : (forall (d : Subrec), FoldTi A R d))
@@
 Definition SubrecFi(C : kMo) : kMo := 
-  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), Algi C X -> X C d.
+  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), Algi X -> X C d.
 Definition Subreci := Mui Subrec SubrecFi.
@@
 state that all the elements of the prefix returned by \verb|span|
-satisfy \verb|p|.  These lemmas are proved using an indexed algebra
-where the indexed anchor type is not used (so the carriers are
-constant indexed functors returning the types shown).  But
-\verb|GuardPresF| uses the indexed anchor type (its argument \verb|S|),
-to express that whenever \verb|spanh| returns a suffix \verb|r|, that
-suffix satisfies the indexed anchor type.  This enables us to invoke
-an outer induction hypothesis on this suffix, when reasoning
-subsidiarily about \verb|span|.  Using these lemmas, we can write a short proof by
-subsidiary induction of the following, where
+satisfy \verb|p|.  These lemmas are proved using indexed algebras with
+constant (indexed) carriers.  In contrast, \verb|GuardPresF| uses its
+argument \verb|S| to express that whenever \verb|spanh| returns a
+suffix \verb|r|, that suffix satisfies \verb|S|.  This enables us to
+invoke an outer induction hypothesis on this suffix, when reasoning
+subsidiarily about \verb|span|.  Using these lemmas, we can write a
+short proof by subsidiary induction of the following, where
 \verb|rld : list (nat * A) -> list A| is the obvious decoding

### 2022-01-30 paper/paper.tex
bug fixes, small improvements
Almost in shape to submit to ITP 2022

@@
 Algebras for our subsidiary recursion are more complex than
-$F$-algebras.  For reasons we will explain further below, the carrier
-of the algebra will be a functor \verb|X : Set -> Set|.  The algebra
-is presented with:
+$F$-algebras.  Let us begin with an informal explanation.  For reasons
+we will explain further below, the carrier of the algebra will be a
+functor \verb|X : Set -> Set|.  The algebra is presented with:
 
@@
 
-Our development is parametrized by a signature functor \verb|F| of
-type \verb|Set -> Set|.  It implements \verb|Subrec : Set|, which is
-the type to use for subsidiary recursion.  This type comes with
+The type over which one can recurse with our scheme of subsidiary
+recursion is called \verb|Subrec|.  It is parametrized by a signature
+functor \verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
 \verb|inn : F Subrec -> Subrec|, which behaves computationally like a
-constructor.  \verb|Subrec.v| does not itself provide an induction
-principle based on \verb|inn|, however.  Induction is derived later
-(Section~\ref{sec:interfacei}). \verb|Subrec.v| makes critical use of
+constructor.  We will later derive an induction principle for this type
+(Section~\ref{sec:interfacei}). The definition of \verb|Subrec| uses
 retractive-positive recursive types, to take a fixed-point of a
@@
 one must define a signature functor for the datatype.  For lists, this
-is \verb|ListF| of Figure~\ref{fig:listf}.  \verb|List| is defined to
+is \verb|ListF|, which we saw at the start of Section~\ref{sec:interface}.  \verb|List| is defined to
 be \verb|Subrec|, with \verb|F| instantiated to \verb|ListF A|.  This
@@
 replacing them.  The figure also shows constructors \verb|mkNil| and
-\verb|mkCons| for \verb|List|, and types for conversion functions
-between \verb|List| and \verb|list| (code elided). One direction uses
-Coq's structural recursion, the other uses subsidiary recursion, which
-we will see next.
+\verb|mkCons| for \verb|List|, and typings for conversion functions
+between \verb|List| and \verb|list| (code elided).
 
@@
 Finally, for a recursion scheme, one would like to see not just the
-typed interface, but also the computation law.  This is stated as a
+typed interface, but also the computation law.  This is shown as a
 theorem in Figure~\ref{fig:comp}.  Intuitively, it states that
@@
 \end{minted}
+\caption{Computation law for subsidiary recursion, stated as a theorem}
+\label{fig:comp}
 \end{figure}
@@
 \begin{minted}{coq}
-Definition SpanAlg(p : A -> bool)
-  : Alg (ListF A) SpanF :=
+Definition SpanAlg(p : A -> bool) : Alg (ListF A) SpanF :=
   rollAlg (fun R fo span xs => 
@@
       end).
-
 Definition wordsByr{R : Set}(fo:FoldT (Alg (ListF A)) R)
@@
 \begin{minted}{coq}
-Definition MapThroughAlg{B : Set}(f:mappedT A B) : Alg (ListF A) (Const (list B)) :=
+Definition MapThroughAlg{B : Set}(f:mappedT A B)
+  : Alg (ListF A) (Const (list B)) :=
   rollAlg (fun R fo mapThrough xs => 
@@
     end).
-
 Definition mapThroughr{R : Set}(fo:FoldT (Alg (ListF A)) R)
@@
 \end{minted}
-\noindent It will be invoked by the code for \verb|mapThrough| with a
-fold function \verb|fo| of type \verb|FoldT (Alg (ListF A)) R|.  It then has the
+\noindent It is invoked by the code for \verb|mapThrough| with \verb|fo : FoldT (Alg (ListF A)) R|.  It then has the
 responsibility of extracting from the tail at type \verb|R| (second input) 
@@
 
-The definition of \verb|Mu| weakens
-this to a strictly positive approximation:
-\begin{minted}{coq}
-  Inductive Mu : Set := 
-    mu : forall (R : Set), (R -> Mu) -> F R -> Mu.
-\end{minted}
-\noindent Instead of taking in \verb|F Mu|, constructor \verb|mu|
+The definition of \verb|Mu| in Figure~\ref{fig:mu} weakens
+this to a strictly positive approximation. Instead of taking in \verb|F Mu|, constructor \verb|mu|
 accepts an input of type \verb|F R|, for some type \verb|R| for which
 we have a function of type \verb|R -> Mu|.  The impredicative
-quantification of \verb|R| is essential here: we instantiate it with
+quantification of \verb|R| is essential here: we will instantiate it with
 \verb|Mu| itself in the definition of \verb|inMu|
@@
 \end{minted}
-\noindent Since \verb|fmap f (NilF A)| equals just \verb|NilF B| for
+\noindent Since \verb|fmap f (NilF A)| equals \verb|NilF B| for
 \verb|f : A -> B|, if we apply \verb|outMu| (of Figure~\ref{fig:mu})
@@
 with realistic examples.  It seems that generalizing the carriers of
-algebras to functors is the critical step enabling such examples.
-Like Uustalu and Vene, we find that subsidiary recursion subsumes
-course-of-value recursion.  Our method is also similar to the
+algebras to functors is the critical step enabling such examples. In
+our Coq development, we show that Uustalu and Vene's scheme is
+derivable from subsidiary recursion. Our method is also similar to the
 technique of sized types, in providing a type-based method for
@@
 type $\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to
-  X}$. We have adopted this idea to the constructor of the type
+  X}$. We have applied this idea to the constructor of the type
 \verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
@@
 encoding~\cite{firsov+18b,firsov+18a}.  Here, we do not derive
-inductive types using these methods, but rather apply them to justify
+inductive types, but rather 
 a terminating recursion scheme for existing datatypes.

### 2022-01-31 paper/paper.tex
tweaks to related work

@@
 advocated in the literature on programming languages under the name
-\emph{strong functional programming}~\cite{turner95}.  Uustalu and
-Vene developed a categorical view of a recursion scheme allowing one
-level of subsidiary recursion, and illustrated it in Haskell with an
-artificial example~\cite{uustalu11}.  In contrast, our scheme allows
-arbitrary finite nestings of recursion, and we illustrate it in Coq
-with realistic examples.  It seems that generalizing the carriers of
-algebras to functors is the critical step enabling such examples. In
-our Coq development, we show that Uustalu and Vene's scheme is
-derivable from subsidiary recursion. Our method is also similar to the
-technique of sized types, in providing a type-based method for
-termination~\cite{barthe04}.  With sized types, datatypes are indexed
-with abstract sizes, which must then be propagated through code, using
-dependent types.  Our \verb|Subrec| does not require dependent types,
-resulting in just polymorphically typed code for our examples
-(\verb|Subreci|, for proofs, of course does use dependent types).
+\emph{strong functional programming}~\cite{turner95}.  Our method is
+similar to the technique of sized types, in providing a type-based
+method for termination~\cite{barthe04}.  With sized types, datatypes
+are indexed with abstract sizes, which must then be propagated through
+code, using dependent types.  In contrast, our approach relies just on
+polymorphism, and does not require dependent types for writing
+subsidiary recursions.  (\verb|Subreci|, for reasoning about such
+recursions, of course does use dependent types).
+
+Uustalu and Vene developed a categorical view of a recursion scheme
+allowing one level of subsidiary recursion, and illustrated it in
+Haskell with an artificial example~\cite{uustalu11}.  In contrast, our
+scheme allows arbitrary finite nestings of recursion, and we
+illustrate it in Coq with realistic examples.  It seems that
+generalizing the carriers of algebras to functors is the critical step
+enabling such examples. In our Coq development, we show that Uustalu
+and Vene's scheme is derivable from subsidiary recursion.
 

### 2022-02-02 paper/paper.tex
Pass through introduction.

@@
   may be initiated within outer ones, in such a way that outer
-  recursive calls may be made on results from inner ones.  The
+  recursive calls may be made on the results from inner ones.  The
   derivation utilizes a novel (necessarily weakened) form of
@@
 \end{minted}
-\noindent The sole recursive call is \verb|span p xs|, and it occurs
-in a clause where the input list is of the form \verb|x:xs|.  Hence it
-is structurally decreasing.  In the appropriate syntax, this
+\noindent The sole recursive call is \verb|span p xs|. This call is
+structurally decreasing, since it occurs in a clause where the input
+list is of the form \verb|x:xs|.  In the appropriate syntax, this
 definition can be accepted without additional effort by all the
-mentioned provers.
-
-This paper is about a more expressive form of terminating recursion,
+previously mentioned provers.
+
+This paper presents a more expressive form of terminating recursion,
 called \textbf{subsidiary recursion}.  While performing an outer
 recursion on some input \verb|x|, one may initiate an inner recursion
-on \verb|x| (or possibly some of its subdata), preserving the
-possibility of further invocations of the outer recursive function.
-Let us see a simple example.  The function \verb|wordsBy|
-(\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
-elements do not satisfy a predicate \verb|p|.  For example,
-\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|.
-Code is in Figure~\ref{fig:wordsBy}.  Recall that
-\verb|break p| is equivalent to \verb|span (not . p)|. The first recursive call,
-\verb|wordsBy p tl|, is structural.  But in the second, we invoke
-\verb|wordsBy p| on a value obtained from another recursion, namely
-\verb|span|.  This is not allowed under structural termination, but
-will be permitted by subsidiary recursion.
+on \verb|x| (or possibly some of its subdata), while still preserving
+the possibility of further invocations of the outer recursive
+function.  As a simple example, consider the function \verb|wordsBy|
+(\verb|Data.List.Extra|), which breaks a list into a set of its
+maximal sublists whose elements do not satisfy a predicate \verb|p|.
+For example, \verb|wordsBy isSpace " good day "| returns
+\verb|["good","day"]|. The code for this function is in
+Figure~\ref{fig:wordsBy}, where \verb|break p| is defined as
+\verb|span (not . p)|. The first recursive call, \verb|wordsBy p tl|,
+is structurally decreasing. In contrast, the second invocation of
+\verb|wordsBy p| uses a value obtained from a call to \verb|span|.
+This recursive call is not allowed under the structural termination
+condition, but is permitted by subsidiary recursion.
 
@@
 other modifications to Coq, except the \verb|-impredicative-set|
-flag. Using this derivation, we present several example functions like
-\verb|wordsBy|, and prove theorems about them.  A nice example is a
-definition of run-length encoding using \verb|span| as a subsidiary
-recursion, where we prove that encoding and then decoding returns the
-original list.  Our approach applies to the standard datatypes in the
-Coq library, and does not require switching libraries or datatype
-definitions.
-
-An important technical novelty is a derivation of a weakened form of
-positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
-datatypes $D$ to be strictly positive: in the input types of
-constructors of $D$, $D$ cannot occur to the left of any arrows.  Our
-derivation needs to use positive-recursive types, where $D$ may occur
-to the left of an even number (only) of arrows.  We present a way to
-derive a weakened form of positive-recursive type that is sufficient
-for our examples (Section~\ref{sec:mu}).  The weakening is to require
-only that $F\ (\mu F)$ is a retract of $\mu F$.  Usually these types are
-isomorphic.  Hence, we dub these \textbf{retractive-positive}
-recursive types.  This weakening leads to noncanonical elements of
-$\mu$, but we will see how to work around this.  Our definition of
-retractive-positive recursive types makes essential use of
-impredicative quantification, and hence is not legal in predicative
-theories like Agda's.
+flag. Using this derivation, we present several example functions in
+the vein of \verb|wordsBy|, and prove theorems about their behavior.  A nice
+example is a definition of run-length encoding using \verb|span| as a
+subsidiary recursion, where we prove that encoding and then decoding
+returns the original list.  Our approach applies to the standard
+datatypes in the Coq library, and does not require switching libraries
+or datatype definitions.
+
+An important technical novelty of our development is a derivation of a
+weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
+restrict datatypes $D$ to be strictly positive: in the argument types
+of constructors of $D$, $D$ cannot occur to the left of any arrows.
+Our derivation needs to use positive-recursive types, where $D$ may
+also occur to the left of an even number (only) of arrows.  We present
+a way to derive a weakened form of positive-recursive type that is
+sufficient for our examples (Section~\ref{sec:mu}).  The weakening is
+to require only that $F\ (\mu F)$ is a retract of $\mu F$.  (Usually
+these types are isomorphic.)  Hence, we dub these
+\textbf{retractive-positive} recursive types.  This weakening leads to
+noncanonical elements of $\mu$, but we will see how to work around
+this.  Our definition of retractive-positive recursive types makes
+essential use of impredicative quantification, and hence is not legal
+in predicative theories like Agda's.
 
 We begin by summarizing the interface our derivation provides for
-subsidiary recursion (Section~\ref{sec:interface}), and then see
-examples (Section~\ref{sec:examples}).  We next explain how the
-interface is actually implemented (Section~\ref{sec:deriv}), including
-our retractive-positive recursive types (Section~\ref{sec:mu}).  The
-interface for subsidiary induction is covered next
-(Section~\ref{sec:interfacei}), and example proofs using it
-(Section~\ref{sec:examplesi}).  Related work is discussed in
-Section~\ref{sec:related}.
+subsidiary recursion (Section~\ref{sec:interface}), and we then
+present some examples (Section~\ref{sec:examples}).  We next explain
+how the interface is actually implemented (Section~\ref{sec:deriv}),
+including our definition of retractive-positive recursive types
+(Section~\ref{sec:mu}).  We then cover the interface for subsidiary
+induction (Section~\ref{sec:interfacei}), and give some example proofs
+using it (Section~\ref{sec:examplesi}). We conclude with a discussion
+of related work (Section~\ref{sec:related}).
 

### 2022-02-03 paper/paper.tex
Additional edits through Section 2.

@@
 present some examples (Section~\ref{sec:examples}).  We next explain
-how the interface is actually implemented (Section~\ref{sec:deriv}),
+the implementation of our interface (Section~\ref{sec:deriv}),
 including our definition of retractive-positive recursive types
-(Section~\ref{sec:mu}).  We then cover the interface for subsidiary
+% (Section~\ref{sec:mu})
+.  We then cover the interface for subsidiary
 induction (Section~\ref{sec:interfacei}), and give some example proofs
-using it (Section~\ref{sec:examplesi}). We conclude with a discussion
+that use it (Section~\ref{sec:examplesi}). We conclude with a discussion
 of related work (Section~\ref{sec:related}).
@@
 algebra and category theory to describe inductive datatypes and their
-recursion principles (cf.~\cite{traytel12,cockett92,hagino87}).  On this approach, one describes transformations
-to be performed on data as \emph{algebras}, which can then be
-\textit{folded} over data.  The simplest form of algebras, namely
-$F$-algebras for functor $F$, are morphisms from $F\ A$ to $A$, for carrier object
-$A$.  From a programming perspective, an $F$-algebra is given input of
-type $F\ A$, and must compute a result of type $A$.  An example of $F$
-is the signature functor for lists, which we will use below:
-
-\begin{minted}{coq}
-Inductive ListF(X : Set) : Set :=
+recursion principles (cf.~\cite{traytel12,cockett92,hagino87}). Under
+this approach, inductive datatypes are equipped with a signature
+functor which explains how to 'unroll' one layer of the type. The
+signature functor for lists is:
+\begin{minted}{coq}
+Inductive ListF (X : Set) : Set :=
 | Nil : ListF X
@@
 \end{minted}
-
+Transformations on data with functor $F$ are described as
+\emph{algebras} which are then \textit{folded} over data.  The
+simplest form of algebras are morphisms from $F\ A$ to $A$, for some
+carrier object $A$.  From a programming perspective, an $F$-algebra is
+given input of type $F\ A$, and must compute a result of type $A$. The
+$ListF$-algebra for computing the length of a list is:
+\begin{minted}{coq}
+Definition LengthAlg (l : ListF nat) : nat :=
+match l with
+| Nil => 0
+| Cons a len' => 1 + len'
+end.
+\end{minted}
 Algebras for our subsidiary recursion are more complex than
@@
 we will explain further below, the carrier of the algebra will be a
-functor \verb|X : Set -> Set|.  The algebra is presented with:
+functor \verb|X : Set -> Set|.  Our algebras are presented with:
 
@@
 
-We will use Coq inductive types for the signature functors \verb|F| of
-various datatypes, thus enabling recursions to use Coq's
-pattern-matching on the subdata structure \verb|d|.  So the style of
-coding against this interface retains a similar feel to structural
-recursion.  Unlike with structural termination, though, the interface
-here is type-based and hence compositional.
+We use Coq inductive types for the signature functors \verb|F| of
+various datatypes. This allows algebras to use Coq's
+pattern-matching on the subdata structure \verb|d|, enabling a style
+of coding that retains a similar feel to structural recursion. Unlike
+structurally recursive functions, however, this interface uses a
+type-based (and hence compositional) criteria to ensure termination.
 
@@
 ``universe'', we have a kind of space (here, \verb|R|), which one
-cannot escape using certain operations.  Other examples are the
-ordinal $\epsilon_0$ and $\omega^-$, and the physical universe and
-traveling at the speed of light.  Staying in the recursion universe is
-good, because we may recurse (via \verb|rec|) on any value of type
-\verb|R|.  Some points must still be explained: why \verb|X| has type
-\verb|Set -> Set|, and the definition of \verb|FoldT|.  Let us see
-these details next.
+cannot escape using the provided operations.  Other examples include the
+ordinal $\epsilon_0$ and operator $\omega^-$, and the physical
+universe and traveling at the speed of light.  Staying in the
+recursion universe is good, because we may recurse (via \verb|rec|) on
+any value of type \verb|R|.  Some points must still be explained: in
+particular why \verb|X| has type \verb|Set -> Set|, and the definition
+of \verb|FoldT|. We consider those details next.
 
@@
 \end{minted}
-  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
+  \caption{Some basics from \texttt{List.v}, specializing the
+    functor-generic derivation of subsidiary recursion to
+    lists parameterized by an element type \texttt{A}  (\texttt{List.v})}
   \label{fig:listf}
@@
 
-The type over which one can recurse with our scheme of subsidiary
-recursion is called \verb|Subrec|.  It is parametrized by a signature
+The type over which one can recurse using our subsidiary
+recursion scheme is called \verb|Subrec|.  It is parametrized by a signature
 functor \verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
@@
 
-For our examples, we will consider the specialization to the case of
-lists, parametrized by the type \verb|A| of elements.  In general, to
-use our development to get subsidiary recursion over some datatype,
-one must define a signature functor for the datatype.  For lists, this
-is \verb|ListF|, which we saw at the start of Section~\ref{sec:interface}.  \verb|List| is defined to
-be \verb|Subrec|, with \verb|F| instantiated to \verb|ListF A|.  This
+Our examples specialize this scheme to the case of polymorphic lists,
+presented in Figure~\ref{fig:listf}.  In general, to use our
+development to get subsidiary recursion over some datatype, one must
+define a signature functor for the datatype, e.g., \verb|ListF| from
+the start of this section. We define \verb|List| to be
+\verb|Subrec| with \verb|F| instantiated to \verb|ListF A|.  This
 type \verb|List| is not to be confused with the type \verb|list| of
-lists in Coq's standard library.  As noted previously, our development
+lists from Coq's standard library.  As noted previously, our development
 is meant to be used in extension of existing inductive datatypes, not
@@
 \verb|mkCons| for \verb|List|, and typings for conversion functions
-between \verb|List| and \verb|list| (code elided).
+between \verb|List| and \verb|list| (we elide their definitions).
 
@@
 
-We need a fixed-point here due to the input \verb|fold| (\verb|Algf|)
-of type \verb|FoldT Alg R|.  This is the type for fold functions which
+We need a fixed-point here due to the \verb|fold| parameter
+\verb|AlgF| expecting an argument of type \verb|FoldT Alg R|.  This is the type for fold functions which
 apply algebras (\verb|Alg|) to data of type \verb|R|.  The variable \verb|Alg| occurs only positively (but not
@@
 \verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
-nothing about \verb|R| except that it supports the following
-operations.  Next we have a local
-\texttt{fold} function, which will allow us to fold another algebra over data
+nothing about \verb|R| except that it supports the
+\verb|fold| and \verb|rec| operations that follow. The first is a local
+\verb|fold| function, which allows us to fold another algebra over data
 of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
-recursions.  Then there is \verb|rec|, for recursive calls on data of
-type \verb|R|.
+recursions.  The second, \verb|rec|, is used to make recursive calls
+on values of type \verb|R|.
 
@@
 \verb|FoldT Alg C|, (i) recursive calls may compute a result of type
-\verb|X R|, mentioning the abstract type \verb|R| for that recursion;
+\verb|X R|, mentioning the abstract type \verb|R| inside the recursion;
 and (ii) outside that recursion, the result will have type \verb|X C|.
@@
 \verb|R|, but view those results as having the type \verb|C| outside
-the recursion.  The function \verb|fold| in the figure initiates
-top-level folds.  We also can have functions
-between \verb|Alg| and its \verb|Algf|-unfolding.  We will
+the recursion.  The function \verb|fold| in Figure~\ref{fig:algf} initiates
+top-level folds.  We can also have functions
+between \verb|Alg| and its \verb|AlgF|-unfolding.  We will
 return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

### 2022-02-03 paper/paper.tex
Additional polishing through Section 3.

@@
 prefix whose elements satisfy \verb|p|, and the second is the
-remaining suffix.  This is the typing for a top-level recursion.  More
-generally, though, given a type \verb|R : Set| along with a
-fold function for that type (i.e., of type
-\verb|FoldT (Alg (ListF A)) R|), we would like to map an input list of
+remaining suffix. Note that this is the typing for the top-level function.  More
+generally, though, given a type \verb|R : Set| and a
+fold function for \verb|R| (i.e., of type
+\verb|FoldT (Alg (ListF A)) R|), we want to map an input list of
 type \verb|R| to a pair of type \verb|list A * R|.  The first
@@
 cannot have type \verb|R|; we cannot statically ensure that outer
-recursions on it are legal.  But the second component will be a
-subdatum of the input list, and so can still have type \verb|R|,
+recursions on it are legal.  Its second component will always be a
+subdatum of the input list, and can thus have the type \verb|R|,
 enabling outer recursive calls.  So we want:
@@
 \end{minted}
-\noindent The idea is that the recursion will signal if it is in the
-one tricky case where \verb|p| does not match the first element, by
-returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
-via \verb|SpanSomeMatch|, a prefix and the suffix at type
+\noindent The idea is that the recursion will return
+\verb|SpanNoMatch| to signal that it is in the one tricky case where
+\verb|p| does not match the first element.  Otherwise, it will be able
+to return, via \verb|SpanSomeMatch|, a prefix and the suffix at type
 \verb|R|. The prefix will be nonempty, and hence the suffix will be at
@@
 
-Figure~\ref{fig:spanalg} show the algebra
-\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
-an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
-\verb|SpanF| of the required type \verb|Set -> Set|.  We use
-\verb|rollAlg| to create an algebra from something whose type is an
-application of \verb|AlgF|.  This takes in all the components of the
-recursion universe: the abstract type \verb|R|, the fold function
-(\verb|fo|) for any subsidiary recursions (not needed here), a
-function we choose to name \verb|span| for making recursive calls, and
-finally \verb|xs : ListF A R|.  The algebra pattern-matches on this
-\verb|xs|.  In the cases where it is empty or where its head
+Figure~\ref{fig:spanalg} shows the algebra \verb|SpanAlg|, whose type
+is \verb|Alg (ListF A) SpanF|.  So we are defining an algebra
+(\verb|Alg|) for the \verb|ListF A| functor, with carrier \verb|SpanF|
+of the required type \verb|Set -> Set|.  We use \verb|rollAlg| to
+create an algebra from something whose type is an application of
+\verb|AlgF|.  This takes in all the components of the recursion
+universe: the abstract type \verb|R|, the fold function (\verb|fo|)
+for any subsidiary recursions (not used here), a function we choose to
+name \verb|span| for making recursive calls, and finally
+\verb|xs : ListF A R|.  The algebra does case analysis on this last
+argument.  In the cases where it is empty or where its head
 (\verb|hd|) does not satisfy \verb|p|, we return \verb|SpanNoMatch|.
@@
 suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
-have \verb|tl : R| here, so we can do this.  In either case (for
-return value of \verb|span tl|), we add the head to the front of the
-prefix.
+have \verb|tl : R| here, so we can do this.  For either possible
+return value of \verb|span tl|, we add the current head \verb|hd| to
+the front of the prefix.
 
@@
 \subsection{The \texttt{wordsBy} function (\texttt{WordsBy.v})}
-
-Let us now see how to write \verb|wordsBy|, our example function from
-Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.  The code
+We now consider how to write the \verb|wordsBy| function from
+Section~\ref{sec:intro}, by using \verb|breakr| subsidiarily.  The code
 is in Figure~\ref{fig:wordsby}, assuming a type \verb|A : Set|.  The
 setup is similar to that for \verb|span|.  We first define an algebra
-\verb|WordsBy|, parametrized by the predicate \verb|p|, of type
+\verb|WordsByAlg|, parametrized by a predicate \verb|p|, of type
 \verb|Alg (ListF A) (Const (list (list A)))|.  This says that
-\verb|WordsBy p| is an algebra (\verb|Alg|) for the \verb|ListF A|
+\verb|WordsByAlg p| is an algebra (\verb|Alg|) for the \verb|ListF A|
 functor, with carrier \verb|Const (list (list A))|.  \verb|Const| is
@@
 function).  We use \verb|Const| where the return type of the algebra
-will not depend on its abstract type \verb|R|.  Here, we are
-constructing from scratch a list of lists, so it will not be legal to
+will not depend on its abstract type \verb|R|.  Since we are
+constructing a list of lists from scratch, it will not be legal to
 recurse on the list itself, or its (list) elements.  So we just use
@@
 
-The code for \verb|WordsBy| is essentially
-the same as what we saw in Section~\ref{sec:intro}.  We pattern match
-on \verb|xs : ListF A R|.  Recall that for this function, we are
-trying to drop elements which satisfy \verb|p|, and return a list of
-the sublists between maximal sequences of such elements.  In the
-\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we are
-supposed to drop it and recurse.  This is legal, because \verb|tl : R|
-and \verb|wordsBy : R -> list (list A)|.  In the \verb|else| case,
-we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
-does not satisfy \verb|p|, and the remaining suffix \verb|z|.
+The code for \verb|WordsByAlg| is essentially the same as what we saw
+in Section~\ref{sec:intro}.  Recall that this function drop elements
+which satisfy \verb|p|, and returns a list of the sublists between
+maximal sequences of such elements. This algebra pattern matches on
+\verb|xs : ListF A R|. In the \verb|Cons| case, if the head
+(\verb|hd|) satisfies the predicate, then we drop it and recurse.
+Note that this recursive call is legal because \verb|tl| and
+\verb|wordsBy| have type \verb|R| and \verb|R -> list (list A)|,
+respectively. Otherwise , we use \verb|breakr| to obtain the maximal
+prefix \verb|w| of \verb|tl| that does not satisfy \verb|p|, and the
+remaining suffix \verb|z|.
 
@@
 with \verb|ListF A|).  This means that from the invocation of
-\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
-can indeed apply \verb|wordsBy : R -> list (list A)| to
-\verb|z| to recurse.  The figure also shows the code for the subsidiary
+\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  Thus, it is
+legal to recursively apply \verb|wordsBy : R -> list (list A)| to
+\verb|z|. The figure also shows the code for the subsidiary
 recursion \verb|wordsByr|.
@@
 \begin{minted}{coq}
-  Definition WordsBy(p : A -> bool)
+Definition WordsByAlg (p : A -> bool)
     : Alg (ListF A) (Const (list (list A))) :=
@@
                    (p : A -> bool)(xs : R) : list (list A) :=
-  fo (Const (list (list A))) (FunConst (list (list A))) (WordsBy p) xs.
+  fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
 \end{minted}
@@
 
-In this example, we see how to write a combinator that factors out a
-subsidiary recursion.  The Haskell library \verb|Data.List.Extra| has
-a function \verb|repeatedly|, defined essentially as in
-Figure~\ref{fig:mapthroughhs}, though we attempt a more informative
-name.  This is like the standard \verb|map| function on lists, except
+This example shows how to write a combinator that factors out a
+subsidiary recursion.  The Haskell library \verb|Data.List.Extra| defines
+a function \verb|repeatedly|, in essentially the same way as \verb|mapThrough| in
+Figure~\ref{fig:mapthroughhs}, although we attempt a more informative
+name.  This function behaves like the standard \verb|map| function on lists, except
 that the function \verb|f| that we are mapping (or ``mapping
@@
 the output list, and whatever other list it wishes, upon which
-\verb|mapThrough| will recurse.
+\verb|mapThrough| will then recurse.
 
@@
 mapped function can initiate a subsidiary recursion, returning a value
-in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  So
-the type we will use for mapped functions is:
+in the abstract type \verb|R| of \verb|mapThrough|'s recursion.
+The type we will use for mapped functions is:
 \begin{minted}{coq}
@@
 
-Given this definition, the code is in Figure~\ref{fig:mapthrough}.
-\verb|MapThroughAlg| is similar to the
-Haskell code above, though when we call \verb|f|, we must
-supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
-from the definition of \verb|mappedT|, we have that \verb|b : B| and
-\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> list B|
+Given this definition, the Coq definition of \verb|mapThrough| is
+presented in Figure~\ref{fig:mapthrough}.  \verb|MapThroughAlg| is
+similar to the Haskell code above, though when we call \verb|f|, we
+must supply the abstract type \verb|R| and fold function \verb|fo|.
+From the definition of \verb|mappedT|, we have that \verb|b : B| and
+\verb|c : R|, so we may indeed invoke \verb|mapThrough : R -> list B|
 on \verb|c|.  Note that as we are building up a new list from scratch
@@
 \end{minted}
-\caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
-  code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
-  we omit it (\texttt{MapThrough.v})}
+\caption{The algebra \texttt{MapThroughAlg} defining the functions
+  \texttt{mapThrough} and \texttt{mapThroughr}. We omit the code for
+  those two functions as it follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby})}
 \label{fig:mapthrough}
@@
 Finally, we have an example using our \verb|mapThrough| combinator
-together with a subsidiary recursion, to implement \emph{run-length
-encoding}.  This is a basic data-compression algorithm where maximal
+together with a subsidiary recursion, to implement a \emph{run-length
+encoding} of lists.  This is a basic data-compression algorithm where maximal
 sequences of $n$ occurrences of element $e$ are summarized by the pair
-$(n,e)$~\cite{datacomp}.  Haskell code is in Figure~\ref{fig:rlehs}.
+$(n,e)$~\cite{datacomp}.  The Haskell code for this function is shown in Figure~\ref{fig:rlehs}.
 Recall that \verb|(== a)| tests its input for equality with \verb|a|.
@@
 \verb|a :: p|.  The \verb|mapThrough| combinator then iterates
-\verb|compressSpan| through the suffix \verb|s|.
+\verb|compressSpan| over the suffix \verb|s|.
 
@@
 will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
-\verb|compressSpan| has type:
+we see that \verb|compressSpan| has the type:
 \begin{minted}{coq}
@@
 \end{minted}
-\noindent It is invoked by the code for \verb|mapThrough| with \verb|fo : FoldT (Alg (ListF A)) R|.  It then has the
-responsibility of extracting from the tail at type \verb|R| (second input)
-a result upon which \verb|mapThrough| should recurse (second component
-of the output pair).  Then we define an algebra \verb|RleAlg| by
-supplying \verb|compressSpan| as the function to map through, to
-\verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).  Following the
-pattern seen above, we define function \verb|rle| for top-level
-recursions using \verb|fold| (we could also define a subsidiary version
-\verb|rler|).
+\noindent It is invoked by the code for \verb|mapThrough| with
+\verb|fo : FoldT (Alg (ListF A)) R|.  Note that \verb|compressSpan|
+has the responsibility of extracting from the tail at type \verb|R|
+(second input) a result upon which \verb|mapThrough| should recurse
+(second component of the output pair). Then we can define an algebra
+\verb|RleAlg| by supplying \verb|compressSpan| as the function to map
+through, to \verb|MapThroughAlg| (Figure~\ref{fig:mapthrough}).
+Following the pattern seen above, we can write the top-level function
+\verb|rle| using \verb|fold| (we could also define a subsidiary
+version \verb|rler|).
 

### 2022-02-04 paper/paper.tex

@@
 require algebras, and also for the definition of \verb|Subrec| (which
-we will see in more detail in the next section).  Full
+we will see in more detail in the next section).  Unfortunately, Full
 positive-recursive types are incompatible with Coq's type
@@
 types~\cite{blanqui05}, but this requires changing the underlying
-theory.  Here we exploit Coq's
-impredicative polymorphism for a different solution.
-
-Assume \verb|F : Set -> Set|, with an \verb|fmap| function
-(morphism part of the functor) of type
+theory.  Here we exploit Coq's impredicative polymorphism to obtain a
+different solution.
+
+Our starting point is a Functor \verb|F : Set -> Set| equipped with an
+\verb|fmap| function (functor morphism) of type
 \begin{minted}{coq}
@@
 \end{minted}
-\noindent Then we make the definitions of Figure~\ref{fig:mu}.  The critical
-idea is embodied in the definition of \verb|Mu|.  Ideally, we would like
-to have a definition like
+\noindent Ideally, we would like to have a definition like
 \begin{minted}{coq}
@@
 \end{minted}
-\noindent This is exactly what is used in approaches to modular
-datatypes in functional programming, like
-Swierstra's~\cite{swierstra08}.  But this definition is (rightly)
-rejected by Coq, as instantiations of
-\verb|F| that are not strictly positive would be unsound.
+\noindent This is exactly what is used in many approaches to modular
+datatypes in functional programming, as in
+Swierstra~\cite{swierstra08}.  But this definition is (rightly)
+rejected by Coq, as instantiations of \verb|F| that are not strictly
+positive would be unsound.
 
@@
 
-The definition of \verb|Mu| in Figure~\ref{fig:mu} weakens
-this to a strictly positive approximation. Instead of taking in \verb|F Mu|, constructor \verb|mu|
-accepts an input of type \verb|F R|, for some type \verb|R| for which
-we have a function of type \verb|R -> Mu|.  The impredicative
-quantification of \verb|R| is essential here: we will instantiate it with
-\verb|Mu| itself in the definition of \verb|inMu|
-(Figure~\ref{fig:mu}).  So this approach would not work in a
-predicative theory like Agda's.  The quantification of \verb|R| can be
-seen as applying a technique due to Mendler, of introducing
-universally quantified variables for problematic type occurrences, to
-a datatype constructor.  We will review this in
-Section~\ref{sec:related}.
+Instead, we define \verb|Mu| as in Figure~\ref{fig:mu}, weakening
+\verb|Mu'| to a strictly positive approximation. Instead of taking
+\verb|F Mu|, the constructor \verb|mu| accepts an input of type
+\verb|F R|, for some type \verb|R| for which we have a function of type
+\verb|R -> Mu|.  The impredicative quantification of \verb|R| is
+essential here: we will instantiate it with \verb|Mu| itself in the
+definition of \verb|inMu| (Figure~\ref{fig:mu}).  Thus, this approach
+would not work in a predicative theory like Agda's.  The
+quantification of \verb|R| can be seen as applying a technique due to
+Mendler, of introducing universally quantified variables for
+problematic type occurrences, to a datatype constructor.  We will
+review this connection further in Section~\ref{sec:related}.
 
@@
 For a simple example: suppose we instantiate \verb|F| with
-\verb|ListF| (of Figure~\ref{fig:listf}).  Our derivation uses a
-different type that wraps \verb|F|, but this will show the issue in a
-simple form.  Let us temporarily define \verb|List A| as
+\verb|ListF| (from Section~\ref{sec:recu}).  Our derivation uses a
+different type that wraps \verb|F|, but using \verb|ListF|
+demonstrates a simple form of the underlying issue.  Let us temporarily define \verb|List A| as
 \verb|Mu (ListF A)| (again, for subsidiary recursion do not use just
 \verb|ListF| directly).  The canonical way to define the empty list
-would be, implicitly instantiating \verb|F| to \verb|ListF A|,
+would be like so:
 \begin{minted}{coq}
@@
 \end{minted}
-\noindent But given this, there are infinitely many other equivalent
-definitions.  For any \verb|Q : Set|, we could take
+\noindent Note that there are infinitely many other equivalent
+definitions of \verb|mkNil|.  For any arbitrary \verb|Q : Set|, for
+example, we could define
 \begin{minted}{coq}
@@
 But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
-definitionally nor provably.  One can define a function that puts
-\verb|Mu| values in normal form by folding \verb|inMu| over them.
-Then \verb|mkNil| and \verb|mkNil'| will have the same normal form,
-and be equivalent in that sense.  But the fact that they are not
-provably equal is what we term noncanonicity.
+definitionally nor provably.  Of course, one could then define a
+function that puts \verb|Mu| values in normal form by folding
+\verb|inMu| over them.  Then \verb|mkNil| and \verb|mkNil'| would be
+equivalent, in the sense that they have the same normal form, but they
+would still not be provably equal.
 
@@
 The type \verb|Subrec| is defined in Figure~\ref{fig:subrec}, as a
-fixed-point of \verb|SubrecF : Set -> Set|.  We take this fixed-point
-with \verb|Mu|, discussed in the previous section, and obtain \verb|roll|
+fixed-point of \verb|SubrecF : Set -> Set|.  We build this fixed-point
+using \verb|Mu| from the previous section, and obtain \verb|roll|
 and \verb|unroll| functions between \verb|SubrecF Subrec| and \verb|Subrec|.
-Unrolling \verb|Subrec| gives us the type
+Unrolling a \verb|Subrec| value gives us a term of type
 \begin{minted}{coq}
@@
 $\textit{Alg}\ X$ is $F\ X \to X$.  We elide the implementation of the
-\verb|roll| and \verb|unroll| functions, but note that \verb|unroll|
+\verb|roll| and \verb|unroll| functions, but we note that \verb|unroll|
 makes use of functoriality of carriers \verb|X|.
@@
 \begin{minted}{coq}
-Definition SubrecF(C : Set) :=
+Definition SubrecF(C : Set) : Set :=
   forall (X : Set -> Set) (FunX : Functor X), Alg X -> X C.
@@
 
+\begin{figure}
+\begin{minted}{coq}
+Definition fold : FoldT Alg Subrec :=
+  fun X FunX alg d => unroll d X FunX alg.
+
+Definition inn : F Subrec -> Subrec :=
+  fun d => roll (fun X xmap alg =>
+                 unrollAlg alg Subrec fold (fold X xmap alg) d).
+
+Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
+  fo F FunF (rollAlg (fun R' _ _ d => d)).
+\end{minted}
+\caption{The rest of the interface for \texttt{Subrec} (\texttt{Subrec.v})}
+\label{fig:subrecb}
+\end{figure}
+
 The rest of the interface for \verb|Subrec| is shown in
 Figure~\ref{fig:subrecb}.  To \verb|fold| an algebra
-\verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
+\verb|alg| with carrier \verb|X| (with the \verb|fmap| function given by
 \verb|FunX|) over \verb|d : Subrec|, we \verb|unroll| the
@@
 \item \verb|Subrec| is passed as the value for the abstract type \verb|R|; this is what enables all
-  the rest of the components to have the desired types, since we will pass values that have \verb|Subrec|
+  the rest of the components to have the desired types, since we will
+  pass values that have type \verb|Subrec|
   where the interface mentions \verb|R|.
@@
 recursion it has type \verb|F R'| where \verb|R'| is the abstract type
-of the subsidiary recursion.  So \verb|out| implements the idea that
+of the subsidiary recursion.  Intuitively, \verb|out| implements the idea that
 unfolding an abstract type one step is just a trivial case of
@@
 \begin{minted}{coq}
-Definition fold : FoldT Alg Subrec :=
-  fun X FunX alg d => unroll d X FunX alg.
-
-Definition inn : F Subrec -> Subrec :=
-  fun d => roll (fun X xmap alg =>
-                 unrollAlg alg Subrec fold (fold X xmap alg) d).
-
-Definition out{R:Set}(fo:FoldT Alg R) : R -> F R :=
-  fo F FunF (rollAlg (fun R' _ _ d => d)).
-\end{minted}
-\caption{The rest of the interface for \texttt{Subrec} (\texttt{Subrec.v})}
-\label{fig:subrecb}
+Definition kMo := Subrec -> Prop.
+Definition KAlgi := (kMo -> kMo) -> Set.
+Definition FoldTi(alg : KAlgi)(C : kMo) : kMo :=
+  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X),
+           alg X -> C d -> X C d.
+
+Definition AlgFi(A: KAlgi)(X : kMo -> kMo) : Set :=
+  forall (R : kMo)
+    (fo : (forall (d : Subrec), FoldTi A R d))
+    (ih : (forall (d : Subrec), R d -> X R d))
+    (d : Subrec),
+    Fi R d -> X R d.
+
+Definition Algi := MuAlgi Subrec AlgFi.
+
+Definition SubrecFi(C : kMo) : kMo :=
+  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), Algi X -> X C d.
+Definition Subreci := Mui Subrec SubrecFi.
+
+Definition foldi(i : Subrec) : FoldTi Algi Subreci i.
+Definition inni(i : Subrec)(fd : Fi Subreci i) : Subreci i.
+\end{minted}
+\caption{Interface for subsidiary induction (\texttt{Subreci.v})}
+\label{fig:subreci}
 \end{figure}
@@
 
-We have seen how to write subsidiary recursions in Coq.  But can one
-reason about these?  To wrap up this paper, we will briefly see the
-interface to our development of subsidiary induction in Coq, and
-example proofs written using this interface.  Subsidiary induction is
+We have now covered how to write subsidiary recursions in Coq.  But can one
+reason about these?  To wrap up our development, we briefly consider our
+interface for subsidiary induction in Coq, and present some
+example proofs that use this interface.  Subsidiary induction is
 the natural extension of subsidiary recursion, which worked over
@@
 \verb|Subrec|-indexed propositions (i.e., predicates).  Just as
-functors need an \verb|fmap| function, we here need an indexed
+functors need an \verb|fmap| function, here we need an indexed
 version, of type \verb|fmapiT Subrec Fi| (definition elided.)
@@
 type.  \verb|AlgFi| and \verb|Algi| are indexed versions
-of the algebras we saw for recursion.  The \verb|rec| function
-(Figure~\ref{fig:algf}) has now become an induction hypothesis: given
+of the algebras we saw for recursion.  The \verb|rec| function from
+Figure~\ref{fig:algf} is now an induction hypothesis: given
 any \verb|d| where \verb|R d| holds, \verb|ih| proves \verb|X R d|.  A
@@
 is the natural indexed version of \verb|SubrecF|.
-
-\begin{figure}
-\begin{minted}{coq}
-Definition kMo := Subrec -> Prop.
-Definition KAlgi := (kMo -> kMo) -> Set.
-Definition FoldTi(alg : KAlgi)(C : kMo) : kMo :=
-  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X),
-           alg X -> C d -> X C d.
-
-Definition AlgFi(A: KAlgi)(X : kMo -> kMo) : Set :=
-  forall (R : kMo)
-    (fo : (forall (d : Subrec), FoldTi A R d))
-    (ih : (forall (d : Subrec), R d -> X R d))
-    (d : Subrec),
-    Fi R d -> X R d.
-
-Definition Algi := MuAlgi Subrec AlgFi.
-
-Definition SubrecFi(C : kMo) : kMo :=
-  fun d => forall (X : kMo -> kMo) (xmap : fmapiT Subrec X), Algi X -> X C d.
-Definition Subreci := Mui Subrec SubrecFi.
-
-Definition foldi(i : Subrec) : FoldTi Algi Subreci i.
-Definition inni(i : Subrec)(fd : Fi Subreci i) : Subreci i.
-\end{minted}
-\caption{Interface for subsidiary induction (\texttt{Subreci.v})}
-\label{fig:subreci}
-\end{figure}
 
@@
 
-For proving the main theorem about run-length encoding, we need
-several lemmas about \verb|span|, shown in Figure~\ref{fig:spanlem}.
+To prove the main theorem about our run-length encoding, we need
+the three lemmas about \verb|span| shown in Figure~\ref{fig:spanlem}.
 For lack of space, we just state the properties.  The first says that
@@
 (module some conversions to \verb|list| from \verb|List|).  The second
-uses the inductive type \verb|Forall| from Coq's standard library to
+uses the inductive proposition \verb|Forall| from Coq's standard library to
 state that all the elements of the prefix returned by \verb|span|
@@
 subsidiarily about \verb|span|.  Using these lemmas, we can write a
-short proof by subsidiary induction of the following, where
+short proof by subsidiary induction of the following theorem, where
 \verb|rld : list (nat * A) -> list A| is the obvious decoding
@@
 well-foundedness then propagates through code.  In contrast, our
-approach here, while less general, does not clutter code with proofs.
+approach here, while less general, does not pollute computational code
+with proof terms.
 Subsidiary recursion can be seen as a generalization of \emph{nested
@@
 generalized to the form \verb|f (g x)|, where \verb|g| could be
-\verb|f| or another recursively defined function.  For more on
-partiality and recursion in theorem provers, see~\cite{bove+16}.
+\verb|f| or another recursively defined function.  Bove et al. have a
+more complete discussion of partiality and recursion in theorem provers~\cite{bove+16}.
 
@@
 \emph{strong functional programming}~\cite{turner95}.  Our method is
-similar to the technique of sized types, in providing a type-based
+similar to the technique of sized types, by providing a type-based
 method for termination~\cite{barthe04}.  With sized types, datatypes
@@
 subsidiary recursion, where recursions may be nested and outer
-recursive calls may be made on results of inner recursions.  We saw
+recursive calls may be made on the results of inner recursions.  We saw
 examples invoking the \verb|span| function as a subsidiary recursion,

### 2022-02-06 paper/paper.tex
applying some of Ben's edits by hand

--- 2022-02-03 paper/paper.tex
@@
   This paper describes a functor-generic derivation in Coq of
-  subsidiary recursion.  On this recursion scheme, inner recursions
+  subsidiary recursion.  With this recursion scheme, inner recursions
   may be initiated within outer ones, in such a way that outer
-  recursive calls may be made on the results from inner ones.  The
+  recursive calls may be made on results from inner ones.  The
   derivation utilizes a novel (necessarily weakened) form of
@@
 \end{minted}
-\noindent The sole recursive call is \verb|span p xs|. This call is
-structurally decreasing, since it occurs in a clause where the input
-list is of the form \verb|x:xs|.  In the appropriate syntax, this
+\noindent The sole recursive call is \verb|span p xs|, and it occurs
+in a clause where the input list is of the form \verb|x:xs|.  Hence it
+is structurally decreasing.  In the appropriate syntax, this
 definition can be accepted without additional effort by all the
-previously mentioned provers.
-
-This paper presents a more expressive form of terminating recursion,
+mentioned provers.
+
+This paper is about a more expressive form of terminating recursion,
 called \textbf{subsidiary recursion}.  While performing an outer
 recursion on some input \verb|x|, one may initiate an inner recursion
-on \verb|x| (or possibly some of its subdata), while still preserving
-the possibility of further invocations of the outer recursive
-function.  As a simple example, consider the function \verb|wordsBy|
-(\verb|Data.List.Extra|), which breaks a list into a set of its
-maximal sublists whose elements do not satisfy a predicate \verb|p|.
-For example, \verb|wordsBy isSpace " good day "| returns
-\verb|["good","day"]|. The code for this function is in
-Figure~\ref{fig:wordsBy}, where \verb|break p| is defined as
-\verb|span (not . p)|. The first recursive call, \verb|wordsBy p tl|,
-is structurally decreasing. In contrast, the second invocation of
-\verb|wordsBy p| uses a value obtained from a call to \verb|span|.
-This recursive call is not allowed under the structural termination
-condition, but is permitted by subsidiary recursion.
+on \verb|x| (or possibly some of its subdata), preserving the
+possibility of further invocations of the outer recursive function.
+Let us see a simple example.  The function \verb|wordsBy|
+(\verb|Data.List.Extra|) breaks a list into its maximal sublists whose
+elements do not satisfy a predicate \verb|p|.  For example,
+\verb|wordsBy isSpace " good day "| returns \verb|["good","day"]|.
+Code is in Figure~\ref{fig:wordsBy}.  Recall that
+\verb|break p| is equivalent to \verb|span (not . p)|. The first recursive call,
+\verb|wordsBy p tl|, is structural.  But in the second, we invoke
+\verb|wordsBy p| on a value obtained from another recursion, namely
+\verb|span|.  This is not allowed under structural termination, but
+will be permitted by subsidiary recursion.
 
@@
 other modifications to Coq, except the \verb|-impredicative-set|
-flag. Using this derivation, we present several example functions in
-the vein of \verb|wordsBy|, and prove theorems about their behavior.  A nice
-example is a definition of run-length encoding using \verb|span| as a
-subsidiary recursion, where we prove that encoding and then decoding
-returns the original list.  Our approach applies to the standard
-datatypes in the Coq library, and does not require switching libraries
-or datatype definitions.
-
-An important technical novelty of our development is a derivation of a
-weakened form of positive-recursive type in Coq.  Coq (Agda, and Lean)
-restrict datatypes $D$ to be strictly positive: in the argument types
-of constructors of $D$, $D$ cannot occur to the left of any arrows.
-Our derivation needs to use positive-recursive types, where $D$ may
-also occur to the left of an even number (only) of arrows.  We present
-a way to derive a weakened form of positive-recursive type that is
-sufficient for our examples (Section~\ref{sec:mu}).  The weakening is
-to require only that $F\ (\mu F)$ is a retract of $\mu F$.  (Usually
-these types are isomorphic.)  Hence, we dub these
-\textbf{retractive-positive} recursive types.  This weakening leads to
-noncanonical elements of $\mu$, but we will see how to work around
-this.  Our definition of retractive-positive recursive types makes
-essential use of impredicative quantification, and hence is not legal
-in predicative theories like Agda's.
+flag. Using this derivation, we present several example functions like
+\verb|wordsBy|, and prove theorems about them.  A nice example is a
+definition of run-length encoding using \verb|span| as a subsidiary
+recursion, where we prove that encoding and then decoding returns the
+original list.  Our approach applies to the standard datatypes in the
+Coq library, and does not require switching libraries or datatype
+definitions.
+
+An important technical novelty is a derivation of a weakened form of
+positive-recursive type in Coq.  Coq (Agda, and Lean) restrict
+datatypes $D$ to be strictly positive: in the input types of
+constructors of $D$, $D$ cannot occur to the left of any arrows.  Our
+derivation needs to use positive-recursive types, where $D$ may occur
+to the left of an even number (only) of arrows.  We present a way to
+derive a weakened form of positive-recursive type that is sufficient
+for our examples (Section~\ref{sec:mu}).  The weakening is to require
+only that $F\ (\mu F)$ is a retract of $\mu F$.  Usually, these types are
+isomorphic.  Hence, we dub these \textbf{retractive-positive}
+recursive types.  This weakening leads to noncanonical elements of
+$\mu$, but we will see how to work around this.  Our definition of
+retractive-positive recursive types makes essential use of
+impredicative quantification, and hence is not legal in predicative
+theories like Agda's.
 
 We begin by summarizing the interface our derivation provides for
-subsidiary recursion (Section~\ref{sec:interface}), and we then
-present some examples (Section~\ref{sec:examples}).  We next explain
-the implementation of our interface (Section~\ref{sec:deriv}),
-including our definition of retractive-positive recursive types
-% (Section~\ref{sec:mu})
-.  We then cover the interface for subsidiary
-induction (Section~\ref{sec:interfacei}), and give some example proofs
-that use it (Section~\ref{sec:examplesi}). We conclude with a discussion
-of related work (Section~\ref{sec:related}).
+subsidiary recursion (Section~\ref{sec:interface}), and then see
+examples (Section~\ref{sec:examples}).  We next explain how the
+interface is actually implemented (Section~\ref{sec:deriv}), including
+our retractive-positive recursive types (Section~\ref{sec:mu}).  The
+interface for subsidiary induction is covered next
+(Section~\ref{sec:interfacei}), and example proofs using it
+(Section~\ref{sec:examplesi}).  Related work is discussed in
+Section~\ref{sec:related}.
 
@@
 algebra and category theory to describe inductive datatypes and their
-recursion principles (cf.~\cite{traytel12,cockett92,hagino87}). Under
-this approach, inductive datatypes are equipped with a signature
-functor which explains how to 'unroll' one layer of the type. The
-signature functor for lists is:
-\begin{minted}{coq}
-Inductive ListF (X : Set) : Set :=
+recursion principles (cf.~\cite{traytel12,cockett92,hagino87}).  With
+this approach, one describes transformations to be performed on data
+as \emph{algebras}, which can then be \textit{folded} over data.  The
+simplest form of algebras, namely $F$-algebras for functor $F$ (called
+the \emph{signature functor} of the datatype), are morphisms from
+$F\ A$ to $A$, for carrier object $A$.  From a programming
+perspective, an $F$-algebra is given input of type $F\ A$, and must
+compute a result of type $A$.  An example of $F$ is the signature
+functor for lists, which we will use below:
+
+\begin{minted}{coq}
+Inductive ListF(X : Set) : Set :=
 | Nil : ListF X
@@
 \end{minted}
-Transformations on data with functor $F$ are described as
-\emph{algebras} which are then \textit{folded} over data.  The
-simplest form of algebras are morphisms from $F\ A$ to $A$, for some
-carrier object $A$.  From a programming perspective, an $F$-algebra is
-given input of type $F\ A$, and must compute a result of type $A$. The
-$ListF$-algebra for computing the length of a list is:
-\begin{minted}{coq}
-Definition LengthAlg (l : ListF nat) : nat :=
-match l with
-| Nil => 0
-| Cons a len' => 1 + len'
-end.
-\end{minted}
+
 Algebras for our subsidiary recursion are more complex than
@@
 we will explain further below, the carrier of the algebra will be a
-functor \verb|X : Set -> Set|.  Our algebras are presented with:
+functor \verb|X : Set -> Set|.  The algebra is presented with:
 
@@
 
-We use Coq inductive types for the signature functors \verb|F| of
+We will use Coq inductive types for the signature functors \verb|F| of
 various datatypes. This allows algebras to use Coq's
-pattern-matching on the subdata structure \verb|d|, enabling a style
-of coding that retains a similar feel to structural recursion. Unlike
-structurally recursive functions, however, this interface uses a
-type-based (and hence compositional) criteria to ensure termination.
+pattern-matching on the subdata structure \verb|d|.  So the style of
+coding against this interface retains a similar feel to structural
+recursion.  Unlike with structural termination, though, the interface
+here is type-based and hence compositional.
 
@@
 ``universe'', we have a kind of space (here, \verb|R|), which one
-cannot escape using the provided operations.  Other examples include the
-ordinal $\epsilon_0$ and operator $\omega^-$, and the physical
-universe and traveling at the speed of light.  Staying in the
-recursion universe is good, because we may recurse (via \verb|rec|) on
-any value of type \verb|R|.  Some points must still be explained: in
-particular why \verb|X| has type \verb|Set -> Set|, and the definition
-of \verb|FoldT|. We consider those details next.
+cannot escape using certain operations.  Other examples are the
+ordinal $\epsilon_0$ and $\omega^-$, and the physical universe and
+traveling at the speed of light.  Staying in the recursion universe is
+good, because we may recurse (via \verb|rec|) on any value of type
+\verb|R|.  Some points must still be explained: why \verb|X| has type
+\verb|Set -> Set|, and the definition of \verb|FoldT|.  Let us see
+these details next.
 
@@
 \end{minted}
-  \caption{Some basics from \texttt{List.v}, specializing the
-    functor-generic derivation of subsidiary recursion to
-    lists parameterized by an element type \texttt{A}  (\texttt{List.v})}
+  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists parametrized by an element type \texttt{A} (\texttt{List.v})}
   \label{fig:listf}
@@
 
-The type over which one can recurse using our subsidiary
-recursion scheme is called \verb|Subrec|.  It is parametrized by a signature
-functor \verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
+The type over which one can recurse using our scheme of subsidiary recursion is
+called \verb|Subrec|.  It is parametrized by a signature functor
+\verb|F| of type \verb|Set -> Set|.  \verb|Subrec| comes with
 \verb|inn : F Subrec -> Subrec|, which behaves computationally like a
-constructor.  We will later derive an induction principle for this type
-(Section~\ref{sec:interfacei}). The definition of \verb|Subrec| uses
-retractive-positive recursive types, to take a fixed-point of a
+constructor.  We will later derive an induction principle for this
+type (Section~\ref{sec:interfacei}). The definition of \verb|Subrec|
+uses retractive-positive recursive types, to take a fixed-point of a
 construction based on \verb|F|.  We present these recursive types in
@@
 
-Our examples specialize this scheme to the case of polymorphic lists,
-presented in Figure~\ref{fig:listf}.  In general, to use our
-development to get subsidiary recursion over some datatype, one must
-define a signature functor for the datatype, e.g., \verb|ListF| from
-the start of this section. We define \verb|List| to be
-\verb|Subrec| with \verb|F| instantiated to \verb|ListF A|.  This
-type \verb|List| is not to be confused with the type \verb|list| of
-lists from Coq's standard library.  As noted previously, our development
-is meant to be used in extension of existing inductive datatypes, not
+In this paper, we use just the specialization to the case of lists,
+with signature functor \verb|ListF A| shown above.  The parameter
+\verb|A| is the type for the list elements.  \verb|List| is then
+defined to be \verb|Subrec|, with \verb|F| instantiated to
+\verb|ListF A|. In general, to use our development to get subsidiary
+recursion over some datatype, one must define a signature functor for
+the datatype.  Note that \verb|List| is different from the type
+\verb|list| of lists in Coq's standard library.  Our development is
+meant to be used in extension of existing inductive datatypes, not
 replacing them.  The figure also shows constructors \verb|mkNil| and
 \verb|mkCons| for \verb|List|, and typings for conversion functions
-between \verb|List| and \verb|list| (we elide their definitions).
+between \verb|List| and \verb|list| (definitions elided).
 
@@
 
-We need a fixed-point here due to the \verb|fold| parameter
-\verb|AlgF| expecting an argument of type \verb|FoldT Alg R|.  This is the type for fold functions which
-apply algebras (\verb|Alg|) to data of type \verb|R|.  The variable \verb|Alg| occurs only positively (but not
-strictly positively) in \verb|AlgF|, because it occurs negatively in
-\verb|FoldT Alg R| which occurs negatively in \verb|AlgF Alg X|.  So
-we can indeed take a fixed-point of \verb|AlgF| to define the constant
-\verb|Alg|.
+We need a fixed-point here because \verb|Alg| occurs in the definition
+of \verb|Algf|.  This is an essential circularity, because we are
+trying to express that algebras take in \verb|fold| functions, which
+themselves may accept algebras.  The variable \verb|Alg| occurs
+negatively in \verb|FoldT Alg R| which occurs negatively in
+\verb|AlgF Alg X|.  Hence it occurs positively but not strictly
+positively.  So we can indeed take a fixed-point of \verb|AlgF| to
+define the constant \verb|Alg|.
 
@@
 \verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
-nothing about \verb|R| except that it supports the
-\verb|fold| and \verb|rec| operations that follow. The first is a local
-\verb|fold| function, which allows us to fold another algebra over data
+nothing about \verb|R| except that it supports the following
+operations.  We have a local
+\texttt{fold} function, which will allow us to fold another algebra over data
 of type \verb|R|.  We will use \verb|fold| to initiate subsidiary
-recursions.  The second, \verb|rec|, is used to make recursive calls
-on values of type \verb|R|.
+recursions.  Then there is \verb|rec|, for recursive calls on data of
+type \verb|R|.
 
@@
 \verb|FoldT Alg C|, (i) recursive calls may compute a result of type
-\verb|X R|, mentioning the abstract type \verb|R| inside the recursion;
+\verb|X R|, mentioning the abstract type \verb|R| for that recursion;
 and (ii) outside that recursion, the result will have type \verb|X C|.
@@
 \verb|R|, but view those results as having the type \verb|C| outside
-the recursion.  The function \verb|fold| in Figure~\ref{fig:algf} initiates
-top-level folds.  We can also have functions
-between \verb|Alg| and its \verb|AlgF|-unfolding.  We will
+the recursion.  The function \verb|fold| (Figure~\ref{fig:algf}) initiates
+top-level folds.  We also can have functions 
+between \verb|Alg| and its \verb|Algf|-unfolding.  We will
 return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.
@@
 fold function for that type (i.e., of type
-\verb|FoldT (Alg (ListF A)) R|), we would like to map an input list of
+\verb|FoldT (Alg (ListF A)) R|), we will map an input list of
 type \verb|R| to a pair of type \verb|list A * R|.  The first
@@
 cannot have type \verb|R|; we cannot statically ensure that outer
-recursions on it are legal.  But the second component will be a
+recursions on it are legal.  But the second component will always be a
 subdatum of the input list, and so can still have type \verb|R|,
@@
 \end{minted}
-\noindent The idea is that the recursion will signal if it is in the
-one tricky case where \verb|p| does not match the first element, by
-returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
+\noindent The idea is that the recursion will return \verb|SpanNoMatch| to signal that it is in the
+one tricky case where \verb|p| does not match the first element.  Otherwise, it will be able to return,
 via \verb|SpanSomeMatch|, a prefix and the suffix at type
@@
 
-Figure~\ref{fig:spanalg} show the algebra
+Figure~\ref{fig:spanalg} shows the algebra
 \verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
@@
 suffix in the pair we return (via \verb|SpanSomeMatch|).  Happily, we
-have \verb|tl : R| here, so we can do this.  In either case (for
-return value of \verb|span tl|), we add the head to the front of the
+have \verb|tl : R| here, so we can do this.  For either possible return
+value of \verb|span tl|, we add the head to the front of the
 prefix.
@@
 
-Let us now see how to write \verb|wordsBy|, our example function from
+We now consider how to write the \verb|wordsBy| function from
 Section~\ref{sec:intro}, using \verb|breakr| subsidiarily.  The code
@@
 setup is similar to that for \verb|span|.  We first define an algebra
-\verb|WordsBy|, parametrized by the predicate \verb|p|, of type
-\verb|Alg (ListF A) (Const (list (list A)))|.  This says that
-\verb|WordsBy p| is an algebra (\verb|Alg|) for the \verb|ListF A|
+\verb|WordsByAlg| of type
+\verb|Alg (ListF A) (Const (list (list A)))|, parametrized
+by a predicate \verb|p|.  This type expresses that
+\verb|WordsByAlg p| is an algebra (\verb|Alg|) for the \verb|ListF A|
 functor, with carrier \verb|Const (list (list A))|.  \verb|Const| is
@@
 
-The code for \verb|WordsBy| is essentially
-the same as what we saw in Section~\ref{sec:intro}.  We pattern match
-on \verb|xs : ListF A R|.  Recall that for this function, we are
-trying to drop elements which satisfy \verb|p|, and return a list of
-the sublists between maximal sequences of such elements.  In the
-\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we are
-supposed to drop it and recurse.  This is legal, because \verb|tl : R|
-and \verb|wordsBy : R -> list (list A)|.  In the \verb|else| case,
+The code for \verb|WordsByAlg| is essentially
+the same as what we saw in Section~\ref{sec:intro}.  Recall that this function drops elements
+that satisfy \verb|p|, and returns the list of
+sublists between maximal sequences of such elements.  The algebra
+pattern-matches on \verb|xs : ListF A R|. In the
+\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we
+drop it and recurse.  Legality of the recursive call follows by typing: \verb|tl : R|
+has the type expected by \verb|wordsBy : R -> list (list A)|.  Otherwise,
 we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
@@
 with \verb|ListF A|).  This means that from the invocation of
-\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  And so we
-can indeed apply \verb|wordsBy : R -> list (list A)| to
+\verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  Thus, it is
+legal to apply \verb|wordsBy : R -> list (list A)| to
 \verb|z| to recurse.  The figure also shows the code for the subsidiary
@@
 \begin{minted}{coq}
-  Definition WordsBy(p : A -> bool)
+  Definition WordsByAlg(p : A -> bool)
     : Alg (ListF A) (Const (list (list A))) :=
@@
 
-In this example, we see how to write a combinator that factors out a
-subsidiary recursion.  The Haskell library \verb|Data.List.Extra| has
-a function \verb|repeatedly|, defined essentially as in
-Figure~\ref{fig:mapthroughhs}, though we attempt a more informative
-name.  This is like the standard \verb|map| function on lists, except
-that the function \verb|f| that we are mapping (or ``mapping
-through'') takes in not just the current element \verb|a|, but also
-the tail \verb|as|.  It then returns the value \verb|b| to include in
-the output list, and whatever other list it wishes, upon which
-\verb|mapThrough| will recurse.
+This example shows how to write a combinator that factors out a
+subsidiary recursion.  The Haskell library \verb|Data.List.Extra|
+defines a function \verb|repeatedly| in essentially the same way as
+\verb|mapThrough| in Figure~\ref{fig:mapthroughhs} (we propose a more
+informative name).  This function behaves like the standard \verb|map|
+function on lists, except that the function \verb|f| that we are
+mapping (or ``mapping through'') takes in not just the current element
+\verb|a|, but also the tail \verb|as|.  It then returns the value
+\verb|b| to include in the output list, and whatever other list it
+wishes, upon which \verb|mapThrough| will then recurse.
 
@@
 mapped function can initiate a subsidiary recursion, returning a value
-in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  So
-the type we will use for mapped functions is:
+in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  
+The type we will use for mapped functions is:
 \begin{minted}{coq}
@@
 
-Given this definition, the code is in Figure~\ref{fig:mapthrough}.
+Given this definition, the Coq definition of \verb|mapThrough| is shown in Figure~\ref{fig:mapthrough}.  
 \verb|MapThroughAlg| is similar to the
 Haskell code above, though when we call \verb|f|, we must
-supply the abstract type \verb|R| and fold function \verb|fo|.  Then,
-from the definition of \verb|mappedT|, we have that \verb|b : B| and
-\verb|c : R|.  So we may indeed invoke \verb|mapThrough : R -> list B|
+supply the abstract type \verb|R| and fold function \verb|fo|.  
+From the definition of \verb|mappedT|, we have that \verb|b : B| and
+\verb|c : R|, so we may indeed invoke \verb|mapThrough : R -> list B|
 on \verb|c|.  Note that as we are building up a new list from scratch
@@
 \end{minted}
-\caption{The algebra \texttt{MapThroughAlg} defining function \texttt{mapThrough} and \texttt{mapThroughr}; the
-  code for those follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} (Figure~\ref{fig:wordsby}), so
-  we omit it (\texttt{MapThrough.v})}
+\caption{The algebra \texttt{MapThroughAlg} defining the functions \texttt{mapThrough} and \texttt{mapThroughr}; 
+  code for the latter is omitted, as it follows the pattern of \texttt{wordsBy} and \texttt{wordsByr} of Figure~\ref{fig:wordsby} (\texttt{MapThrough.v})}
 \label{fig:mapthrough}
@@
 sequences of $n$ occurrences of element $e$ are summarized by the pair
-$(n,e)$~\cite{datacomp}.  Haskell code is in Figure~\ref{fig:rlehs}.
+$(n,e)$~\cite{datacomp}.  A Haskell implementation of this algorithm
+is in Figure~\ref{fig:rlehs}.
 Recall that \verb|(== a)| tests its input for equality with \verb|a|.
@@
 The function \verb|compressSpan| is written at the type \verb|mappedT A (nat * A)| that
-will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|,
-\verb|compressSpan| has type:
+will be required by \verb|mapThrough|.  Unfolding the definition of \verb|mappedT|, we see that
+\verb|compressSpan| has this type:
 \begin{minted}{coq}
@@
 \end{minted}
-\noindent It is invoked by the code for \verb|mapThrough| with \verb|fo : FoldT (Alg (ListF A)) R|.  It then has the
-responsibility of extracting from the tail at type \verb|R| (second input)
-a result upon which \verb|mapThrough| should recurse (second component
+\noindent It is invoked by the code for \verb|mapThrough| with \verb|fo : FoldT (Alg (ListF A)) R|.  Then
+\verb|compressSpan| will extract from the tail at type \verb|R| (second input) 
+the suffix upon which \verb|mapThrough| should recurse (second component
 of the output pair).  Then we define an algebra \verb|RleAlg| by
@@
 theory.  Here we exploit Coq's
-impredicative polymorphism for a different solution.
-
-Assume \verb|F : Set -> Set|, with an \verb|fmap| function
+impredicative polymorphism to obtain a different solution.
+
+Our starting point is a type scheme \verb|F : Set -> Set|, with an \verb|fmap| function
 (morphism part of the functor) of type
@@
 \end{minted}
-\noindent This is exactly what is used in approaches to modular
+\noindent This is exactly what is used in many approaches to modular
 datatypes in functional programming, like
@@
 
-The definition of \verb|Mu| in Figure~\ref{fig:mu} weakens
-this to a strictly positive approximation. Instead of taking in \verb|F Mu|, constructor \verb|mu|
-accepts an input of type \verb|F R|, for some type \verb|R| for which
-we have a function of type \verb|R -> Mu|.  The impredicative
-quantification of \verb|R| is essential here: we will instantiate it with
-\verb|Mu| itself in the definition of \verb|inMu|
-(Figure~\ref{fig:mu}).  So this approach would not work in a
-predicative theory like Agda's.  The quantification of \verb|R| can be
-seen as applying a technique due to Mendler, of introducing
-universally quantified variables for problematic type occurrences, to
-a datatype constructor.  We will review this in
-Section~\ref{sec:related}.
+Instead, we define \verb|Mu| in Figure~\ref{fig:mu}, to weaken this
+ideal \verb|Mu'| to a strictly positive approximation. Instead of taking in
+\verb|F Mu|, the constructor \verb|mu| accepts an input of type
+\verb|F R|, for some type \verb|R| for which we have a function of
+type \verb|R -> Mu|.  The impredicative quantification of \verb|R| is
+essential here: we will instantiate it with \verb|Mu| itself in the
+definition of \verb|inMu| (Figure~\ref{fig:mu}).  So this approach
+would not work in a predicative theory like Agda's.  The
+quantification of \verb|R| can be seen as applying a technique due to
+Mendler, of introducing universally quantified variables for
+problematic type occurrences, to a datatype constructor.  We will
+review this in Section~\ref{sec:related}.
 
@@
 For a simple example: suppose we instantiate \verb|F| with
-\verb|ListF| (of Figure~\ref{fig:listf}).  Our derivation uses a
-different type that wraps \verb|F|, but this will show the issue in a
-simple form.  Let us temporarily define \verb|List A| as
-\verb|Mu (ListF A)| (again, for subsidiary recursion do not use just
-\verb|ListF| directly).  The canonical way to define the empty list
-would be, implicitly instantiating \verb|F| to \verb|ListF A|,
+\verb|ListF A| (from Section~\ref{sec:recu}).  Our derivation uses a
+different type that wraps \verb|F|, but using \verb|ListF A|
+demonstrates the issue in a simple form.  Let us temporarily define
+\verb|List A| as \verb|Mu (ListF A)| (again, for subsidiary recursion
+do not use just \verb|ListF| directly).  The canonical way to define
+the empty list would be:
 \begin{minted}{coq}
@@
 But critically, \verb|mkNil| and \verb|mkNil'| are not equal, neither
-definitionally nor provably.  One can define a function that puts
-\verb|Mu| values in normal form by folding \verb|inMu| over them.
-Then \verb|mkNil| and \verb|mkNil'| will have the same normal form,
-and be equivalent in that sense.  But the fact that they are not
-provably equal is what we term noncanonicity.
-
-Noncanonicity must be handled carefully when reasoning about functions
-defined with our interface.  We will see an example in
+definitionally nor provably.  Of course, one could define a function
+that puts \verb|Mu| values in canonical form by folding \verb|inMu|
+over them.  Then \verb|mkNil| and \verb|mkNil'| would be equivalent.
+But they would still not be provably equal, which is the problem of
+noncanonicity.  We will see how to work around this in
 Section~\ref{sec:examplesi}.  First, though, let us complete the
@@
 The type \verb|Subrec| is defined in Figure~\ref{fig:subrec}, as a
-fixed-point of \verb|SubrecF : Set -> Set|.  We take this fixed-point
-with \verb|Mu|, discussed in the previous section, and obtain \verb|roll|
+fixed-point of \verb|SubrecF : Set -> Set|.  We build this fixed-point
+using \verb|Mu| from the previous section, and obtain \verb|roll|
 and \verb|unroll| functions between \verb|SubrecF Subrec| and \verb|Subrec|.
-Unrolling \verb|Subrec| gives us the type
+Unrolling a \verb|Subrec| term gives us a term of type
 \begin{minted}{coq}
@@
 We have seen how to write subsidiary recursions in Coq.  But can one
-reason about these?  To wrap up this paper, we will briefly see the
+reason about these?  We turn now briefly to the
 interface to our development of subsidiary induction in Coq, and
-example proofs written using this interface.  Subsidiary induction is
+some example proofs written using this interface.  Subsidiary induction is
 the natural extension of subsidiary recursion, which worked over
@@
 of the algebras we saw for recursion.  The \verb|rec| function
-(Figure~\ref{fig:algf}) has now become an induction hypothesis: given
+from Figure~\ref{fig:algf} is now an induction hypothesis: given
 any \verb|d| where \verb|R d| holds, \verb|ih| proves \verb|X R d|.  A
@@
 
-For proving the main theorem about run-length encoding, we need
-several lemmas about \verb|span|, shown in Figure~\ref{fig:spanlem}.
+To prove the main theorem about run-length encoding, we need
+the three lemmas about \verb|span| shown in Figure~\ref{fig:spanlem}.
 For lack of space, we just state the properties.  The first says that
@@
 generalized to the form \verb|f (g x)|, where \verb|g| could be
-\verb|f| or another recursively defined function.  For more on
-partiality and recursion in theorem provers, see~\cite{bove+16}.
+\verb|f| or another recursively defined function.  See the survey by
+Bove et al. for more on partiality and recursion in theorem
+provers~\cite{bove+16}.
 
@@
 generalizing the carriers of algebras to functors is the critical step
-enabling such examples. In our Coq development, we show that Uustalu
-and Vene's scheme is derivable from subsidiary recursion.
+enabling such examples.
 

### 2022-02-06 paper/paper.tex
more small edits

@@
 negatively in \verb|FoldT Alg R| which occurs negatively in
-\verb|AlgF Alg X|.  Hence it occurs positively but not strictly
+\verb|AlgF Alg X|.  Hence it occurs positively in \verb|Algf|, though not strictly
 positively.  So we can indeed take a fixed-point of \verb|AlgF| to
@@
 \end{minted}
-\caption{Functions  \texttt{wordsBy} and \texttt{wordsByr}, and the algebra they
-  fold (\texttt{WordsBy.v})}
+\caption{The algebra \texttt{WordsByAlg}, and functions \texttt{wordsBy} and \texttt{wordsByr} folding it (\texttt{WordsBy.v})}
 \label{fig:wordsby}
@@
 To write this combinator using our infrastructure for subsidiary
-recursion, we need to supply the mapped function with the fold
-function for \verb|mapThrough|'s recursion.  This is so that the
-mapped function can initiate a subsidiary recursion, returning a value
-in the abstract type \verb|R| of \verb|mapThrough|'s recursion.  
-The type we will use for mapped functions is:
+recursion, we will use this type for mapped functions:
 \begin{minted}{coq}
@@
 \end{minted}
-\noindent This type is more informative than the Haskell type,
-since it shows that the second component of the returned value
-must have type \verb|R|, and hence must be (hereditarily) a tail
-of the input.
+\noindent This \verb|mappedT| type is more informative than the
+Haskell type, since it shows that the second component of the returned
+value must have type \verb|R|, and hence must be (hereditarily) a tail
+of the input.  We need to supply mapped functions with the fold
+function to use, which will come from \verb|mapThrough|'s recursion.
+Mapped functions need this to initiate subsidiary recursions,
+returning a value in the abstract type \verb|R| of \verb|mapThrough|'s
+recursion.
 

### 2022-02-07 paper/paper.tex
final edits

@@
 \verb|span|.  This is not allowed under structural termination, but
-will be permitted by subsidiary recursion.
+is permitted by subsidiary recursion.
 
@@
 recursion, where we prove that encoding and then decoding returns the
-original list.  Our approach applies to the standard datatypes in the
-Coq library, and does not require switching libraries or datatype
-definitions.
+original list.  Our approach does not require switching libraries or
+datatype definitions.
 
@@
 compute a result of type $A$.  An example of $F$ is the signature
-functor for lists, which we will use below:
+functor for lists, paramatrized by the type \verb|A| of elements:
 
@@
 
-We have previously dubbed this interface a \emph{recursion
-universe}~\cite{stump20}.  As in other domains using the term
+We have previously dubbed interfaces for recursion \emph{recursion
+universes}~\cite{stump20}.  As in other domains using the term
 ``universe'', we have a kind of space (here, \verb|R|), which one
@@
 
-In this paper, we use just the specialization to the case of lists,
-with signature functor \verb|ListF A| shown above.  The parameter
-\verb|A| is the type for the list elements.  \verb|List| is then
-defined to be \verb|Subrec|, with \verb|F| instantiated to
-\verb|ListF A|. In general, to use our development to get subsidiary
+In this paper, our examples use \verb|ListF A| (shown above) to
+instantiate the parameter \verb|F|.  \verb|List| is then defined to be
+\verb|Subrec|. In general, to use our development to get subsidiary
 recursion over some datatype, one must define a signature functor for
@@
 
-\subsection{Algebras for subsidiary recursion}
-
-\verb|Subrec.v| also implements the notion of algebra we introduced
-informally above.  The central definitions are in
-Figure~\ref{fig:algf}.  \verb|KAlg| is the kind for the
-type-constructor for algebras, as we see in the definition of
-\verb|Alg|.  This type-constructor \verb|Alg| is a fixed-point of the
-type \verb|AlgF|.  The fixed-point is taken using \verb|MuAlg|, which
-implements our retractive-positive recursive types
-(Section~\ref{sec:mu}) at kind \verb|KAlg|.
-
-We need a fixed-point here because \verb|Alg| occurs in the definition
+\subsection{Algebras for subsidiary recursion (\texttt{Subrec.v})}
+
+Let us now look more formally at the notion of algebra we introduced
+informally above.  The central definitions are in Figure~\ref{fig:algf}.
+\verb|KAlg| is the kind for the type-constructor for algebras, as we
+see in the definition of \verb|Alg|.  This type-constructor \verb|Alg|
+is a fixed-point of the type \verb|AlgF|.  The fixed-point is taken
+using \verb|MuAlg|, which implements our retractive-positive recursive
+types (Section~\ref{sec:mu}) at kind \verb|KAlg|.
+
+We need a fixed-point here since \verb|Alg| occurs in the definition
 of \verb|Algf|.  This is an essential circularity, because we are
@@
 
-Let us look at \verb|AlgF|.  As noted already, each recursion is based
+Let us look further at \verb|AlgF|.  As noted already, each recursion is based
 on an abstract type \verb|R|, representing the data upon which we will
@@
 \verb|AlgF Alg X|.  Reasoning parametrically, an algebra can assume
-nothing about \verb|R| except that it supports the following
-operations.  We have a local
+nothing about \verb|R| except that it supports the 
+operations that follow.  We have a local
 \texttt{fold} function, which will allow us to fold another algebra over data
@@
 recursions.  Then there is \verb|rec|, for recursive calls on data of
-type \verb|R|.
+type \verb|R|.  Finally, we have the subdata structure \verb|d : F R|.
 
@@
 flexibility to type results inside a recursion with the abstract type
-\verb|R|, but view those results as having the type \verb|C| outside
+\verb|R|, but view those results with type \verb|C| outside
 the recursion.  The function \verb|fold| (Figure~\ref{fig:algf}) initiates
-top-level folds.  We also can have functions 
+top-level folds.  We also have functions \verb|rollAlg| and \verb|unrollAlg|
 between \verb|Alg| and its \verb|Algf|-unfolding.  We will
-return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.
+fill in missing definitions (for \verb|Subrec|, \verb|inn|, \verb|fold|, etc.) in Section~\ref{sec:deriv}.
 
@@
 for the \verb|rec| function; and \verb|d| for the subdata structure.
+(\verb|FmapId| expresses the identity law for functors, needed
+for the proof.)
 
@@
 
-This first example does not invoke subsidiary recursions, but will
-itself be used as a subsidiary recursion in other examples to follow.
-Given a predicate \verb|p : A -> bool|, and a value of type
-\verb|List A|, we would like to compute a pair of type
-\verb|list A * List A|, where the first component is the maximal
-prefix whose elements satisfy \verb|p|, and the second is the
-remaining suffix.  This is the typing for a top-level recursion.  More
-generally, though, given a type \verb|R : Set| along with a
-fold function for that type (i.e., of type
-\verb|FoldT (Alg (ListF A)) R|), we will map an input list of
-type \verb|R| to a pair of type \verb|list A * R|.  The first
-component of this pair is going to be built up from scratch, and so
-cannot have type \verb|R|; we cannot statically ensure that outer
-recursions on it are legal.  But the second component will always be a
-subdatum of the input list, and so can still have type \verb|R|,
-enabling outer recursive calls.  So we want:
+Our first example is \verb|span| discussed above.  It does not invoke
+subsidiary recursions, but will be used as a subsidiary
+recursion in later examples.  Given a predicate
+\verb|p : A -> bool|, and a value of type \verb|List A|, we would like
+to compute a pair of type \verb|list A * List A|, where the first
+component is the maximal prefix whose elements satisfy \verb|p|, and
+the second is the remaining suffix.  This is the typing for a
+top-level recursion.  More generally, though, given a type
+\verb|R : Set| along with a fold function for that type (i.e., of type
+\verb|FoldT (Alg (ListF A)) R|), we will map an input list of type
+\verb|R| to a pair of type \verb|list A * R|.  The first component of
+this pair is going to be built up from scratch, and so cannot have
+type \verb|R|; we cannot statically ensure that outer recursions on it
+would be legal.  But the second component will always be a subdatum of the
+input list, and so can still have type \verb|R|.  The typing is:
 \begin{minted}{coq}
@@
 as desired.  To work around this, we will have our recursion return a value
-of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
+of the following type \verb|SpanF R| (with \verb|X| implicit for the constructors):
 \begin{minted}{coq}
@@
 Figure~\ref{fig:spanalg} shows the algebra
-\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  So we are defining
+\verb|SpanAlg|, whose type is \verb|Alg (ListF A) SpanF|.  We are defining
 an algebra (\verb|Alg|) for the \verb|ListF A| functor, with carrier
@@
 recursion universe: the abstract type \verb|R|, the fold function
-(\verb|fo|) for any subsidiary recursions (not needed here), a
+\verb|fo| for any subsidiary recursions (not needed here), a
 function we choose to name \verb|span| for making recursive calls, and
@@
 \verb|SpanSomeMatch l r|, then we return the nonempty
-prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| 
-for subsidiary recursion (e.g., in \verb|wordsBy|, below).
+prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version \verb|breakr| 
+for subsidiary recursion.
 
@@
 
-The code for \verb|WordsByAlg| is essentially
-the same as what we saw in Section~\ref{sec:intro}.  Recall that this function drops elements
-that satisfy \verb|p|, and returns the list of
-sublists between maximal sequences of such elements.  The algebra
-pattern-matches on \verb|xs : ListF A R|. In the
-\verb|Cons| case, if the head (\verb|hd|) satisfies the predicate, then we
-drop it and recurse.  Legality of the recursive call follows by typing: \verb|tl : R|
-has the type expected by \verb|wordsBy : R -> list (list A)|.  Otherwise,
-we use \verb|breakr| to obtain the maximal prefix \verb|w| of \verb|tl| that
-does not satisfy \verb|p|, and the remaining suffix \verb|z|.
+The code for \verb|WordsByAlg| is essentially the same as
+\verb|wordsBy| in Haskell, which we saw in Section~\ref{sec:intro}.
+Recall that we must drop elements that satisfy \verb|p|, and return
+the list of sublists between maximal sequences of such elements.  The
+algebra pattern-matches on \verb|xs : ListF A R|. In the \verb|Cons|
+case, if the head (\verb|hd|) satisfies the predicate, then we drop it
+and recurse.  Legality of the recursive call follows by typing, since
+\verb|tl : R| has the input type required by
+\verb|wordsBy : R -> list (list A)|.  Otherwise, we use \verb|breakr|
+to obtain the maximal prefix \verb|w| of \verb|tl| that does not
+satisfy \verb|p|, and the remaining suffix \verb|z|.
 
@@
 comes from the type \verb|FoldT (ListF A) Alg R| of
-\verb|fo|, from the definition
-of \verb|AlgF| in Figure~\ref{fig:algf} (instantiating the functor
-with \verb|ListF A|).  This means that from the invocation of
+\verb|fo|.  This means that from the invocation of
 \verb|breakr|, we get \verb|w : list A| and \verb|z : R|.  Thus, it is
@@
 \verb|z| to recurse.  The figure also shows the code for the subsidiary
-recursion \verb|wordsByr|.
+recursion \verb|wordsByr| and top-level recursion \verb|wordsBy|.
 
@@
   fo (Const (list (list A))) (FunConst (list (list A))) (WordsByAlg p) xs.
+Definition wordsBy(p : A -> bool)(xs : List A) : list (list A) :=
+  wordsByr (fold (ListF A)) p xs.
 \end{minted}
@@
 \verb|a|, but also the tail \verb|as|.  It then returns the value
-\verb|b| to include in the output list, and whatever other list it
+\verb|b| to include in the output list, and whatever sublist it
 wishes, upon which \verb|mapThrough| will then recurse.
@@
 
-Given this definition, the Coq definition of \verb|mapThrough| is shown in Figure~\ref{fig:mapthrough}.  
+Given this definition, Coq code for \verb|mapThrough| is shown in Figure~\ref{fig:mapthrough}.  
 \verb|MapThroughAlg| is similar to the
-Haskell code above, though when we call \verb|f|, we must
+Haskell code in Figure~\ref{fig:mapthroughhs}, though when we call \verb|f|, we must
 supply the abstract type \verb|R| and fold function \verb|fo|.  
@@
 types~\cite{blanqui05}, but this requires changing the underlying
-theory.  Here we exploit Coq's
-impredicative polymorphism to obtain a different solution.
+theory.  Here we derive a different solution.
 
@@
 idea is embodied in the definition of \verb|Mu|.  Ideally, we would like
-to have a definition like
+to use this definition:
 \begin{minted}{coq}
@@
 
-For a simple example: suppose we instantiate \verb|F| with
-\verb|ListF A| (from Section~\ref{sec:recu}).  Our derivation uses a
+For a simple example of noncanonicity: suppose we instantiate \verb|F| with
+\verb|ListF A| (from Section~\ref{sec:recu}).  Our derivation actually uses a
 different type that wraps \verb|F|, but using \verb|ListF A|
@@
 and \verb|unroll| functions between \verb|SubrecF Subrec| and \verb|Subrec|.
-Unrolling a \verb|Subrec| term gives us a term of type
+The type \verb|SubrecF Subrec| is definitionally equal to
 \begin{minted}{coq}
@@
 \end{minted}
-\noindent So we see that \verb|Subrec| is the type of functions which,
+\noindent So \verb|Subrec| is the type of functions which,
 for all algebras with functorial carrier
@@
 \verb|alg| with carrier \verb|X| (with \verb|fmap| function given by
-\verb|FunX|) over \verb|d : Subrec|, we \verb|unroll| the
-definition of \verb|Subrec| and apply that to the algebra (with its
+\verb|FunX|) over \verb|d : Subrec|, we \verb|unroll| \verb|d| and apply that to the algebra (with its
 carrier).
@@
 We have seen how to write subsidiary recursions in Coq.  But can one
-reason about these?  We turn now briefly to the
-interface to our development of subsidiary induction in Coq, and
+reason about these?  We turn now briefly to our interface for subsidiary induction in Coq, and
 some example proofs written using this interface.  Subsidiary induction is
@@
 expresses provability of \verb|X C| for \verb|d|, based on an indexed
-algebra and a value of type \verb|C d|, where \verb|C| is the (indexed) anchor
-type.  \verb|AlgFi| and \verb|Algi| are indexed versions
+algebra and a value of type \verb|C d|.  \verb|AlgFi| and \verb|Algi| are indexed versions
 of the algebras we saw for recursion.  The \verb|rec| function
@@
 recursion, but in constructive type theory, evidence of
-well-foundedness then propagates through code.  In contrast, our
-approach here, while less general, does not clutter code with proofs.
+well-foundedness then propagates through code.  Our
+approach, while less general, does not clutter code with proofs.
 Subsidiary recursion can be seen as a generalization of \emph{nested
